---
title: "veteran-lung-cancer-overview"
format: html
---

```{r setup, include=FALSE}
# packages
library(survival)
library(survminer)   # ggforest
library(dplyr)
library(forcats)
library(broom)
library(purrr)
library(ggplot2)
library(ranger)
library(glue)

# Load dataset explicitly from survival package
veteran <- survival::veteran
```

# Introduction

Survival analysis is a branch of statistics designed to model the time until an event of interest occurs — most often death, relapse, or system failure. Unlike standard regression methods, it can account for censoring (when the event hasn’t occurred by the end of observation) and produces estimates that describe both the probability of survival over time and the influence of predictors on that time. This makes it invaluable in clinical trials, epidemiology, and engineering reliability studies.

The Veteran dataset comes from a randomized clinical trial conducted by the U.S. Veterans Administration to compare a standard chemotherapy regimen with an experimental treatment for advanced, inoperable lung cancer. It includes information on patient demographics, disease characteristics, treatment allocation, and survival time, making it a compact but rich teaching dataset for survival modelling.

This page provides a high-level overview of the analysis, including key descriptive statistics, primary survival estimates, and major findings from Cox proportional hazards models.

The project is organised into three parts:

Summary — the page you’re currently reading; a quick reference with key numbers and visualisations.

Full Report — a detailed, step-by-step account of the entire analysis, methods, and results.

Teaching & Learning Reference — a theory-heavy companion that explains the statistical reasoning behind each step, serving as both a technical appendix and a reusable “cheat sheet” for future survival analyses.

In the sections that follow, we start with a concise study overview describing the dataset and patient population. We then highlight the main survival patterns using Kaplan–Meier curves, followed by key hazard ratio estimates from Cox proportional hazards models. The goal is to give you, at a glance, the most important findings of the analysis — with the full technical detail available in the accompanying report and reference materials.

# Study Overview

This dataset records information from 137 male patients with advanced, inoperable lung cancer who were enrolled in a randomised controlled trial conducted by the U.S. Veterans Administration. Participants were allocated to receive either a standard chemotherapy protocol or an experimental regimen, with the primary aim of assessing whether the new treatment improved survival.

Key study characteristics:

    Population: 137 male patients, predominantly older adults.

    Condition: Advanced, inoperable lung cancer (various histological subtypes).

    Intervention: Standard chemotherapy vs experimental chemotherapy.

    Primary Outcome: Time from treatment initiation to death, measured in days.

    Follow-up: All patients were followed until death or censoring.

Additional recorded variables include Karnofsky performance score (a measure of functional status), age at diagnosis, time from diagnosis to randomisation, tumour cell type, and whether patients had received prior therapy.

This combination of survival time, treatment allocation, and clinically relevant covariates makes the Veteran dataset an ideal teaching resource for exploring survival analysis techniques such as Kaplan–Meier estimation, Cox proportional hazards modelling, and assumption diagnostics.

# Key Descriptive Statistics

```{r key-descriptive-stats, collapse=TRUE}
# Load and clean dataset
veteran2 <- veteran %>%
  mutate(
    trt      = factor(trt, levels = c(1, 2), labels = c("Standard", "Experimental")),
    celltype = factor(celltype, levels = c("squamous", "smallcell", "adeno", "large"),
                      labels = c("Squamous", "Small cell", "Adenocarcinoma", "Large cell")),
    prior    = factor(prior, levels = c(0, 10), labels = c("No", "Yes"))
  )

# Counts by treatment group
group_counts <- veteran2 %>%
  count(trt, name = "n")

# Age & Karnofsky summaries
age_summary <- summary(veteran2$age)
karno_summary <- summary(veteran2$karno)

# Cell type distribution
celltype_dist <- prop.table(table(veteran2$celltype)) * 100

# Prior therapy proportion
prior_dist <- prop.table(table(veteran2$prior)) * 100

# Kaplan–Meier medians by group
km_fit <- survfit(Surv(time, status) ~ trt, data = veteran2)
km_meds <- surv_median(km_fit)

# Log-rank test
logrank <- survdiff(Surv(time, status) ~ trt, data = veteran2)
logrank_p <- 1 - pchisq(logrank$chisq, length(logrank$n) - 1)

# Event breakdown
events <- sum(veteran2$status == 1)
censored <- sum(veteran2$status == 0)

# Print results
cat("\n**Sample size:**", nrow(veteran2), "patients\n")
cat("Treatment groups:\n")
print(group_counts)

cat("\nMedian age:", round(age_summary[["Median"]]), 
    "years (IQR:", age_summary[["1st Qu."]], "-", age_summary[["3rd Qu."]], ")\n")

cat("Median Karnofsky score:", round(karno_summary[["Median"]]), 
    "(IQR:", karno_summary[["1st Qu."]], "-", karno_summary[["3rd Qu."]], ")\n")

cat("\nCell type distribution (%):\n")
print(round(celltype_dist, 1))

cat("\nPrior therapy (%):\n")
print(round(prior_dist, 1))

cat("\nMedian survival (days):\n")
print(km_meds)

cat("\nLog-rank p-value:", signif(logrank_p, 3), "\n")

cat("\nDeaths:", events, "(", round(events / nrow(veteran2) * 100, 1), "%)\n")
cat("Censored:", censored, "(", round(censored / nrow(veteran2) * 100, 1), "%)\n")

# Overall
km_all <- survfit(Surv(time, status) ~ 1, data = veteran)
overall <- surv_median(km_all)

# By treatment (1=Standard, 2=Experimental)
vet2 <- within(veteran, trt <- factor(trt, levels = c(1,2), labels = c("Standard","Experimental")))
km_by <- survfit(Surv(time, status) ~ trt, data = vet2)
byarm  <- surv_median(km_by)

# Nicely formatted output
print(overall)
print(byarm)
```
# Key Descriptive Statistics Summary

    Sample size: 137 patients

    Treatment groups: Standard = 69, Experimental = 68

Baseline characteristics

    Median age: 62 years (IQR: 51–66)

    Median Karnofsky score: 60 (IQR: 40–75)

    Cell type distribution: Squamous 25.5%, Small cell 35.0%, Adenocarcinoma 19.7%, Large cell 19.7%

    Prior therapy: 29.2% “Yes” (70.8% “No”)

Survival outcomes

    Log-rank test: p = 0.928 (no evidence of a difference between treatment arms)

    Median survival (all / by arm): [insert from KM table below]

Event status

    Deaths: 128 (93.4%)

    Censored: 9 (6.6%)


```{r key-stats-bullets, collapse=TRUE}
# Data (explicit to be bulletproof)
v2 <- veteran %>%
  mutate(
    trt      = factor(trt, levels = c(1, 2), labels = c("Standard", "Experimental")),
    celltype = factor(celltype, levels = c("squamous","smallcell","adeno","large"),
                      labels = c("Squamous","Small cell","Adenocarcinoma","Large cell")),
    prior    = factor(prior, levels = c(0, 10), labels = c("No","Yes"))
  )

# N and group counts
n_total <- nrow(v2)
n_by_trt <- v2 %>% count(trt, name = "n")
n_std <- n_by_trt$n[n_by_trt$trt == "Standard"]
n_exp <- n_by_trt$n[n_by_trt$trt == "Experimental"]

# Age & Karnofsky medians + IQR
q_age   <- quantile(v2$age,    probs = c(0.25, 0.5, 0.75), na.rm = TRUE)
q_karno <- quantile(v2$karno,  probs = c(0.25, 0.5, 0.75), na.rm = TRUE)
med_age    <- round(q_age[2]);    iqr_age_l <- round(q_age[1]);    iqr_age_u <- round(q_age[3])
med_karno  <- round(q_karno[2]);  iqr_kar_l <- round(q_karno[1]);  iqr_kar_u <- round(q_karno[3])

# Cell type distribution (%)
cell_pct <- round(100 * prop.table(table(v2$celltype)), 1)
cell_lines <- paste0(names(cell_pct), " (", cell_pct, "%)")
cell_line  <- paste(cell_lines, collapse = ", ")

# Prior therapy (% Yes / No)
prior_pct <- round(100 * prop.table(table(v2$prior)), 1)
prior_yes <- unname(prior_pct["Yes"])
prior_no  <- unname(prior_pct["No"])

# KM medians (overall + by arm) with 95% CI
km_all  <- survfit(Surv(time, status) ~ 1,  data = v2)
overall <- surv_median(km_all)  # columns: median, lower, upper
med_all <- round(overall$median);  ci_all_l <- round(overall$lower);  ci_all_u <- round(overall$upper)

km_by  <- survfit(Surv(time, status) ~ trt, data = v2)
byarm  <- surv_median(km_by)
# rows are by trt strata in level order (Standard, Experimental)
med_std   <- round(byarm$median[byarm$strata == "trt=Standard"])
ci_std_l  <- round(byarm$lower [byarm$strata == "trt=Standard"])
ci_std_u  <- round(byarm$upper [byarm$strata == "trt=Standard"])
med_exp   <- round(byarm$median[byarm$strata == "trt=Experimental"])
ci_exp_l  <- round(byarm$lower [byarm$strata == "trt=Experimental"])
ci_exp_u  <- round(byarm$upper [byarm$strata == "trt=Experimental"])

# Log-rank test
lr <- survdiff(Surv(time, status) ~ trt, data = v2)
lr_p <- signif(1 - pchisq(lr$chisq, length(lr$n) - 1), 3)

# Event breakdown
deaths   <- sum(v2$status == 1)
censored <- sum(v2$status == 0)
deaths_pct   <- round(100 * deaths   / n_total, 1)
censored_pct <- round(100 * censored / n_total, 1)

# ---- Print Markdown bullets ----
cat(glue("
- **Sample size:** {n_total} patients  
- **Treatment groups:** Standard = **{n_std}**, Experimental = **{n_exp}**

**Baseline characteristics**  
- **Median age:** **{med_age}** years (IQR: **{iqr_age_l}–{iqr_age_u}**)  
- **Median Karnofsky score:** **{med_karno}** (IQR: **{iqr_kar_l}–{iqr_kar_u}**)  
- **Cell type distribution:** {cell_line}  
- **Prior therapy:** **{prior_yes}%** \"Yes\" (**{prior_no}%** \"No\")

**Survival outcomes**  
- **Median survival (all patients):** **{med_all}** days (95% CI: **{ci_all_l}–{ci_all_u}**)  
- **By treatment group:**  
  - **Standard:** **{med_std}** days (95% CI: **{ci_std_l}–{ci_std_u}**)  
  - **Experimental:** **{med_exp}** days (95% CI: **{ci_exp_l}–{ci_exp_u}**)  
- **Log-rank test:** *p* = **{lr_p}**  

**Event status**  
- **Deaths:** **{deaths}** ({deaths_pct}%)  
- **Censored:** **{censored}** ({censored_pct}%)
"))
```


```{r}
# factor housekeeping to get nice labels
veteran2 <- veteran %>%
  mutate(
    trt      = factor(trt, levels = c(1, 2), labels = c("Standard", "Experimental")),
    celltype = factor(celltype, levels = c("squamous","smallcell","adeno","large"),
                      labels = c("Squamous","Small cell","Adeno","Large")),
    prior    = factor(prior, levels = c(0, 10), labels = c("No","Yes"))
  )

# multivariable Cox model (tweak as per your analysis)
cox_multi <- coxph(Surv(time, status) ~ trt + celltype + karno + age + diagtime + prior,
                   data = veteran2)

# publishable forest plot
ggforest(
  cox_multi, data = veteran2,
  main = "Hazard Ratios (Cox PH): Veteran Lung Cancer Trial",
  fontsize = 0.9,
  noDigits = 2,
  refLabel = "Reference"
)

```

```{r}

# choose covariates to scan (you can add/remove)
covars <- c("trt","celltype","karno","age","diagtime","prior")

uni_df <- map_dfr(covars, function(v) {
  f <- as.formula(paste0("Surv(time, status) ~ ", v))
  fit <- coxph(f, data = veteran2)
  broom::tidy(fit, exponentiate = TRUE, conf.int = TRUE) %>%
    mutate(term_group = v)
})

# pretty labels
uni_df <- uni_df %>%
  mutate(
    pretty = case_when(
      term == "trtExperimental" ~ "Treatment: Experimental vs Standard",
      term == "priorYes"        ~ "Prior therapy: Yes vs No",
      grepl("celltype", term)   ~ sub("celltype", "Cell type: ", term),
      term == "karno"           ~ "Karnofsky (per point)",
      term == "age"             ~ "Age (per year)",
      term == "diagtime"        ~ "Months to randomization (per month)",
      TRUE ~ term
    )
  )

ggplot(uni_df, aes(x = forcats::fct_rev(pretty), y = estimate)) +
  geom_point() +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0) +
  geom_hline(yintercept = 1, linetype = 2) +
  coord_flip() +
  labs(
    x = NULL, y = "Hazard Ratio (95% CI)",
    title = "Univariable Cox Models"
  ) +
  theme_minimal(base_size = 12)

```

```{r}

multi_tidy <- broom::tidy(cox_multi, exponentiate = TRUE, conf.int = TRUE) %>%
  mutate(
    pretty = case_when(
      term == "trtExperimental" ~ "Treatment: Experimental",
      term == "priorYes"        ~ "Prior therapy: Yes",
      grepl("celltype", term)   ~ sub("celltype", "Cell type: ", term),
      term == "karno"           ~ "Karnofsky (per point)",
      term == "age"             ~ "Age (per year)",
      term == "diagtime"        ~ "Months to randomization (per month)",
      TRUE ~ term
    )
  )

ggplot(multi_tidy, aes(x = forcats::fct_rev(pretty), y = estimate)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0) +
  geom_hline(yintercept = 1, linetype = 2) +
  coord_flip() +
  theme_minimal(base_size = 12) +
  labs(
    x = NULL, y = "Adjusted Hazard Ratio (95% CI)",
    title = "Multivariable Cox Model"
  )

```

```{r}

# KM (overall)
km_fit  <- survfit(Surv(time, status) ~ 1, data = veteran2)
# Cox baseline survival
cox_fit <- survfit(cox_multi)  # from the model above
# Random forest survival
rf_fit  <- ranger(Surv(time, status) ~ trt + celltype + karno + age + diagtime + prior,
                  data = veteran2, mtry = 3, splitrule = "logrank", num.trees = 1000)

# build a combined df like the article, then ggplot the three curves
# (structure mirrors the RViews code)
```
