---
title: "Veteran Lung Cancer Survival Analysis: Overview"
format:
  html:
    toc: true
    theme: cosmo
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(survival)
library(survminer)
library(dplyr)
library(forcats)
library(broom)
library(purrr)
library(ggplot2)
library(splines)
library(glue)
library(ggfortify)

# Cleaned dataset with factor labels used throughout
veteran2 <- survival::veteran %>%
  mutate(
    trt      = factor(trt, levels = c(1, 2), labels = c("Standard", "Experimental")),
    celltype = factor(celltype, levels = c("squamous", "smallcell", "adeno", "large"),
                      labels = c("Squamous", "Small cell", "Adenocarcinoma", "Large cell")),
    prior    = factor(prior, levels = c(0, 10), labels = c("No", "Yes"))
  )
```

# Introduction

Survival analysis models the time until an event of interest occurs — most
often death, relapse, or system failure. Unlike standard regression, it
accounts for censoring (patients still alive at study end) and estimates both
the probability of survival over time and the influence of covariates on that
time.

This analysis uses the Veterans Administration lung cancer trial dataset
(Kalbfleisch & Prentice, 1980): 137 male patients with advanced, inoperable
lung cancer randomised to standard vs experimental chemotherapy. It is a
well-established teaching dataset for survival modelling — compact,
clinically relevant, and rich enough to illustrate the full Cox PH workflow.

The project is organised into three parts:

- **Overview** — this page; key numbers and visualisations at a glance.
- **Full Report** — step-by-step analysis, model building, and diagnostics.
- **Teaching & Learning Reference** — statistical theory, code explanations,
  and a reusable reference for future survival analyses.

# Study Overview

```{r descriptive-stats}
# KM medians overall and by arm
km_all <- survfit(Surv(time, status) ~ 1,   data = veteran2)
km_by  <- survfit(Surv(time, status) ~ trt, data = veteran2)
overall <- surv_median(km_all)
byarm   <- surv_median(km_by)

med_all   <- round(overall$median)
ci_all_l  <- round(overall$lower);  ci_all_u  <- round(overall$upper)
med_std   <- round(byarm$median[byarm$strata == "trt=Standard"])
ci_std_l  <- round(byarm$lower [byarm$strata == "trt=Standard"])
ci_std_u  <- round(byarm$upper [byarm$strata == "trt=Standard"])
med_exp   <- round(byarm$median[byarm$strata == "trt=Experimental"])
ci_exp_l  <- round(byarm$lower [byarm$strata == "trt=Experimental"])
ci_exp_u  <- round(byarm$upper [byarm$strata == "trt=Experimental"])

# Baseline summaries
n_total   <- nrow(veteran2)
n_by_trt  <- veteran2 %>% count(trt, name = "n")
n_std_n   <- n_by_trt$n[n_by_trt$trt == "Standard"]
n_exp_n   <- n_by_trt$n[n_by_trt$trt == "Experimental"]

q_age   <- quantile(veteran2$age,   probs = c(0.25, 0.5, 0.75))
q_karno <- quantile(veteran2$karno, probs = c(0.25, 0.5, 0.75))
med_age   <- round(q_age[2]);   iqr_age_l <- round(q_age[1]);   iqr_age_u <- round(q_age[3])
med_karno <- round(q_karno[2]); iqr_kar_l <- round(q_karno[1]); iqr_kar_u <- round(q_karno[3])

cell_pct  <- round(100 * prop.table(table(veteran2$celltype)), 1)
prior_pct <- round(100 * prop.table(table(veteran2$prior)),    1)
prior_yes <- unname(prior_pct["Yes"])

lr   <- survdiff(Surv(time, status) ~ trt, data = veteran2)
lr_p <- signif(1 - pchisq(lr$chisq, length(lr$n) - 1), 3)

deaths      <- sum(veteran2$status == 1)
censored    <- sum(veteran2$status == 0)
deaths_pct  <- round(100 * deaths   / n_total, 1)
cens_pct    <- round(100 * censored / n_total, 1)

cat(glue("
**Study population**
- {n_total} male patients; Standard = **{n_std_n}**, Experimental = **{n_exp_n}**
- Median age: **{med_age}** years (IQR: {iqr_age_l}–{iqr_age_u})
- Median Karnofsky score: **{med_karno}** (IQR: {iqr_kar_l}–{iqr_kar_u})
- Cell type: Squamous {cell_pct['Squamous']}%, Small cell {cell_pct['Small cell']}%,
  Adenocarcinoma {cell_pct['Adenocarcinoma']}%, Large cell {cell_pct['Large cell']}%
- Prior therapy: **{prior_yes}%** yes

**Survival outcomes**
- Median survival (all): **{med_all}** days (95% CI: {ci_all_l}–{ci_all_u})
- Standard arm: **{med_std}** days (95% CI: {ci_std_l}–{ci_std_u})
- Experimental arm: **{med_exp}** days (95% CI: {ci_exp_l}–{ci_exp_u})
- Log-rank *p* = **{lr_p}**

**Events**
- Deaths: **{deaths}** ({deaths_pct}%) | Censored: **{censored}** ({cens_pct}%)
"))
```

# Plot 1 — Survival by Treatment Group

The primary question of the trial: does the experimental regimen improve
survival over standard chemotherapy?

```{r km-treatment, fig.width=8, fig.height=6}
ggsurvplot(
  km_by,
  data        = veteran2,
  risk.table  = TRUE,
  pval        = TRUE,
  conf.int    = TRUE,
  palette     = c("#2E86AB", "#E84855"),
  legend.title = "Treatment",
  xlab        = "Time (days)",
  ylab        = "Survival probability",
  title       = "Kaplan–Meier Survival Curves by Treatment Arm",
  ggtheme     = theme_minimal(base_size = 13)
)
```

The survival curves for the two treatment arms overlap almost entirely
throughout follow-up. The log-rank test confirms no statistically significant
difference (p = `r lr_p`). The experimental chemotherapy did not improve
survival in this cohort.

# Plot 2 — Survival by Karnofsky Performance Score

Karnofsky score — a measure of physical functional status scored 0–100 — is
the dominant prognostic factor in this dataset and in oncology more broadly.

```{r km-karnofsky, fig.width=8, fig.height=6}
veteran2$karno_group <- cut(
  veteran2$karno,
  breaks = c(0, 50, 70, 100),
  labels = c("Low (≤50)", "Medium (51–70)", "High (>70)"),
  right  = TRUE
)

km_karno <- survfit(Surv(time, status) ~ karno_group, data = veteran2)

ggsurvplot(
  km_karno,
  data         = veteran2,
  risk.table   = TRUE,
  pval         = TRUE,
  conf.int     = FALSE,
  palette      = c("#E84855", "#F4A259", "#2E86AB"),
  legend.title = "Karnofsky group",
  legend.labs  = c("Low (≤50)", "Medium (51–70)", "High (>70)"),
  xlab         = "Time (days)",
  ylab         = "Survival probability",
  title        = "Kaplan–Meier Survival Curves by Karnofsky Score Group",
  ggtheme      = theme_minimal(base_size = 13)
)
```

The separation between groups is striking: patients with high functional status
survived substantially longer than those in the low group. This gradient
persists and is highly significant, foreshadowing Karnofsky score's dominance
in the multivariable model.

# Plot 3 — Final Cox Model: Adjusted Hazard Ratios

The final Cox proportional hazards model adjusts simultaneously for treatment
group, prior therapy, age (modeled with a natural cubic spline), Karnofsky
score, and cell type.

```{r cox-forest, fig.width=9, fig.height=6}
cox_final <- coxph(
  Surv(time, status) ~ trt + prior + ns(age, df = 3) + karno + celltype,
  data = veteran2
)

ggforest(
  cox_final,
  data      = veteran2,
  main      = "Adjusted Hazard Ratios — Final Cox PH Model",
  fontsize  = 0.9,
  noDigits  = 2,
  refLabel  = "Reference"
)
```

Karnofsky score (HR = 0.97 per point, p < 0.001) and cell type are the
dominant predictors. Adenocarcinoma and small cell carcinoma carry
substantially higher hazard than squamous cell (HR ≈ 3.3 and 2.2
respectively). Treatment group remains non-significant after adjustment
(HR ≈ 1.4, p = 0.097). The model concordance is 0.735 — a marked improvement
over a treatment-only model (concordance ≈ 0.51).

# Plot 4 — Aalen Additive Model: Time-Varying Effects

The Cox model assumes that each covariate's effect is constant over time
(proportional hazards). The Aalen additive model relaxes this assumption,
estimating cumulative covariate effects as a function of follow-up time. A
steep, consistent slope indicates a strong, stable effect; a flat or erratic
line suggests little or no independent contribution.

```{r aalen, fig.width=10, fig.height=8}
aa_fit <- aareg(
  Surv(time, status) ~ trt + celltype + karno + age + prior,
  data = veteran2
)

autoplot(aa_fit) +
  theme_minimal(base_size = 12) +
  labs(title = "Aalen Additive Model: Cumulative Regression Functions")
```

The Aalen plots reinforce the Cox model findings from a different methodological
angle. Karnofsky score shows the steepest, most consistent negative slope —
confirming its role as the strongest and most stable predictor of survival.
Cell type subtypes (particularly small cell and adenocarcinoma) show clear
positive slopes relative to the intercept. Treatment group and prior therapy
show flat trajectories close to zero, consistent with their non-significant
status in the Cox model. The convergence of two distinct modelling approaches
on the same substantive conclusions strengthens confidence in the findings.
