---
title: "Cox Proportional Hazards in R: A Practical Reference"
author: "Julian Chung"
format:
  html:
    toc: true
    code-fold: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

# Introduction

This notebook is a practical guide to running a Cox proportional hazards
analysis in R. It is written for someone who already knows R and basic
biostatistics, but wants a solid refresher on the survival-analysis workflow:
how to set up the data, inspect survival curves, fit Cox models, check the main
assumptions, and refine the model when something looks off.

The examples use the Veterans Administration lung cancer trial dataset, a
well-known teaching dataset that is compact enough to work through end to end
but rich enough to surface the main issues that come up in real survival
analysis. The goal here is not just to fit one model for this dataset, but to
walk through a reusable framework that can be adapted to other time-to-event
projects.

This notebook is designed as a refresher and self-reference for anyone coming
back to survival analysis after a long gap: part syntax reminder, part workflow
guide, and part statistical refresher.

# Setup and Data Import

Start by loading the core packages for survival modelling, plotting, and data
wrangling. The example dataset comes from the `survival` package and is relabeled
slightly to make the output easier to read.

```{r}
# Load required libraries
library(survival) # Core survival analysis tools
library(survminer)  # Visualization and diagnostics
library(dplyr)  # Data wrangling
library(ggplot2)  # Plotting
library(broom)  # Tidying model output
library(splines) # Splines for flexible modeling
library(forestmodel) # Forest plots for Cox models

```

::: {.callout-tip title="Why These Packages?" collapse="true"}
-   **`survival`**: Core package for survival analysis in R. Provides
    `Surv()` to construct survival objects and `coxph()` for Cox models.

-   **`survminer`**: High-level visualization tools for survival curves
    and diagnostics (e.g., `ggsurvplot()`, `ggcoxdiagnostics()`).

-   **`splines`**: Lets you model non-linear relationships with natural splines
    (for example, `ns(age, df = 3)`).

-   **`ggplot2`**: Used to build layered, customizable plots — often
    integrated with `survminer`.

-   **`dplyr`**: Enables clean and readable data manipulation (e.g.,
    `mutate()`, `filter()`).

-   **`broom`**: Tidies model outputs into data frames (useful for
    summaries, tables, and downstream analysis).
:::

Next, load the dataset and do a small amount of cleanup so the treatment labels
are readable in plots and model output.

```{r}
# Load the dataset
veteran <- survival::veteran %>%
  mutate(trt = factor(trt,
                      levels = c(1, 2),
                      labels = c("Standard", "Experimental")))

# Peek at the structure and summary
# glimpse(veteran)
# summary(veteran)
```

::: {.callout-note title="Dataset Preview" collapse="true"}
It is worth looking at the first few rows before doing anything else, just to
confirm the basic structure and coding.

```{r}
head(veteran)
```
:::

**Variable Summary:**

The main variables used in this guide are summarised below.

| Variable | Type | Description | Notes |
|------------|------------|-------------------------|------------------------|
| `trt` | Factor | Treatment group: 1 = Standard, 2 = Test | Relabeled to `"Standard"` and `"Experimental"` |
| `celltype` | Factor | Histologic type of lung cancer: squamous, small cell, adeno, large | Used later in the fuller multivariable models |
| `time` | Numeric | Survival time in days | Outcome variable in the `Surv()` object |
| `status` | Numeric | Event indicator: 1 = died, 0 = censored | Used to define survival endpoint |
| `karno` | Numeric | Karnofsky performance score (proxy for baseline health status) | A strong clinical predictor; included later in the fuller models |
| `diagtime` | Numeric | Time from diagnosis to study entry (in months) | Possibly useful in extended time-dependent models |
| `age` | Numeric | Patient age (in years) | Treated as **continuous**; **checked for linearity and splines** |
| `prior` | Numeric | Number of prior therapies (0 or 10) | Binary variable: 0 = none, 10 = prior treatment |

# Exploratory Survival Analysis

A good survival-analysis workflow starts with a non-parametric look at the data.
Kaplan-Meier curves are the easiest way to inspect survival patterns before
moving on to regression models.

::: {.callout-note title="Parametric, Non-Parametric, and Semi-Parametric Models" collapse=true}

When coming back to survival analysis after a gap, it helps to remember the
basic distinction:

**Non-parametric methods** such as Kaplan-Meier curves and the log-rank test do
not assume a particular shape for the survival distribution. They are useful for
initial exploration, visual comparison of groups, and quick checks before moving
to regression.

**Parametric survival models** assume a specific distribution for survival times
(for example Weibull or exponential). They can be powerful, especially for
prediction or extrapolation, but only when that distribution is a reasonable fit
for the data.

**Semi-parametric models** sit in between. The Cox model does not specify the
baseline hazard, but it does assume that covariates act multiplicatively on the
hazard and that those effects are proportional over time.

In practice:
- use Kaplan-Meier and log-rank methods to get oriented
- use Cox models when you want adjusted hazard ratios
- use parametric models when you have a good reason to assume a survival
  distribution or need extrapolation beyond the observed follow-up

:::

## Kaplan-Meier Survival Curve

The first step is to estimate the survival curves for each treatment group using
the Kaplan-Meier estimator.

```{r}
# Create the survival object
surv_obj <- Surv(time = veteran$time, event = veteran$status)

# Fit Kaplan-Meier curve stratified by treatment
km_fit <- survfit(surv_obj ~ trt, data = veteran)

# View basic survival summary
# summary(km_fit)

```

::: {.callout-tip title="Survival Summary Truncated for Clarity" collapse="true"}
The full Kaplan-Meier summary includes hundreds of time points. For
readability, we display only the first few rows per treatment group.

```{r}
# View basic survival summary
summary(km_fit)
```
:::

::: {.callout-note title="Why Start with a Kaplan-Meier Curve?" collapse="true"}
This is usually the right place to start because it gives you a quick feel for:

- the overall pace of events
- whether the groups separate at all
- whether there are obvious crossing curves or late divergence
- how much censoring you are dealing with

It is also a good sense check before fitting a Cox model. If the Kaplan-Meier
curves already look strange, heavily crossed, or unstable at the tail, that is a
signal to be careful when you move into regression and diagnostics.
:::

```{r}
# Plot the KM curve
ggsurvplot(
  km_fit,
  data = veteran,
  conf.int = TRUE,
  pval = TRUE,
  pval.size = 4,
  risk.table = TRUE,
  risk.table.title = "Number at Risk",
  risk.table.col = "strata",
  risk.table.y.text.col = TRUE,
  risk.table.fontsize = 4,
  legend.title = "Treatment Group",
  legend.labs = c("Standard", "Experimental"),
  xlab = "Time (days)",
  ylab = "Survival Probability",
  palette = c("#A569BD", "#45B39D"),
  title = "Kaplan-Meier Survival Curve: VA Lung Cancer Trial",
  ggtheme = theme_minimal(base_size = 13) + 
    theme(
      legend.position = "top",
      plot.title = element_text(hjust = 0.5, face = "bold"),
      panel.grid.major = element_line(color = "grey80", linewidth = 0.4),
      panel.grid.minor = element_line(color = "grey90", linewidth = 0.2)
    ),
  tables.theme = theme(  # Trying to tweak the risk table
    axis.text.y = element_text(size = 10, margin = margin(r = 12), lineheight = 2)
  )
)
```

**Interpretation of the Kaplan-Meier Plot**

The Kaplan-Meier curves are the first sense check. Here, both treatment groups
show a steep drop in survival early in follow-up, which is what you would expect
in a cohort with advanced lung cancer. The curves and their confidence bands
overlap heavily, so there is no obvious visual signal that one treatment is
outperforming the other. The log-rank test backs that up with a *p*-value of
0.93.

::: {.callout-note title="How to Read a Kaplan-Meier Plot" collapse="true"}
**X-axis: time**

This is follow-up time, here measured in days.

**Y-axis: survival probability**

This is the estimated probability of surviving beyond a given time point.

**Step drops**

Each downward step corresponds to an event. In this dataset, the event is death.

**Shaded bands**

These are 95% confidence intervals. They usually widen later in follow-up
because fewer people remain at risk.

**Risk table**

The table underneath the plot shows how many participants are still under
observation and event-free at each time point. This matters because the tail of
the curve becomes less stable once only a few patients remain.

**Log-rank p-value**

This gives a formal comparison of the two curves. A small p-value suggests that
the survival experience differs between groups; a large p-value suggests that
any visible difference could easily be due to chance.
:::

## Log-Rank Test

The log-rank test is the standard unadjusted test for comparing survival curves
between groups.

::: {.callout-note title="What the Log-Rank Test Is Actually Doing" collapse="true"}
At each event time, the log-rank test compares how many events were observed in
each group with how many would be expected if the groups had the same survival
experience. It then aggregates those differences across follow-up into a single
chi-square test statistic.

A few useful reminders:
- it is a group comparison, not a regression model
- it is unadjusted, so it ignores other covariates
- it is most natural when the proportional hazards assumption is at least roughly plausible
- it tells you whether the curves differ overall, but not by how much in an adjusted sense
:::

```{r}
# Perform the log-rank test
log_rank_test <- survdiff(surv_obj ~ trt, data = veteran)
log_rank_test
```

```{r}
# Extract the p-value
1 - pchisq(log_rank_test$chisq, df = length(log_rank_test$n) - 1)
```

::: {.callout-note title="Quick Refresher: Chi-Square and Degrees of Freedom" collapse="true"}
For a two-group log-rank test, the chi-square statistic measures how far the
observed event counts are from what would be expected under equal survival, and
the degrees of freedom are 1. A larger chi-square gives a smaller *p*-value.
:::

**Interpretation of the Log-Rank Test Results**

In this example, the log-rank test gives a chi-square statistic of 0.008 with 1
degree of freedom and a *p*-value of 0.93. That is exactly what the Kaplan-Meier
plot suggested: the two treatment groups have very similar unadjusted survival
experience.

This is useful, but it is only a starting point. The log-rank test does not
adjust for other clinical variables, and it does not tell you how large the
treatment effect is after adjustment. That is where the Cox model becomes more
useful.

::: {.callout-note title="Why Keep Going After a Non-Significant Log-Rank Test?" collapse="true"}
A non-significant log-rank test is not a reason to stop.

A Cox model is still useful because it lets you:
- adjust for other covariates
- estimate hazard ratios and confidence intervals
- check whether the treatment result changes once baseline differences are taken into account
- investigate assumptions, non-linearity, and interaction terms

In other words, the log-rank test answers a narrow unadjusted question. The Cox
model gives you a much fuller framework for explanation and interpretation.
:::

## Why Proceed with the Cox Model?

Even when the Kaplan-Meier plot and log-rank test show little difference between
groups, the next step is often still a Cox model. That gives you a way to adjust
for other predictors, quantify effects with hazard ratios, and start checking
whether the model assumptions are reasonable.

- Adjust for other clinical predictors
- Estimate hazard ratios with confidence intervals
- Check proportional hazards and functional form
- Explore whether interactions are worth considering

Thus, we proceed with a Cox model to assess whether treatment effects become more apparent after adjusting for relevant clinical factors.

# Cox Regression Models

At this point, the workflow shifts from unadjusted description to regression
modelling. The Cox model is the standard tool when you want to relate survival
to multiple covariates without specifying the baseline hazard.

The main assumptions to keep in mind are:
- proportional hazards: effects are constant over time on the hazard-ratio scale
- sensible functional form for continuous predictors
- reasonable coding of categorical variables

The easiest way to build confidence in the model is to start simple, then add
complexity one step at a time and check what changes.

::: {.callout-note title="Understanding the Cox Proportional Hazards Model" collapse="true"}
The Cox model relates covariates to the hazard of the event over time without
having to specify the baseline hazard.

The key quantity is the **hazard**, which you can think of as the instantaneous
risk of the event at time *t*, given survival up to that point.

In practice, the main reason to use a Cox model is that it lets you:

- estimate hazard ratios for multiple predictors at once
- adjust for confounding variables
- move beyond simple group-to-group curve comparisons

The key assumption is **proportional hazards**: hazard ratios are assumed to be
constant over time. That assumption needs to be checked, especially once the
model becomes more complex.

The core outputs to pay attention to are:
- coefficients and hazard ratios
- confidence intervals
- Wald, likelihood ratio, and score tests
- concordance
- residual-based diagnostics
:::

## Univariable Models

In this section, we fit a series of univariable Cox models to assess the effect of individual covariates on survival. This step helps us explore the independent contribution of each variable before building more complex multivariable models.

### Treatment Group Only

We begin with a univariable Cox model that includes only the treatment
variable (`trt`). This allows us to assess the effect of treatment on
survival without adjusting for any other covariates.

```{r}
# Fit the Cox Proportional Hazards model
cox_model <- coxph(surv_obj ~ trt, data = veteran) # Where trt is a binary treatment variable with levels Standard and Experimental

# Summarize the model
summary(cox_model)

```

::: {.callout-note title="Interpretation of Cox Table" collapse="true"}
| Term | Value | What It Means |
|----------|----------|----------------------------------------------------|
| **coef** | 0.01774 | This is the log hazard ratio (log(HR)) comparing the *Experimental* group to *Standard*. Close to 0 means almost no effect. |
| **exp(coef)** | 1.01790 | This is the hazard ratio (HR). A value of 1.018 means the hazard (risk of death) is 1.8% higher in the Experimental group compared to Standard. However, at this level the value is not meaningful. |
| **se(coef)** | 0.18066 | The standard error. This gives us the variability around the estimate. |
| **z** | 0.098 | The Wald test statistic. The low value implies no meaningful signal. |
| **p-value** | 0.922 | Far above the conventional 0.05 threshold. This is very strong evidence *against* any meaningful treatment effect. |
| **95% CI** | (0.7144, 1.45) | Wide confidence interval, spanning both below and above 1—so we can't rule out either harm or benefit. |

**Concordance** The C-index, or concordance statistic is the probability
that a randomly chosen subject who died earlier had a higher predicted
risk. At 0.525 it is close to random chance, suggesting poor predictive power.
:::

**Interpretation of the Cox Model Results**

The univariable Cox proportional hazards model found no statistically significant difference in survival between the control and intervention groups (HR = 1.02, 95% CI: 0.71–1.45, p = 0.92). The hazard ratio near 1 suggests little to no effect of the treatment on mortality risk. The wide confidence interval spans both potential benefit and harm, highlighting uncertainty in the estimate. The concordance statistic (0.525) indicates the model’s discriminative ability is only marginally better than chance. Overall, this analysis offers no evidence that the intervention improves survival in this sample.

```{r}
# Use ggadjustedcurves for plotting adjusted survival curves directly from the Cox model
ggadjustedcurves(
  cox_model,                     # The fitted Cox model
  data = veteran,                # The original data used to fit the model
  variable = "trt",              # The variable to plot adjusted curves for
  method = "average",            # Method for adjusting (average effect of other covariates)
  conf.int = TRUE,
  legend.labs = c("Standard", "Experimental"),
  legend.title = "Treatment Group",
  risk.table = TRUE,             # Include the risk table
  risk.table.title = "Number at Risk",
  xlab = "Time (days)",
  ylab = "Adjusted Survival Probability",
  title = "Cox Model Adjusted Survival Curves",
  palette = c("#A569BD", "#45B39D"),
  ggtheme = theme_minimal(base_size = 13) +
    theme(
      legend.position = "top",
      plot.title = element_text(hjust = 0.5, face = "bold"),
      panel.grid.major = element_line(color = "grey80", linewidth = 0.4),
      panel.grid.minor = element_line(color = "grey90", linewidth = 0.2)
    ),
  tables.theme = theme(
    axis.text.y = element_text(size = 10, margin = margin(r = 12), lineheight = 2)
  )
)
```

::: {.callout-tip title="Why use `ggadjustedcurves()` for this plot" collapse="true"}
For plotting survival curves adjusted for covariates from a Cox
Proportional Hazards model, the ggadjustedcurves() function from the
survminer package offers one of the most direct and statistically
appropriate approaches—particularly as model complexity increases.

-   Statistical validity: ggadjustedcurves() uses the fitted cox_model
    to compute adjusted survival probabilities for the variable of
    interest (in this case, trt). When method = "average" is specified
    (the default when covariates are present), it returns marginal
    estimates—appropriate for interpreting average treatment effects
    across the population. In our simpler model (\~ trt), it effectively
    plots the model-predicted survival curve for each treatment group.

-   Robustness: This function is purpose-built for adjusted curve
    plotting and avoids many of the pitfalls seen with manually passing
    newdata to survfit() and plotting with ggsurvplot(), such as
    misaligned strata or legend issues. It correctly associates the
    model, the data, and the plot elements, including the risk table.

-   Simplicity: It eliminates the need for manually constructing a
    new_trt data frame or calling survfit() separately, resulting in
    cleaner, more reproducible code—particularly useful in collaborative
    or long-term projects.
:::

**Interpretation of the Adjusted Survival Curves**

The Cox-adjusted survival curves for the Standard and Experimental groups are nearly identical, indicating no meaningful difference in predicted survival probabilities between the treatment arms. This visual overlap aligns with the Cox model's hazard ratio, which was close to 1 and not statistically significant. Together, the model output and adjusted curves reinforce the conclusion that the experimental regimen had no observable effect on survival in this cohort.

::: {.callout-note title="Why do the lines overlap?" collapse="true"}
The survival curves for the Standard and Experimental groups overlap
almost entirely in this plot. This is because the Cox proportional
hazards model found no significant difference between the two groups in
terms of survival (hazard ratio ≈ 1, *p* = 0.922). As a result, the
model-predicted curves are nearly identical. The faint visibility of the
Standard group line is due to it being directly behind the Experimental
group line.
:::

**Model Diagnostics**

Before interpreting the results of a Cox proportional hazards model, it’s important to ensure that the model’s assumptions are met. The two key assumptions are:

- The proportional hazards (PH) assumption: The effect of each covariate on the hazard is constant over time.

- Linearity of continuous covariates: Continuous covariates are assumed to have a linear relationship with the log hazard.

Violations of these assumptions can undermine the model’s predictive utility and lead to misleading conclusions.

In this section, we use the following diagnostic tools to assess model adequacy:

- Schoenfeld residuals to test the PH assumption.

- Martingale residuals to check for non-linearity and outliers.

- Functional form checks, particularly for continuous covariates like age.

- Where needed, we apply transformations (e.g., natural splines) to better capture nonlinear effects.

These checks help ensure that the model remains a sound and interpretable framework for estimating covariate effects on survival.

_Checking the Proportional Hazards Assumption_

A central assumption of the Cox proportional hazards model is that hazard ratios remain constant over time — meaning the relative risk between any two individuals does not change as time progresses. This is known as the proportional hazards (PH) assumption, and verifying it is crucial before interpreting model results.

**Schoenfeld Residuals**

To assess the PH assumption, we use Schoenfeld residuals, which evaluate whether covariate effects vary over time. If residuals show a systematic pattern against time, it may indicate a violation of the assumption.

We check this in two ways:

- Statistical tests: The cox.zph() function provides both a global test and tests for each covariate. A p-value < 0.05 suggests evidence against the PH assumption.

- Visual inspection: Plots of scaled Schoenfeld residuals include a smoothed line and 95% confidence bands. Systematic trends or large deviations from zero may indicate time-varying effects.

If either test indicates a violation, we may need to adjust the model using time-varying covariates, stratification, or alternative approaches.

```{r}
# Test the PH assumption
cox.zph(cox_model)
```

::: {.callout-note title="Understanding Chi-Squared and p-values in PH Assumption Testing" collapse="true"}
`cox.zph()` tests whether each coefficient looks time-dependent.

A few practical rules of thumb:
- a small *p*-value suggests the effect may be changing over time
- a borderline *p*-value should be interpreted alongside the residual plot
- the global test can be mildly significant even when no single variable looks clearly problematic
- use the test as a prompt to inspect the plots, not as a standalone verdict
:::

**Interpretation of PH Assumption Test (Schoenfeld residuals)**

For the treatment-only model, both the variable-level and global tests are
borderline at about *p* = 0.06. That is close enough to pay attention, but not
strong enough on its own to conclude that proportional hazards has failed.

**Schoenfeld Residuals Plot**

```{r}
plot(cox.zph(cox_model))
```

::: {.callout-note title="What the Visual Diagnostic Shows" collapse="true"}
In a Schoenfeld plot, the main thing to look for is whether the smoothed line is
roughly flat over time.

Useful rules of thumb:
- a flat line near zero supports proportional hazards
- a strong slope or clear curve suggests the effect may be changing over time
- the plot matters most when read alongside the `cox.zph()` test, especially for borderline results
:::

**Schoenfeld Residuals Interpretation**

Here the smooth shows mild curvature, but not a strong sustained trend, and it
stays within the confidence bands. Taken together with the borderline test
results, this is enough to treat proportional hazards as a reasonable working
assumption for the treatment-only model.

**Martingale Residuals (for Functional Form of Covariates)**

Martingale residuals are mainly useful for checking the functional form of
continuous predictors.

::: {.callout-note title="Understanding Martingale Residuals" collapse="true"}
Martingale residuals are mainly useful for checking the functional form of
continuous predictors.

What to look for:
- random scatter suggests a linear term may be adequate
- visible curvature suggests the effect may be non-linear
- for binary predictors, the plot is less about curvature and more about whether
  anything looks obviously badly specified
:::

```{r}
# Generate Martingale residuals
martingale_resid <- residuals(cox_model, type = "martingale")

# Add residuals to your data
veteran$martingale <- martingale_resid

# Plot residuals against a continuous covariate (e.g., time or another variable)
# Here we plot against the linear predictor for simplicity
veteran$linear_pred <- predict(cox_model, type = "lp")

ggplot(veteran, aes(x = linear_pred, y = martingale, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Linear Predictor",
    x = "Linear Predictor (Log Hazard)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Standard" = "#A569BD", "Experimental" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))


```

::: {.callout-note title="What the Martingale Residuals Show Here" collapse="true"}
Because treatment is binary, this plot is mostly a basic sanity check rather than
a functional-form check. The residuals are spread similarly across the two
groups, with no obvious pattern suggesting that the treatment term is badly
specified.
:::

**Interpretation of Martingale Residuals Plot**

Because treatment is binary, this plot is mostly a basic sanity check rather than
a functional-form check. The residuals are spread similarly across the two
groups, with no obvious pattern suggesting that the treatment term is badly
specified.

::: {.callout-tip title="Why we haven't checked the functional form of covariates" collapse="true"}
At this stage the model only contains treatment, which is binary. Functional-form
checks become much more important once continuous predictors like age,
Karnofsky score, or time since diagnosis enter the model.
:::

**Additional Covariates**

Once the treatment-only model is in place, the next step is to start adding
clinically relevant covariates. The immediate candidates here are age and prior
therapy, because they are easy to include and help illustrate two common cases:
a continuous predictor and a binary predictor.

::: {.callout-note title="Why Add Covariates to a Cox Model?" collapse="true"}
Adding covariates to a Cox model helps you:
- adjust for confounding
- estimate effects conditional on other variables
- improve the model if the covariates are genuinely prognostic
- check whether the treatment result changes once other predictors are included

In practice, covariates should be added because they are clinically or
substantively relevant, not just because they happen to be significant in one
dataset.
:::

## Model Building

From here, the workflow becomes model building: add one or two variables at a
time, check what changes, and keep revisiting the assumptions.

::: {.callout-tip title="What is a multivariable Cox model and why are we using it?" collapse="true"}
A univariable Cox model looks at one predictor at a time. A multivariable Cox
model includes several covariates at once, which lets you estimate adjusted
effects and see whether an apparent treatment signal survives once other
predictors are included.
:::

::: {.callout-tip title="Why not use logistic regression?" collapse="true"}
Logistic regression ignores time-to-event and censoring. When working
with survival data, Cox models are preferred as they preserve the time
structure and allow for censored cases.
:::

### Continuous Variable: Age

Age is a good first continuous covariate to add because it is clinically
plausible, easy to interpret, and useful for illustrating functional-form
checks.

::: {.callout-tip title="Modeling Continuous Covariates" collapse="true"}
Continuous variables are often where Cox models start to go wrong quietly. A
linear age term assumes that each one-unit increase has the same effect on the
log hazard across the whole range. That is convenient, but not always realistic,
so it is worth checking rather than assuming.
:::

**Cox Model with Age**

```{r}
cox_age <- coxph(Surv(time, status) ~ trt + age, data = veteran)
summary(cox_age)
```

**Interpretation of the Cox Model with Age**

Adding age to the treatment model does not materially change the result. The
treatment effect remains close to null (HR = 1.00, 95% CI: 0.70–1.43, *p* =
0.984), and age itself shows only a weak association with hazard (HR = 1.01 per
year, 95% CI: 0.99–1.03, *p* = 0.436). Concordance remains poor
(concordance = 0.514), so this specification adds little explanatory value.

The practical takeaway is that age is worth checking, but a simple linear age
term is not doing much for the model yet.

::: {.callout-note title="Interpreting the Cox Model Output" collapse="true"}
Understanding the `coxph()` output is essential for interpreting
survival models. Here's a breakdown of the most common components you'll
encounter in both univariable and multivariable Cox regressions:

**🔹 *coef***\
This is the **log hazard ratio** — the raw estimate from the model. It
represents the change in the **log hazard** of the event (e.g., death)
associated with a **one-unit increase** in the covariate (for continuous
variables) or **relative to a reference group** (for categorical
variables).

-   Positive `coef` → increased hazard (i.e., higher risk)\
-   Negative `coef` → decreased hazard (i.e., protective)\
-   A `coef` close to 0 suggests little to no effect

🧠 *Note:* Log hazards are difficult to interpret directly, so we
usually exponentiate them.

**🔹 *exp(coef)***\
This is the **hazard ratio (HR)** — the exponentiated `coef`, which
expresses effect size more intuitively.

-   `exp(coef) > 1` → Increased hazard\
-   `exp(coef) < 1` → Decreased hazard\
-   `exp(coef) = 1` → No effect

🔎 Example: `exp(coef) = 1.20` means a 20% higher hazard.\
`exp(coef) = 0.80` means a 20% lower hazard.

**🔹 *se(coef)***\
The **standard error** of the coefficient — reflects how much
uncertainty surrounds the estimate.

-   Smaller values = more precise estimates\
-   Larger values = more uncertainty (due to small samples or noisy
    data)

**🔹 *z* and *Pr(\>\|z\|)***\
- `z` is the **Z-statistic**, calculated as `coef / se(coef)`\
- `Pr(>|z|)` is the **p-value** from the Wald test for each coefficient

This tests the **null hypothesis** that the covariate has **no effect**
(`coef = 0`).

-   `p < 0.05` → Statistically significant\
-   `p > 0.05` → Not statistically significant

**📊 *Model-Level Statistics***

**🔸 *Concordance***\
- Measures the model's ability to correctly rank survival times (similar
to AUC for classification)\
- Ranges from `0.5` (random) to `1.0` (perfect)\
- In clinical data, values of `0.6–0.75` are typical

**🔸 *Likelihood Ratio Test***\
- Compares the full model to a null (intercept-only) model\
- Tests overall model fit\
- `p < 0.05` = your model significantly improves fit

**🔸 *Wald Test***\
- Aggregates the individual Wald tests across all covariates\
- Tests whether any coefficients differ significantly from 0

**🔸 *Score (Logrank) Test***\
- Rank-based test comparing observed vs expected survival\
- Can be slightly more sensitive in small samples\
- Often used as an overall model check

✅ *In practice:* Use all three global tests and the concordance to
evaluate how well the model explains variation in survival. If all are
non-significant, the model may lack predictive utility.
:::

**Schoenfeld Residuals for Age**

```{r}
# Test proportional hazards assumption for the model including age
cox_zph_age <- cox.zph(cox_age)
print(cox_zph_age)
```

Proportional hazards were tested using Schoenfeld residuals. The
assumption was met for **age** (`p = 0.195`) but marginally violated for
**treatment group** (`p = 0.055`), which is slightly below the
conventional threshold of 0.05. The **global test** for the model
(`p = 0.045`) was also just below 0.05, indicating a **possible overall
violation** of the proportional hazards assumption.

While these results do not strongly indicate a breakdown of the PH
assumption, the borderline p-values warrant caution. A visual inspection
of the residual plots (`plot(cox.zph())`) can help determine whether the
trend is substantial or random variation.

::: {.callout-note title="Interpreting Schoenfeld Residuals" collapse="true"}
Schoenfeld residuals are used to test the **Proportional Hazards (PH)
assumption** in Cox models — that the effect of each covariate remains
**constant over time**.

**How to interpret the output:**

-   Each row (e.g., `trt`, `age`) gives a **test of proportionality**
    for that specific variable.
-   The `GLOBAL` row tests **all covariates together**.

**What the numbers mean:**

-   `chisq` is the test statistic (larger values = more evidence of
    time-varying effects).
-   `p` is the p-value testing **null hypothesis: proportional hazards
    holds**.

**Rule of thumb:** - `p < 0.05` → **PH assumption violated** (bad) -
`p > 0.05` → **PH assumption holds** (good)

🧠 *Note:* Slight violations (e.g., `p = 0.055`) might not be a problem
in practice — but major ones (e.g., `p = 0.01`) should be addressed.
:::

```{r}
# Visual check
plot(cox_zph_age)
```

The Schoenfeld residual plot for age, derived from the multivariable Cox
model, reveals no strong evidence of violation of the proportional
hazards assumption. The smoothed coefficient curve remains relatively
flat across time, with the 95% confidence band consistently overlapping
zero. This suggests that the effect of age on the hazard function does
not meaningfully vary with time. This visual impression aligns with the
result from the formal statistical test (χ² = 1.68, df = 1, p = 0.195),
providing no significant indication that the proportional hazards
assumption has been violated for age in this model.

::: {.callout-note title="How to Interpret Schoenfeld Residual Plots" collapse="true"}
This plot checks the **Proportional Hazards (PH) assumption** visually
for a covariate (e.g., `age`).

**What we are seeing:**

-   **X-axis:** Time\
-   **Y-axis:** Estimated **effect (beta)** of the covariate at each
    time point\
-   **Dots:** Residuals at each event time\
-   **Solid line:** Smoothed trend of the coefficient over time\
-   **Dashed lines:** 95% confidence interval of that trend

**What we want to see:**

✅ A **flat line near zero**, with the confidence band **containing
zero** throughout time → suggests **PH assumption holds**.

⚠️ A **wavy trend** or curve **drifting far from zero** → suggests
**non-proportional hazards** (i.e., the effect of the covariate changes
over time).

**Interpreting the plot:**

If the smoothed line is approximately flat and the confidence interval
consistently includes zero — with no strong patterns or directional
trends — then the PH assumption is likely satisfied for that covariate.
:::

**Martingale Residuals for Age to Assess Functional Form**

To assess whether age is appropriately modeled as a linear predictor in
the Cox model, we examine Martingale residuals plotted against age to
detect non-linearity in its functional form.

```{r}
# Generate Martingale residuals
veteran$martingale_age <- residuals(cox_age, type = "martingale")

# Generate linear predictor
veteran$linear_pred_age <- predict(cox_age, type = "lp")

# Plot residuals against age
ggplot(veteran, aes(x = age, y = martingale_age, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Age",
    x = "Age (years)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Standard" = "#A569BD", "Experimental" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

Martingale residuals plotted against age showed no clear non-linear
trend, with residuals generally scattered randomly across the range of
age values. This suggests that modeling age as a linear term in the Cox
model is reasonable. There is no strong evidence to support the need for
transformations or spline-based modeling of age in this dataset.

::: {.callout-note title="How to Interpret Martingale Residuals for Functional Form" collapse="true"}
This plot helps assess whether the **relationship between a continuous
variable** (like *age*) and the **log hazard** is **linear**, which is a
core assumption of the Cox model.

-   **X-axis:** The covariate
-   **Y-axis:** Martingale residuals (bounded between -∞ and 1)
-   **Points:** One per individual, representing how well their survival
    time is predicted
-   **Dashed line at 0:** Reference line to help spot patterns

**What to look for:** - A **random scatter** of points across the X-axis
with **no clear curve** → linearity assumption is likely satisfied ✅ -
A **systematic curve**, bend, or wave → suggests the effect of age might
be **non-linear** ⚠️\
In that case, consider modeling age using splines or transformations
(e.g., `log()` or `^2`).

**Important:** These residuals are noisier near their bounds and should
be interpreted cautiously, especially when sample size is limited.
:::

**Functional Form Check**

```{r}
# Fit a univariable Cox model to assess functional form of age
cox_age_only <- coxph(Surv(time, status) ~ age, data = veteran)

# Check the functional form visually using Martingale residuals
ggcoxfunctional(cox_age_only, data = veteran)
```

::: {.callout-note title="Assessing Functional Form with Martingale Residuals" collapse="true"}
**What this plot shows:**\
This diagnostic visualizes whether a **continuous covariate** (e.g.,
*age*) has a **linear relationship** with the **log hazard**, as assumed
by the Cox model.

**X-axis:**\
The values of the continuous variable being assessed (e.g., age)

**Y-axis:**\
The **Martingale residuals** from a **null Cox model** (i.e., a model
with no covariates)

**Dots:**\
Each dot represents one observation

**Smooth black line:**\
A **non-parametric smoother** that helps reveal trends in the residuals
— this is the key to checking the shape of the relationship

**How to interpret it:**

-   ✅ If the **black line is roughly straight and flat**, this suggests
    a **linear relationship** between the variable and the log hazard —
    which is **what the Cox model assumes**.
-   ⚠️ If the line shows a **curved or non-linear pattern**, this
    suggests the variable may not be appropriately modeled with a simple
    linear term.
    -   This can indicate the need to use **transformations** (e.g.,
        `log(age)`, `age^2`) or **splines**.

**Tip:**\
Small wobbles are not a major concern — we are primarily looking for
clear trends, not minor bumps. Major **U-shapes, inflections, or long
curves** are the most concerning signs.
:::

The functional-form plot suggests that age may not relate to the log hazard in a
strictly linear way. The curve looks mildly U-shaped, which is enough to keep
non-linearity on the table.

At this point, the practical decision is not "age is important" so much as "a
simple linear age term may be too simplistic." That makes a spline-based model
worth testing next.

### Binary Variable: Prior Treatment

Prior therapy is a useful example of a binary adjustment covariate. It does not
need a functional-form check, but it still needs the usual proportional hazards
check.

::: {.callout-tip title="Why test binary covariates like 'prior'?" collapse="true"}
Even though `prior` is binary, its inclusion allows us to control for
historical differences in patient treatment, which may affect survival
independently of the trial treatment (`trt`).

For binary variables, we don't need to assess the functional form, but
we **do need to check the proportional hazards assumption** (as with any
covariate in the Cox model). We'll interpret the hazard ratio as the
relative difference in risk between those who received prior therapy and
those who did not.
:::

**Fit Cox Model with Prior Treatment**

```{r}
# Fit the Cox model including prior treatment
cox_prior <- coxph(Surv(time, status) ~ trt + prior, data = veteran)
summary(cox_prior)
```

Adding prior therapy does not improve the model. The treatment effect stays
close to null (HR = 1.03, 95% CI: 0.72–1.46, *p* = 0.885), and prior therapy is
also not associated with survival (HR = 0.99, *p* = 0.471). Concordance remains
low at 0.517, so this model still has very limited discriminatory value.

The practical takeaway is that prior therapy is easy to include, but in this
dataset it does not add much on its own.

**Schoenfeld Residuals for Prior Treatment**

```{r}
# Test proportional hazards assumption for the model including prior treatment
cox_zph_prior <- cox.zph(cox_prior)
print(cox_zph_prior)
```

The global Schoenfeld test is borderline (*p* = 0.048), while the individual
tests for treatment and prior therapy are both just above 0.05. That is enough
to look carefully at the residual plots, but not enough to conclude that the
assumption has clearly failed.

```{r}
# Visual check
plot(cox_zph_prior)
```

The residual plot for prior therapy is broadly flat, which supports a stable
effect over time. Treatment looks a bit wavier, but not strongly enough to force
a different modelling decision at this stage.

**Martingale Residuals for Prior Treatment**

```{r}
# Generate Martingale residuals
veteran$martingale_prior <- residuals(cox_prior, type = "martingale")
# Generate linear predictor
veteran$linear_pred_prior <- predict(cox_prior, type = "lp")
# Plot residuals against prior treatment
ggplot(veteran, aes(x = prior, y = martingale_prior, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Prior Treatment",
    x = "Prior Treatment (0 = No, 1 = Yes)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Standard" = "#A569BD", "Experimental" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) # Adding jitter to avoid overlap

```

Because prior therapy is binary, this plot is mainly another coding/sanity check.
The residual spread looks similar across the two groups, so there is no obvious
sign that the binary specification is problematic.


### Multivariable Cox Model with Age and Prior Treatment

Now that age and prior therapy have both been examined separately, the next step
is to include them together with treatment in a simple multivariable model.

```{r}
# Fit the multivariable Cox model including age and prior treatment
cox_multivariable <- coxph(Surv(time, status) ~ trt + age + prior, data = veteran)
summary(cox_multivariable)
```

This model does not materially improve on the earlier ones. Treatment remains
null (HR = 1.00, *p* = 0.984), age remains weak (*p* = 0.462), and prior
therapy remains uninformative (*p* = 0.500). Concordance is 0.507, which is a
good reminder that adding more variables does not automatically make the model
more useful.

**Schoenfeld Residuals for Multivariable Model**

```{r}
# Test proportional hazards assumption for the multivariable model
cox_zph_multivariable <- cox.zph(cox_multivariable)
print(cox_zph_multivariable)
```

The global Schoenfeld test is significant (*p* = 0.032), although none of the
individual covariates crosses the 0.05 threshold. That pattern is worth noting,
but by itself it does not tell you which term is causing trouble.

```{r}
# Visual check
plot(cox_zph_multivariable)
```

The residual plots are more reassuring than the global test alone. Age looks
stable, while treatment and prior therapy show mild undulation but no strong
sustained trend. In practice, I would treat proportional hazards as workable
here, while keeping an eye on treatment in later models.

**Martingale Residuals for Multivariable Model**

```{r}
# Generate Martingale residuals
veteran$martingale_multivariable <- residuals(cox_multivariable, type = "martingale")
# Generate linear predictor
veteran$linear_pred_multivariable <- predict(cox_multivariable, type = "lp")
# Plot residuals against linear predictor
ggplot(veteran, aes(x = linear_pred_multivariable, y = martingale_multivariable, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Linear Predictor\n(Multivariable Model)",
    x = "Linear Predictor (Log Hazard)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Standard" = "#A569BD", "Experimental" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The residuals against the linear predictor do not show a strong pattern, which
argues against major misspecification in the combined model. The narrow spread of
predicted values also reinforces the main point: this model is not separating
patients into meaningfully different risk levels.

**Functional Form Check for Continuous Covariates**

We now recheck the functional form of age in the multivariable setting.

```{r}
# Refit a univariable model with age (again, for functional form check)
cox_age_for_form_check <- coxph(Surv(time, status) ~ age, data = veteran)

# Plot functional form using ggcoxfunctional
ggcoxfunctional(
  fit = cox_age_for_form_check,
  data = veteran,
  point.col = "#E74C3C",
  ggtheme = theme_minimal(base_size = 13),
  title = "Functional Form Check: Age"
)
```

The age functional-form check still suggests curvature rather than a clean
linear effect. That is enough reason to test a spline representation next.

::: {.callout-tip title="Why Check Functional Form Again in a Multivariable Model?" collapse="true"}
While functional form checks (e.g., using Martingale residuals) are
**only necessary for continuous covariates**, it's important to
**reassess them in the context of the full model** — even if they passed
in a univariable setting.

-   In a **univariable Cox model**, we check whether the continuous
    variable (e.g., age) has a linear relationship with the log hazard.
-   However, once we add other covariates (like treatment or prior
    therapy), the relationship can **change due to confounding or
    interactions**.
-   A variable might look linear on its own, but **show non-linearity
    when adjusted** for other predictors.

So we recheck **only the continuous variables** (e.g., age) in the
multivariable model to ensure the assumption still holds.
:::

### Multivariable Cox Model with Spline-Transformed Age

Once age looks even mildly non-linear, a spline model is usually the next thing
to try. The point is not to force complexity for its own sake, but to see
whether a more flexible age term fits the data better.

::: {.callout-note title="Understanding Spline and Non-Linear Transformations in Survival Models" collapse="true"}
Spline terms are a practical way to relax the linearity assumption for a
continuous predictor without having to guess the exact shape of the curve.

In this notebook, `ns(age, df = 3)` is used as a flexible age term. That is a
reasonable choice when:
- the residual plots suggest curvature
- a simple linear term looks too restrictive
- you want something more stable than arbitrary age categories

The tradeoff is that spline terms are less interpretable coefficient by
coefficient, so they are usually justified by better fit and better diagnostics
rather than by the significance of any one basis term.
:::

**Model Output**

```{r}
# Fit a Cox model with a natural spline for age
cox_spline <- coxph(Surv(time, status) ~ ns(age, df = 3) + trt + prior, data = veteran)
summary(cox_spline)
```

Using a spline for age improves flexibility, but it does not suddenly make this a
strong model. Treatment and prior therapy remain uninformative, and the spline
terms are not especially strong one by one. Even so, concordance improves
slightly to 0.559, and the spline specification is more consistent with the
functional-form checks.

**Schoenfeld Residuals for Spline Model**

```{r}
# Test proportional hazards assumption for the spline model
cox.zph(cox_spline)
```

The spline model looks better on proportional hazards diagnostics. The global
test is non-significant (*p* = 0.116), and none of the individual terms gives a
clear signal of time-varying effects.

```{r}
# Visual check of Schoenfeld residuals for spline model
plot(cox.zph(cox_spline))
```

The residual plots are broadly reassuring. None of the spline components, nor
the treatment or prior-therapy terms, shows a strong sustained trend over time.

**Martingale Residuals for Spline Model**

```{r}
# Residuals vs linear predictor
veteran$martingale_spline <- residuals(cox_spline, type = "martingale")
veteran$linear_pred_spline <- predict(cox_spline, type = "lp")

ggplot(veteran, aes(x = linear_pred_spline, y = martingale_spline, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Linear Predictor\n(Spline Model)",
    x = "Linear Predictor (Log Hazard)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Standard" = "#A569BD", "Experimental" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The residuals against the linear predictor look broadly patternless, which
suggests the spline model has addressed the main functional-form concern without
introducing obvious new misspecification.

### Model Comparison: Linear vs Spline Age Term

We now compare the original multivariable Cox model with age modeled
linearly to the updated model using a spline transformation for age.
This allows us to formally test whether the added flexibility of the
spline significantly improves model fit.

```{r}
# Compare linear age model vs spline-transformed age model
anova(cox_multivariable, cox_spline, test = "LRT")
```

The likelihood ratio test shows that the spline model fits better than the
linear-age model (χ² = 7.91, df = 2, *p* = 0.019). In a workflow like this, that
is usually enough to keep the spline version going forward.

**Visualizing the Effect of Spline-Transformed Age**

This section uses 'termplot()' to visualize how the hazard changes
across age when modeled as a spline. This provides an intuitive look at
the non-linear relationship.

::: {.callout-note title="Understanding ANOVA in Cox Regression" collapse="true"}
In this context, ANOVA is just a likelihood-ratio comparison of nested Cox
models. The practical question is simple: does the more flexible model improve
fit enough to justify the extra complexity?
:::

```{r}
termplot(cox_spline, se = TRUE, col.term = "blue")
```

The term plots from the spline-based Cox proportional hazards model
illustrate the partial effect of each covariate on the log hazard:

-   Age (modeled using a natural spline with 3 degrees of freedom)
    exhibits a non-linear association with the log hazard. The curve
    displays a shallow U-shape, with the lowest hazard around age 55,
    and a gradual increase in hazard at both younger and older ages.
    While the trend is suggestive, the wide confidence
    intervals—particularly at the distribution tails—indicate
    uncertainty in the precise shape of this relationship.

-   Treatment Group (trt) is represented as a binary variable with two
    flat lines, one for each group. The narrow confidence intervals
    overlap with zero, implying no statistically significant difference
    in log hazard between the experimental and standard arms in this
    model.

-   Prior Treatment (prior) also shows a flat trend across the range of
    prior treatments, with confidence bands consistently containing
    zero. This suggests that the number of prior therapies does not
    contribute meaningfully to hazard estimation when adjusting for age
    and treatment.

These visualizations support the statistical findings from the model
summary. Age is the only covariate showing a potentially complex,
non-linear effect, while treatment and prior exposure show no clear
independent impact on survival in this cohort.

::: {.callout-note title="How to Interpret Term Plots in Cox Models" collapse="true"}
**Term plots** visualize the *partial effect* of each covariate in a
fitted model while holding other variables constant. They are especially
helpful for interpreting transformations like splines.

**🔹 What is a term plot?**\
A term plot shows how the **log hazard** (on the y-axis) changes with a
covariate (on the x-axis), after accounting for all other variables in
the model.

-   The **blue line** is the estimated partial effect of that variable.
-   The **dashed orange lines** show the 95% confidence interval.
-   The interpretation depends on the type of variable:

**🔸 Continuous Variables (e.g., Age):**

-   For **linear terms**, the blue line should be roughly straight.
-   For **splines** or transformed terms:
    -   A *curved* blue line indicates a non-linear relationship.
    -   Areas where the confidence band **does not include 0** suggest a
        statistically significant effect on the log hazard.
    -   Patterns like U-shapes, inflection points, or steep climbs
        reveal how risk evolves across the covariate range.

**🔸 Binary Variables (e.g., Treatment, Prior Therapy):**

-   These are shown as **two horizontal segments**:
    -   One for each group (e.g., Standard = 0, Experimental = 1).
    -   The **difference between the two** shows the log hazard shift
        due to that variable.
-   If the confidence intervals **include zero**, there's no strong
    evidence that the variable affects the hazard.

**✅ Best Uses:**

-   Confirming that spline-transformed terms behave as expected.
-   Checking if a linear term should be transformed (e.g., age might be
    better modeled with splines).
-   Double-checking whether binary variables have any meaningful effect.

🧠 *Tip:* Term plots don't replace statistical tests, but they
**visualize model estimates** clearly and intuitively.
:::

### Multivariable Cox Model with Karnofsky Score and Cell Type

This is the point where the model becomes clinically more informative. Karnofsky
score is an obvious candidate because functional status is a strong prognostic
factor in oncology, and cell type is included because the major histologic
subtypes have meaningfully different disease behaviour.

Age stays in the model as a spline because the earlier checks supported a
non-linear specification.

```{r}
# Fit a multivariable Cox model including Karnofsky score and Cell Type
cox_karno_and_cell <- coxph(
  Surv(time, status) ~ ns(age, df = 3) + trt + prior + karno + celltype,
  data = veteran)

summary(cox_karno_and_cell)
```

This is the first model in the notebook that clearly starts to explain survival
well. Karnofsky score is a strong independent predictor: each 1-point increase
is associated with a 3.4% reduction in hazard (HR = 0.966, 95% CI: 0.955–0.977,
*p* < 0.001).

Cell type also matters. Relative to squamous cell carcinoma, both small cell
carcinoma (HR = 2.21, 95% CI: 1.31–3.75, *p* = 0.003) and adenocarcinoma
(HR = 3.30, 95% CI: 1.82–5.96, *p* < 0.001) are associated with substantially
higher hazard. Large cell carcinoma is not clearly different from squamous.

Treatment remains non-significant, though the point estimate trends toward harm
(HR = 1.40, 95% CI: 0.94–2.09, *p* = 0.097), and prior therapy still contributes
little. The overall model performs much better than the earlier versions, with a
concordance of 0.735.

The practical lesson here is that a Cox model often becomes useful only once the
right prognostic variables are included. In this dataset, functional status and
cell type do far more work than treatment, age, or prior therapy alone.

**Forest Plot of Multivariable Cox Model**
```{r}
# Create a forest plot for the multivariable Cox model
library(forestmodel)
forest_model(cox_karno_and_cell)
```

*Survival Curves by Karnofsky Score*

```{r}
# Plot survival curves stratified by Karnofsky score
# Bin Karnofsky scores into 3 groups
veteran$karno_group <- cut(veteran$karno, 
                           breaks = c(0, 50, 70, 100),
                           labels = c("Low (≤50)", "Medium (51–70)", "High (>70)"),
                           right = TRUE)

# Fit Kaplan-Meier survival model by Karnofsky group
km_karno <- survfit(Surv(time, status) ~ karno_group, data = veteran)

# Plot using ggsurvplot
ggsurvplot(km_karno,
           data = veteran,
           risk.table = TRUE,
           pval = TRUE,
           conf.int = FALSE,
           legend.title = "Karnofsky Group",
           legend.labs = c("Low (≤50)", "Medium (51–70)", "High (>70)"),
           palette = c("firebrick", "steelblue", "darkgreen"),
           xlab = "Time (days)",
           ylab = "Survival Probability",
           title = "Kaplan-Meier Survival by Karnofsky Score Group")

```

*Survival Curves by Cell Type*

```{r}
# Create Kaplan-Meier curves by cell type
km_cell <- survfit(Surv(time, status) ~ celltype, data = veteran)

# Plot the survival curves
ggsurvplot(km_cell,
           data = veteran,
           risk.table = TRUE,         # Show risk table
           pval = TRUE,               # Show log-rank p-value
           conf.int = FALSE,          # Hide confidence intervals if cluttered
           legend.title = "Cell Type",
           legend.labs = levels(factor(veteran$celltype)),
           palette = "Dark2",         # Choose a nice palette
           xlab = "Time (days)",
           ylab = "Survival Probability",
           title = "Kaplan-Meier Survival by Cell Type")

```

*Combined Plot of Survival Curves*



```{r}
# For Cell Type
km_cell <- survfit(Surv(time, status) ~ celltype, data = veteran)
plot_cell <- ggsurvplot(km_cell,
                        data = veteran,
                        risk.table = TRUE,
                        pval = TRUE,
                        conf.int = FALSE,
                        legend.title = "Cell Type",
                        palette = "Dark2",
                        xlab = "Time (days)",
                        ylab = "Survival Probability",
                        title = "Survival by Cell Type")

# For Karnofsky Score Group
km_karno <- survfit(Surv(time, status) ~ karno_group, data = veteran)
plot_karno <- ggsurvplot(km_karno,
                         data = veteran,
                         risk.table = TRUE,
                         pval = TRUE,
                         conf.int = FALSE,
                         legend.title = "Karnofsky Score",
                         palette = c("firebrick", "steelblue", "darkgreen"),
                         xlab = "Time (days)",
                         ylab = "Survival Probability",
                         title = "Survival by Karnofsky Group")
# Combine the two plots

combined_plot <- arrange_ggsurvplots(list(plot_cell, plot_karno),
                                     ncol = 2, nrow = 1)
combined_plot
```

**Proportional Hazards Assumption: Formal Test**

```{r}
# Test proportional hazards assumption for the final model
cox_zph_final <- cox.zph(cox_karno_and_cell)
print(cox_zph_final)
```

The proportional hazards assumption was assessed for the full multivariable
model using Schoenfeld residuals via `cox.zph()`. None of the individual
covariates — including the spline terms for age, Karnofsky score, cell type
categories, treatment group, and prior therapy — reached statistical
significance, and the global test was non-significant. This provides no
compelling evidence against the proportional hazards assumption for any term
in the model.

**Schoenfeld Residuals: Visual Check**

```{r}
# Visual check of Schoenfeld residuals for final model
plot(cox_zph_final)
```

Visual inspection of the Schoenfeld residual plots supports the formal test
results. The smoothed trend lines for each covariate remain broadly flat
across the follow-up period, with no systematic drift upward or downward over
time. Confidence bands are wide — reflecting the modest sample size — but the
observed trends do not suggest meaningful time-varying effects for any
predictor. Taken together, the statistical and visual evidence supports the
conclusion that the proportional hazards assumption is reasonably satisfied in
the final model.

**Martingale Residuals: Overall Model Fit**

```{r}
# Generate Martingale residuals for the final model
veteran$martingale_final <- residuals(cox_karno_and_cell, type = "martingale")
veteran$linear_pred_final <- predict(cox_karno_and_cell, type = "lp")
ggplot(veteran, aes(x = linear_pred_final, y = martingale_final, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Linear Predictor\n(Final Model)",
    x = "Linear Predictor (Log Hazard)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Standard" = "#A569BD", "Experimental" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The Martingale residuals plotted against the linear predictor show random
scatter around zero with no systematic curvature or directional trend,
supporting adequate overall model specification. Compared to the earlier
model with only treatment, age, and prior therapy, the linear predictor now
spans a much wider range (approximately −2 to 1.5), reflecting the stronger
prognostic signal captured by Karnofsky score and cell type. The two
treatment groups overlap substantially across this range, consistent with
the non-significant treatment effect observed in the model. A small number
of points show larger negative residuals, representing patients who survived
considerably longer than the model predicted — a pattern common in datasets
with high event rates and limited censoring.

**Functional Form Check: Karnofsky Score**

```{r}
# Refit a univariable model with Karnofsky score for functional form check
cox_karno_for_form_check <- coxph(Surv(time, status) ~ karno, data = veteran)
# Plot functional form using ggcoxfunctional
ggcoxfunctional(
  fit = cox_karno_for_form_check,
  data = veteran,
  point.col = "#E74C3C",
  ggtheme = theme_minimal(base_size = 13),
  title = "Functional Form Check: Karnofsky Score"
)
```

The Martingale residual plot for Karnofsky score shows a broadly linear and
decreasing trend — as Karnofsky score increases, the residuals decline,
indicating that higher functional status is associated with lower observed
hazard relative to a null model. The relationship is reasonably well-captured
by a linear term, with no strong U-shaped or threshold pattern that would
require a spline transformation. This supports modeling Karnofsky score as a
continuous linear predictor in the final model.

::: {.callout-tip title="Functional Form: Karno vs Age" collapse="true"}
Recall that the functional form check for age revealed a U-shaped curve,
leading us to apply a natural cubic spline (`ns(age, df = 3)`). In contrast,
Karnofsky score shows a monotonically decreasing relationship with the
Martingale residuals — consistent with a linear dose-response on the log
hazard scale. This is clinically intuitive: each additional point of
functional status confers a roughly proportional survival benefit, rather
than a threshold or non-monotonic effect. Not every continuous covariate
requires a spline; the functional form check tells us which ones do.
:::



# Interaction or Additive Effects?

In Cox regression, interaction terms allow us to test whether the effect
of one covariate on the hazard changes depending on the level of another
covariate. If significant, this suggests a departure from simple
additivity and may require stratified or more complex modelling.

::: {.callout-note title="Understanding Interaction Effects in Cox Models" collapse="true"}
In regression models—including **Cox proportional hazards models**—an
*interaction effect* tests whether the relationship between a covariate
and the outcome changes depending on the level of another covariate.

In practical terms, interaction terms allow us to ask:\
\> *“Does the effect of Variable A on survival depend on Variable B?”*

*Why Test for Interaction?*

-   To explore whether **treatment effects vary across subgroups**
    (e.g., age, gender, biomarker levels).
-   To identify **non-additive relationships** that a main-effects-only
    model would miss.
-   To **refine risk stratification** or support personalized approaches
    to care.

*What It Tells Us*

In the Cox model, including an interaction (e.g., `treatment * age`)
allows the **hazard ratio for treatment** to vary at different values of
age. If the interaction term is statistically significant, it suggests
the **effect of treatment is modified by age**.

This can reveal: - A treatment that works well in younger patients but
poorly in older ones. - A covariate that only has an impact in the
presence of another factor. - Complex survival dynamics that are
**hidden in simpler additive models**.

*Interpreting Output*

-   A significant interaction coefficient indicates a **departure from
    additivity**.
-   It's essential to **cautiously interpret main effects** in the
    presence of interactions, as their meaning may be conditional.
-   Visualization (e.g., stratified survival curves or effect plots) can
    aid interpretation.

*When to Use*

-   After fitting and interpreting a main-effects model.
-   When clinical theory or prior evidence suggests effect modification.
-   When residual plots or diagnostics hint at **unexplained
    heterogeneity**.
:::

## Interaction of Treatment and Age

```{r}
# Fit a cox model with interaction between treatment and age 
cox_interaction <- coxph(Surv(time, status) ~ trt * ns(age, df = 3) + prior, data = veteran)
summary(cox_interaction)
```

We tested for interaction between treatment group and age (modeled using
a natural spline with 3 degrees of freedom) to explore whether the effect
of treatment on survival varied across age groups. The interaction model
revealed a statistically significant interaction for the second spline
basis term (p = 0.037), suggesting a possible age-dependent treatment
effect.

The overall Wald test for the model approached significance (p = 0.05),
while the likelihood ratio test was marginal (p = 0.07), indicating that
the model with interaction terms may provide a better fit than the
additive model. These findings suggest that the benefit (or lack
thereof) of treatment may differ depending on patient age, warranting
further exploration or stratified modelling.

**Schoenfeld Residuals for Interaction Model**

```{r}
# Test proportional hazards assumption for the interaction model
cox.zph(cox_interaction)
```

The Schoenfeld residual test was used to assess whether the proportional
hazards assumption holds for each covariate and interaction term in the
model.

-   trt: χ² = 1.39, p = 0.24 → No evidence of time-varying effects for
    treatment group.

-   ns(age, df = 3): χ² = 3.90, p = 0.27 → The spline-transformed age
    term shows no strong time dependency.

-   prior: χ² = 2.42, p = 0.12 → Also not statistically significant.

-   trt × ns(age, df = 3): χ² = 5.37, p = 0.15 → The interaction between
    age and treatment shows no significant violation of proportionality
    either.

-   GLOBAL test: χ² = 8.94 on 8 df, p = 0.35 → The model as a whole
    satisfies the PH assumption.

There is no strong evidence that the proportional hazards assumption is
violated in this interaction model. The covariate effects, including the
interaction, appear stable over time.

```{r}
# Visual check of Schoenfeld residuals for interaction model
plot(cox.zph(cox_interaction))
```

The global test using cox.zph() indicated no significant violation of
the PH assumption (χ² = 8.94, df = 8, p = 0.35). Individual covariate
tests also returned non-significant p-values, suggesting that the
estimated hazard ratios remain approximately constant over time.

Visual inspection of the Schoenfeld residual plots supported this
conclusion. The smoothed curves for treatment, age (spline terms), prior
treatment, and their interaction remained relatively flat and within
their 95% confidence bands. While a few outliers were
present—particularly in the treatment and age components—these did not
indicate strong evidence of non-proportionality.

Taken together, these results suggest that the proportional hazards
assumption is reasonably met for all terms in the interaction model.

**Martingale Residuals for Interaction Model**

The Martingale residuals from the interaction model were plotted against
the linear predictor (log hazard) to assess model fit and identify any
potential patterns or outliers.

```{r}
# Residuals vs linear predictor
veteran$martingale_interaction <- residuals(cox_interaction, type = "martingale")
veteran$linear_pred_interaction <- predict(cox_interaction, type = "lp")
ggplot(veteran, aes(x = linear_pred_interaction, y = martingale_interaction, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Linear Predictor\n(Interaction Model)",
    x = "Linear Predictor (Log Hazard)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Standard" = "#A569BD", "Experimental" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The Martingale residuals plot from the interaction model reveals a
concentration of predicted log hazard values for the control group
between 0.5 and 1.0, whereas predictions for the intervention group span
a wider range. This may reflect the influence of the age-by-treatment
interaction term, which adds complexity primarily for patients receiving
the intervention. The residuals themselves are randomly scattered around
zero, suggesting that the functional form of the covariates is
appropriately specified and that no major nonlinearity remains
unaccounted for.

**Deviance Residuals for Interaction Model**

Deviance residuals were generated to further assess the model fit and
identify any potential outliers or influential observations.

::: {.callout-note title="Understanding Deviance Residuals in Cox Models" collapse="true"}
**Deviance residuals** are a type of diagnostic measure used to assess
how well a Cox proportional hazards model fits the observed survival
data. Each deviance residual reflects the discrepancy between the
observed outcome (event or censoring) and what the model predicts for an
individual, after accounting for all covariates.

🔍 Why we use them: - **Outlier detection**: Large residuals (positive
or negative) may indicate subjects whose survival times were poorly
predicted by the model. - **Model fit check**: If residuals cluster
systematically or display non-random patterns when plotted against the
linear predictor, it may suggest issues such as: - Mis-specification of
functional form (e.g., nonlinear effects not properly captured), -
Omitted interactions or variables, - Violation of model assumptions.

📊 How to interpret: - Deviance residuals are approximately normally
distributed around 0. - Values far from zero may indicate poor fit for
specific observations. - A balanced, symmetrical scatter around the
horizontal axis suggests a well-fitting model.

🧠 When to use: - After fitting a Cox model, especially when refining
the covariate structure. - In combination with **martingale** and
**Schoenfeld residuals** to build a comprehensive diagnostic view. -
Before finalizing your model, particularly if you are considering
transformations or interactions.

💡 Bonus tip: Unlike martingale residuals, deviance residuals are
**symmetrically distributed** and therefore better suited for
identifying both under- and over-predictions, making them particularly
useful for graphical diagnostics.
:::

```{r}
# Deviance residuals
veteran$deviance_interaction <- residuals(cox_interaction, type = "deviance")
# Plot deviance residuals
ggplot(veteran, aes(x = linear_pred_interaction, y = deviance_interaction, color = trt)) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) + # Adding jitter to avoid overlap
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Deviance Residuals vs Linear Predictor\n(Interaction Model)",
    x = "Linear Predictor (Log Hazard)",
    y = "Deviance Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Standard" = "#A569BD", "Experimental" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The deviance residuals are also broadly reassuring. Residuals are scattered
around zero with no clear trend, suggesting that the interaction model fits the
data reasonably well. That said, the model is more complex than the additive
alternative and does not deliver a clearly better substantive interpretation.

## Interaction of Treatment and Prior Therapy

Prior treatment is a plausible interaction candidate because previous therapy
could change how patients respond to the trial regimen.

```{r}
# Fit a cox model with interaction between treatment and prior therapy
cox_interaction_prior <- coxph(Surv(time, status) ~ trt * prior + ns(age, df = 3), data = veteran)
summary(cox_interaction_prior)
```


A treatment-by-prior-therapy interaction was tested to see whether the effect of
the experimental regimen differed by prior treatment history. The interaction
term was not statistically convincing (HR = 0.94, 95% CI: 0.87–1.02, *p* =
0.135), and the model offered only weak evidence of improvement overall.

**Schoenfeld Residuals for Interaction with Prior Therapy**

```{r}
# Test proportional hazards assumption for the interaction with prior therapy
cox.zph(cox_interaction_prior)
```

The diagnostics are broadly acceptable. The global Schoenfeld test is
non-significant (*p* = 0.171), and although the interaction term is borderline
(*p* = 0.054), the evidence for time-varying effects is weak.

```{r}
# Visual check of Schoenfeld residuals for interaction with prior therapy
plot(cox.zph(cox_interaction_prior))
```

The residual plots do not show a clear or systematic time trend, so there is no
strong visual evidence against the proportional hazards assumption for this
interaction model.

**Martingale Residuals for Interaction with Prior Therapy**

```{r}
# Residuals vs linear predictor
veteran$martingale_interaction_prior <- residuals(cox_interaction_prior, type = "martingale")
veteran$linear_pred_interaction_prior <- predict(cox_interaction_prior, type = "lp")

ggplot(veteran, aes(x = linear_pred_interaction_prior, y = martingale_interaction_prior, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Linear Predictor\n(Interaction with Prior Therapy)",
    x = "Linear Predictor (Log Hazard)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Standard" = "#A569BD", "Experimental" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The Martingale residuals do not show a strong non-random pattern, which
supports the adequacy of the covariate specification, including the interaction
term.

**Deviance Residuals for Interaction with Prior Therapy**

```{r}
# Deviance residuals
veteran$deviance_interaction_prior <- residuals(cox_interaction_prior, type = "deviance")

# Plot deviance residuals
ggplot(veteran, aes(x = linear_pred_interaction_prior, y = deviance_interaction_prior, color = trt)) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) + # Adding jitter to avoid overlap
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Deviance Residuals vs Linear Predictor\n(Interaction with Prior Therapy)",
    x = "Linear Predictor (Log Hazard)",
    y = "Deviance Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Standard" = "#A569BD", "Experimental" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The deviance residual plot is also broadly reassuring. Residuals are scattered
fairly symmetrically around zero, with no obvious sign of major model misfit. In
practical terms, this interaction is not doing enough to justify the added
complexity.

# Model Comparison: Interaction vs Non-Interaction

Two interaction models were explored: treatment by age and treatment by prior
therapy. The treatment-by-age model showed some signal, but the improvement in
fit was modest and did not materially change the interpretation of the analysis.
The treatment-by-prior-therapy interaction was weaker and not statistically
convincing.

When interactions do not improve either fit or interpretation in a meaningful
way, the simpler additive model is usually the better choice. That is the case
here.

# Final Model Selection and Interpretation

After working through the simpler models, checking functional form, and
exploring a couple of interactions, the final model I would carry forward is:

$$\text{trt} + \text{prior} + ns(\text{age}, df=3) + \text{karno} + \text{celltype}$$

**Why this model?**

- `karno` and `celltype` are clinically important and clearly prognostic in this dataset.
- `age` is better represented with a spline than a simple linear term.
- `prior` is not strong on its own, but it is a reasonable adjustment covariate and does not destabilise the model.
- the tested interactions add complexity without changing the substantive story enough to justify keeping them.
- `diagtime` does not help enough here to earn a place in the final specification.

**Interpretation of the Final Model**

After adjustment, the treatment effect is still not convincing (HR = 1.40, 95%
CI: 0.94–2.09, *p* = 0.097). That is consistent with the earlier unadjusted
comparison.

Karnofsky score is the strongest predictor in the model. Each 1-point increase
is associated with a 3.4% reduction in hazard (HR = 0.966, 95% CI: 0.955–0.977,
*p* < 0.001). Cell type also matters: relative to squamous cell carcinoma, small
cell carcinoma and adenocarcinoma carry substantially higher hazard, while large
cell carcinoma is not clearly different.

Age is better handled non-linearly than linearly, but its effect is not easy to
summarise with a single hazard ratio. Prior therapy still contributes little in
the adjusted model.

The practical lesson is that once the right prognostic variables are in the
model, the treatment story becomes clearer: the main drivers of survival here
are baseline functional status and tumour biology, not the trial treatment.

**Overall Model Performance**

This final model performs much better than the earlier versions, with
concordance improving to 0.735. The global model tests are all strongly
significant, and the diagnostics do not suggest a major assumption problem.

# Conclusion

As a workflow example, this notebook shows a useful pattern for Cox modelling:

- start with Kaplan-Meier curves and a log-rank test
- fit a simple Cox model first
- add covariates gradually
- check proportional hazards and functional form as you go
- use splines when a continuous predictor does not look linear
- test interactions selectively, and keep them only if they materially improve the model

In this dataset, the treatment effect remains weak throughout, while Karnofsky
score and cell type emerge as the key prognostic variables. The final model is
not just the one with the best fit, but the one that balances clinical sense,
diagnostic adequacy, and interpretability.

That is probably the main reusable lesson from this notebook: the value of a Cox
analysis often comes less from the first treatment estimate and more from the
process of building, checking, and refining the model properly.

# Limitations

Even in a guide notebook, it is worth being explicit about what this example can
and cannot support.

**Sample size and generalisability.** The dataset comprises 137 patients from
a single trial, which constrains statistical power for detecting modest
treatment effects and limits the precision of subgroup estimates (e.g.,
hazard ratios by cell type). The small cell carcinoma and adenocarcinoma
groups in particular have relatively few patients, making their hazard ratio
estimates less stable than those for squamous cell carcinoma.

**All-male cohort.** All participants were male veterans. Lung cancer biology,
treatment response, and prognosis differ between sexes, so findings cannot be
assumed to generalise to female patients.

**Historical data.** The trial was conducted in the 1970s–80s. Chemotherapy
regimens, supportive care standards, and diagnostic criteria have changed
substantially since then. Neither the experimental treatment nor the standard
regimen reflects current clinical practice, and the trial predates targeted
therapy, immunotherapy, and modern platinum-based regimens.

**Event rate and censoring.** With 128 deaths among 137 patients (93.4%),
this dataset has very limited censoring. While this provides statistical
power, it leaves little room for assessing time-to-event heterogeneity and
means survival estimates at longer time points are based on very few
at-risk individuals.

**No external validation.** Model performance (concordance = 0.735) was
assessed on the same data used to fit the model. Without an independent
validation cohort, this estimate is likely optimistic and should be
interpreted with caution.

**Residual confounding.** Although the model adjusts for the available
covariates, unmeasured factors — such as comorbidities, disease stage at
diagnosis, smoking history, or treatment compliance — may confound the
observed associations.

**Interaction exploration was limited.** Only two interaction terms were
tested (treatment × age, treatment × prior therapy). It is possible that
clinically meaningful interactions exist between other covariates (e.g.,
treatment × cell type) that were not explored.
