---
title: "Survival Analysis of Veterans with Advanced Lung Cancer: A Cox Proportional Hazards Approach"
author: "Julian Chung"
format:
  html:
    toc: true
    code-fold: true
    number-sections: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

# Introduction

This report presents a survival analysis of patients with advanced,
inoperable lung cancer who participated in a randomized clinical trial
conducted by the Veterans Administration in the United States. The
objective of the study was to compare the effectiveness of a standard
chemotherapy protocol versus a test (experimental) treatment in
prolonging patient survival.

Participants were followed until death or censoring, with overall
survival time (in days from treatment initiation) as the primary
outcome. This dataset, originally introduced in The Statistical Analysis
of Failure Time Data by Kalbfleisch and Prentice (1980), has become a
foundational paper in the teaching survival analysis due to its manageable size,
clear structure, and strong clinical relevance.

This analysis explores survival outcomes using both Kaplan-Meier
estimation and Cox proportional hazards modeling, followed by diagnostic
checks to assess model assumptions and potential improvements. 

# Setup and Data Import

We begin by loading the necessary R packages and importing the veteran
lung cancer dataset from the `survival` package.

```{r}
# Load required libraries
library(survival) # Core survival analysis tools
library(survminer)  # Visualization and diagnostics
library(dplyr)  # Data wrangling
library(ggplot2)  # Plotting
library(broom)  # Tidying model output
library(splines) # Splines for flexible modeling

```

::: {.callout-tip title="Why These Packages?" collapse="true"}
-   **`survival`**: Core package for survival analysis in R. Provides
    `Surv()` to construct survival objects and `coxph()` for Cox models.

-   **`survminer`**: High-level visualization tools for survival curves
    and diagnostics (e.g., `ggsurvplot()`, `ggcoxdiagnostics()`).

-   **`splines`**: Adds support for modeling non-linear relationships in
    covariates using **restricted cubic splines** (e.g., `ns()`).

-   **`ggplot2`**: Used to build layered, customizable plots — often
    integrated with `survminer`.

-   **`dplyr`**: Enables clean and readable data manipulation (e.g.,
    `mutate()`, `filter()`).

-   **`broom`**: Tidies model outputs into data frames (useful for
    summaries, tables, and downstream analysis).
:::

Next, the dataset is loaded and a quick transformation applied to label
the treatment and control groups.

```{r}
# Load the dataset
veteran <- survival::veteran %>%
  mutate(trt = factor(trt,
                      levels = c(1, 2),
                      labels = c("Control", "Intervention")))

# Peek at the structure and summary
# glimpse(veteran)
# summary(veteran)
```

::: {.callout-note title="Dataset Preview" collapse="true"}
We show the first few rows to get a sense of the data structure.

```{r}
head(veteran)
```
:::

**Variable Summary:**

Below is a summary of the key variable in the dataset.

| Variable | Type | Description | Notes |
|------------|------------|-------------------------|------------------------|
| `trt` | Factor | Treatment group: 1 = Standard, 2 = Test | Relabeled to `"Control"` and `"Intervention"` |
| `celltype` | Factor | Histologic type of lung cancer: squamous, small cell, adeno, large | Not used in this model, but useful in extended analyses |
| `time` | Numeric | Survival time in days | Outcome variable in the `Surv()` object |
| `status` | Numeric | Event indicator: 1 = died, 0 = censored | Used to define survival endpoint |
| `karno` | Numeric | Karnofsky performance score (proxy for baseline health status) | May affect prognosis; not modeled in current analysis |
| `diagtime` | Numeric | Time from diagnosis to study entry (in months) | Possibly useful in extended time-dependent models |
| `age` | Numeric | Patient age (in years) | Treated as **continuous**; **checked for linearity and splines** |
| `prior` | Numeric | Number of prior therapies (0 or 10) | Binary variable: 0 = none, 10 = prior treatment |

# Exploratory Survival Analysis

Before proceeding with formal regression modeling, we begin with a non-parametric exploration of the survival data using the Kaplan-Meier estimator. This allows us to visualize survival patterns and group differences without making assumptions about the underlying hazard function. 

::: {.callout-note title="Parametric vs Non-Parametric Methods in Survival Analysis" collapse=true}

**Non-parametric methods**, like the Kaplan-Meier estimator and log-rank test, do **not assume a specific mathematical form** for the distribution of survival times. Instead, they rely solely on the observed data to estimate survival probabilities. These approaches are ideal for **exploratory analysis**, detecting group differences, and visualizing survival trends without imposing structural assumptions.

**Parametric methods**, on the other hand, assume that survival times follow a known probability distribution (e.g., exponential, Weibull). This allows for more efficient estimation and extrapolation — but only if the model fits well. They are powerful for making predictions and simulating outcomes but **can be misleading if the chosen distribution doesn't reflect the true underlying process**.

**Semi-parametric methods**, like the **Cox proportional hazards model**, strike a balance. The Cox model doesn't assume a specific baseline hazard function (non-parametric), but it does assume a **parametric relationship between covariates and the log hazard** (i.e., proportional hazards). This makes it one of the most flexible and widely used tools in survival analysis.

Use **non-parametric** approaches for initial exploration or when we want minimal assumptions.  
Use **(semi-)parametric** models when we want to assess covariate effects or make adjusted predictions.

:::

## Kaplan-Meier Survival Curve

Firstly we estimate the **survival function** for patients using the
Kaplan-Meier estimator. This provides a non-parametric view of survival
over time, stratified by treatment group.

```{r}
# Create the survival object
surv_obj <- Surv(time = veteran$time, event = veteran$status)

# Fit Kaplan-Meier curve stratified by treatment
km_fit <- survfit(surv_obj ~ trt, data = veteran)

# View basic survival summary
# summary(km_fit)

```

::: {.callout-tip title="Survival Summary Truncated for Clarity" collapse="true"}
The full Kaplan-Meier summary includes hundreds of time points. For
readability, we display only the first few rows per treatment group.

```{r}
# View basic survival summary
summary(km_fit)
```
:::

::: {.callout-note title="Why Start with a Kaplan-Meier Curve?" collapse="true"}
The **Kaplan-Meier (KM) curve** is a fundamental starting point in
survival analysis because it allows us to:

-   **Visualize survival patterns** over time,
-   Compare groups (e.g., treatment vs control) using log-rank tests,
-   And understand the *overall shape* and drop-off of survival before
    making model-based assumptions.

The KM method is **non-parametric**, which means it does **not assume
any specific distribution** for survival times. It simply calculates the
probability of surviving past a given time point based on the observed
data.

This makes it ideal as a **descriptive tool** and as a way to **check
for potential group differences** *before* moving on to more complex
models.

By contrast, the **Cox proportional hazards model** is a
**semi-parametric** model. It assumes: - A baseline hazard function
(unspecified), - And that covariates have a *multiplicative effect* on
the hazard over time.

This adds **power and flexibility**, but also introduces assumptions —
like the **proportional hazards** assumption — that must be checked.

So we start with KM to understand the data and get a rough idea of
group differences — without assuming too much — before fitting a Cox
model.
:::

```{r}
# Plot the KM curve
ggsurvplot(
  km_fit,
  data = veteran,
  conf.int = TRUE,
  pval = TRUE,
  pval.size = 4,
  risk.table = TRUE,
  risk.table.title = "Number at Risk",
  risk.table.col = "strata",
  risk.table.y.text.col = TRUE,
  risk.table.fontsize = 4,
  legend.title = "Treatment Group",
  legend.labs = c("Control", "Intervention"),
  xlab = "Time (days)",
  ylab = "Survival Probability",
  palette = c("#A569BD", "#45B39D"),
  title = "Kaplan-Meier Survival Curve: VA Lung Cancer Trial",
  ggtheme = theme_minimal(base_size = 13) + 
    theme(
      legend.position = "top",
      plot.title = element_text(hjust = 0.5, face = "bold"),
      panel.grid.major = element_line(color = "grey80", linewidth = 0.4),
      panel.grid.minor = element_line(color = "grey90", linewidth = 0.2)
    ),
  tables.theme = theme(  # Trying to tweak the risk table
    axis.text.y = element_text(size = 10, margin = margin(r = 12), lineheight = 2)
  )
)
```

**Interpretation of the Kaplan-Meier Plot**

The Kaplan-Meier plot shows the survival curves for patients in the
intervention and control arms. For both groups there is a steep decline
in survival probability during the first 200 days, indicating rapid
disease progression. The shaded areas represent the 95% confidence
intervals, which largely overlap throughout the follow-up period,
suggesting no significant difference in survival between treatment arms.
The log-rank test yields a p-value of 0.93, providing no evidence to
reject the null hypothesis of equal survival. Therefore, we conclude
that there is no statistically significant difference in survival
between the two chemotherapy regimens.

::: {.callout-note title="Interpretation of Graph Elements" collapse="true"}
**X-Axis: Time (days)**

The X-axis shows the time since enrollment in days. The furthest value
to the right (\~1000 days) marks the maximum observed survival time in
the dataset. Shorter durations towards the left reflect early mortality
events.

**Y-Axis: Survival Probability**

The Y-Axis is the Kaplan-Meier estimate of the probability that a
patient survives beyond a certain timepoint. The axis starts at 1 (100%
survival at time 0), and decreases with each event (death).

**Curves (Blue = Standard, Orange = Test)**

Each step downward represents a death. Tick marks would represent a
censored individual (few in this dataset).

**Shaded Area: 95% Confidence Interval**

Around each curve is a shaded band indicating **uncertainty** in the
estimate. Wider intervals to the right of the chart reflect fewer
patients remaining and hence more variability in survival probability.

**p = 0.93**

The log-rank test result tests the null hypothesis: there is no
difference in survival between the groups. The p-value of 0.93 provides
strong evidence *against* rejecting the H0. In other words, there is no
statistically significant difference in survival between the Standard
and Test treatments. If the curves had diverged strongly at some point,
the P-value would be lower.

**"Number at Risk" Table**

This shows how many participatns in each treatment group are still alive
and uncensored at various timepoints (0, 250, 500, 750, 1000). As time
progresses the number declines due to deaths or censoring. Few
participants remain beyond 500 days, which is why the Confidence
Intervals widen later in the graph.

-   The **p-value** is from a log-rank test (H₀: no difference in
    survival curves).
-   The **shaded areas** represent 95% confidence intervals.
-   The **risk table** shows the number of patients still at risk at
    each time point.

Visually, if the curves separate early and remain apart, it suggests a
potential treatment effect.
:::

## Log-Rank Test

The log-rank test is a statistical test used to compare the survival
distributions of two or more groups. It assesses whether there are
significant differences in survival times between the groups, based on
the observed and expected number of events (deaths) at each time point.

::: {.callout-note title="What is the Log-Rank Test?" collapse="true"}
The **log-rank test** is a non-parametric statistical test used to
compare the survival distributions of two or more independent groups. It
evaluates the **null hypothesis** that there is no difference in the
hazard rates between the groups over time—that is, the survival curves
are statistically indistinguishable.

The test works by comparing the **observed number of events (e.g.,
deaths)** in each group to the **expected number of events** under the
assumption that all groups share the same survival function. At each
time point where an event occurs, the log-rank test calculates the
difference between the observed and expected events across groups,
summing these differences over time to produce a test statistic.

This statistic approximately follows a **chi-squared distribution**, and
the resulting p-value indicates whether the observed differences are
likely due to chance. A **small p-value (typically \< 0.05)** suggests a
statistically significant difference in survival between groups.

In the context of **Cox proportional hazards analysis**, the log-rank
test is often used **before fitting a Cox model** to determine whether a
survival difference exists between categorical groups (e.g., treatment
arms). While the log-rank test is limited to univariate group
comparisons and assumes proportional hazards, it provides a simple and
intuitive starting point for exploring potential treatment effects or
group differences in survival.
:::

```{r}
# Perform the log-rank test
log_rank_test <- survdiff(surv_obj ~ trt, data = veteran)
log_rank_test
```

```{r}
# Extract the p-value
1 - pchisq(log_rank_test$chisq, df = length(log_rank_test$n) - 1)
```

::: {.callout-note title="Understanding Chi-Square and Degrees of Freedom" collapse="true"}
The **chi-square (χ²) statistic** is a measure of how much the observed
data deviate from what we would expect under a specific statistical
model — in this case, the assumption that survival curves are the same
across groups.

In tests like the **log-rank test** or **Cox model diagnostics**, the
chi-square statistic is calculated by comparing the **observed** vs
**expected** number of events (deaths) for each group at every time
point. The greater the deviation between these values, the larger the
chi-square statistic becomes.

The **degrees of freedom (df)** represent the number of independent
comparisons being made. For example: - If you're comparing **two
groups**, there’s only one degree of freedom (df = 1). - If you're
comparing **k groups**, df = k – 1.

Once the chi-square statistic and degrees of freedom are known, we can
compute a **p-value** — the probability of observing such a deviation
(or more extreme) if the null hypothesis were true.

-   **Small p-value (e.g., \< 0.05)**: Strong evidence against the null
    hypothesis → suggests a difference between groups.
-   **Large p-value (e.g., \> 0.05)**: No strong evidence → we fail to
    reject the null hypothesis.

In short:\
**Chi-square** quantifies the discrepancy.\
**Degrees of freedom** tell us how many comparisons we’re making.\
**P-value** tells us whether the discrepancy is statistically
meaningful.
:::

**Interpretation of the Log-Rank Test Results**

A log-rank test was performed to compare survival distributions between
the control and intervention treatment groups. The test yielded a
chi-square statistic of 0.008 with 1 degree of freedom, resulting in a
p-value of 0.93. This indicates no statistically significant difference
in survival between the two treatment arms. The observed and expected
number of events (deaths) were nearly identical in both groups, further
supporting the conclusion that survival outcomes did not differ
meaningfully between treatments in this sample.

However, the log-rank test is a univariate method — it does not account for covariates that may confound or modify treatment effects (e.g., age, prior therapy). Additionally, it assumes proportional hazards and constant hazard differences over time. These limitations justify the use of a Cox proportional hazards model, which allows for multivariable adjustment and more nuanced interpretation of the treatment effect in the context of patient characteristics.

::: {.callout-note title="Why Continue Cox Modeling After a Non-Significant Log-Rank Test?" collapse="true"}
Even if the Kaplan-Meier curve and log-rank test show **no statistically
significant difference** between groups, there are several valid and
important reasons to continue with a Cox proportional hazards model:

**The Kaplan-Meier and log-rank test are unadjusted.**\
They assess only one variable at a time and do **not control for
confounders** like age, prior treatment, or performance score. A Cox
model allows you to adjust for these and may reveal effects that were
previously masked.

**Cox regression provides more than just a p-value.**\
Unlike the log-rank test, the Cox model estimates **hazard ratios
(HRs)** and their **confidence intervals**, which are essential for
understanding the **magnitude and direction** of effects — even if not
statistically significant.

**Exploratory analysis and skill building are valuable.**\
In educational or preliminary research contexts, building and diagnosing
models provides critical insight. You may discover interesting patterns
or generate hypotheses for future studies.

**Diagnostics may reveal model violations or improvements.**\
Even in the absence of significant treatment effects, diagnostic tools
(like residual plots or proportional hazards tests) can highlight
**non-linear effects, time-varying hazards, or outliers** that inform
better modelling strategies.

**Bottom line:** The absence of significance in univariate survival
tests does not mean "stop here." A Cox model is not just a hypothesis
test — it's a framework for **adjusted inference**, **deeper insight**,
and **model-based exploration**.
:::

## Why Proceed with the Cox Model?

While the Kaplan-Meier plot and log-rank test indicate no statistically significant difference in survival between treatment arms, these methods do not account for other variables that may affect outcomes. The Cox proportional hazards model allows us to:

- Adjust for confounding covariates (e.g., age, prior treatment)
- Estimate hazard ratios for each variable
- Explore possible interactions or time-varying effects

Thus, we proceed with a Cox model to assess whether treatment effects become more apparent after adjusting for relevant clinical factors.

# Cox Regression Models

The Cox proportional hazards model is a regression method for analysing time-to-event data. Unlike the Kaplan-Meier estimator, which provides a non-parametric, unadjusted view of survival, the Cox model allows us to examine the effect of multiple covariates on the hazard of death.

This model assumes that covariates have a multiplicative and time-constant effect on the hazard (the proportional hazards assumption). By estimating hazard ratios, we can assess whether factors such as treatment type, age, or prior therapy are independently associated with better or worse survival outcomes.

We begin with a simple model including only treatment group, and progressively build toward a multivariable model incorporating continuous and categorical variables. Each model is followed by diagnostic checks to assess assumptions and guide model refinement.

::: {.callout-note title="Understanding the Cox Proportional Hazards Model" collapse="true"}
The **Cox proportional hazards model** (often just called the *Cox
model*) is a cornerstone of survival analysis, allowing us to
investigate how multiple variables influence the **risk (hazard)** of an
event over time — without needing to specify the baseline hazard
function.

**What is a Hazard?**\
The hazard is the **instantaneous risk of the event** (e.g., death) at a
given time, given that the individual has survived up to that point. The
Cox model expresses this as:

> *h(t \| X) = h₀(t) × exp(β₁X₁ + β₂X₂ + ... + βₖXₖ)*

Where: - *h(t \| X)* is the hazard at time *t* for a subject with
covariates *X*, - *h₀(t)* is the **baseline hazard** (unspecified), -
*β* are the **coefficients** estimating the effect of each covariate.

**Why Use the Cox Model Instead of Kaplan-Meier?**

| Kaplan-Meier | Cox Model |
|---------------------------------|---------------------------------------|
| Non-parametric (no assumptions) | Semi-parametric (assumes proportional hazards) |
| Describes survival by **groups only** | Models the **effect of variables** |
| Great for **initial exploration** | Great for **inference and prediction** |
| Can't adjust for **confounders** | Can include **multiple covariates** |
| Visual tool | Inferential and predictive tool |

In practice, it's common to start with a **Kaplan-Meier plot** to
visualize survival, then use a **Cox model** to quantify the effect of
predictors and control for confounding.

**What Does "Proportional Hazards" Mean?**\
The Cox model assumes that the hazard ratio between two individuals is
**constant over time**.\
For example, if a treatment reduces the hazard by 30%, it is assumed to
reduce it by 30% at every point in time. This is the **proportional
hazards assumption**, and it **must be checked** using diagnostic tools
like **Schoenfeld residuals**.

**What the Model Produces:**\
- **Coefficients (β):** Log hazard ratios - **exp(β):** Hazard ratios
(HR), easier to interpret - **p-values:** From the Wald test, testing
whether β ≠ 0 - **Global tests:** Likelihood ratio, Score (logrank),
Wald

Example: - `HR = 1.50` → 50% higher hazard\
- `HR = 0.70` → 30% lower hazard

**Summary:**\
The Cox model is a flexible, interpretable, and robust way to analyze
the effect of variables on survival, especially when you want to
**adjust for multiple covariates**. However, it does require checking
assumptions and being thoughtful about variable selection and
transformations (e.g., linearity for continuous variables).
:::

## Univariable Models

In this section, we fit a series of univariable Cox models to assess the effect of individual covariates on survival. This step helps us explore the independent contribution of each variable before building more complex multivariable models.

### Treatment Group Only

We begin with a univariable Cox model that includes only the treatment
variable (`trt`). This allows us to assess the effect of treatment on
survival without adjusting for any other covariates.

```{r}
# Fit the Cox Proportional Hazards model
cox_model <- coxph(surv_obj ~ trt, data = veteran) # Where trt is a binary treatment variable with levels Control and Intervention

# Summarize the model
summary(cox_model)

```

::: {.callout-note title="Interpretation of Cox Table" collapse="true"}
| Term | Value | What It Means |
|----------|----------|----------------------------------------------------|
| **coef** | 0.01774 | This is the log hazard ratio (log(HR)) comparing the *Intervention* group to *Control*. Close to 0 means almost no effect. |
| **exp(coef)** | 1.01790 | This is the hazard ratio (HR). A value of 1.018 means the hazard (risk of death) is 1.8% higher in the Intervention group compared to Control. However at this level the value is insigificant. |
| **se(coef)** | 0.18066 | The standard error. This gives us the variability around the estimate. |
| **z** | 0.098 | The Wald test statistic. The low value implies no stastical significange. |
| **p-value** | 0.922 | Far above the conventional 0.05 threshold. This is very strong evidence *against* any meaningful treatment effect. |
| **95% CI** | (0.7144, 1.45) | Wide confidence interval, spanning both below and above 1—so we can't rule out either harm or benefit. |

**Concordance** The C-index, or concordance statistic is the probability
that a candomly hosen subject who died earlier had a higher predicted
risk. At 0.525 it is random chance, suggested poor predictive power.
:::

**Interpretation of the Cox Model Results**

The univariable Cox proportional hazards model found no statistically significant difference in survival between the control and intervention groups (HR = 1.02, 95% CI: 0.71–1.45, p = 0.92). The hazard ratio near 1 suggests little to no effect of the treatment on mortality risk. The wide confidence interval spans both potential benefit and harm, highlighting uncertainty in the estimate. The concordance statistic (0.525) indicates the model’s discriminative ability is only marginally better than chance. Overall, this analysis offers no evidence that the intervention improves survival in this sample.

```{r}
# Use ggadjustedcurves for plotting adjusted survival curves directly from the Cox model
ggadjustedcurves(
  cox_model,                     # The fitted Cox model
  data = veteran,                # The original data used to fit the model
  variable = "trt",              # The variable to plot adjusted curves for
  method = "average",            # Method for adjusting (average effect of other covariates)
  conf.int = TRUE,
  legend.labs = c("Control", "Intervention"),
  legend.title = "Treatment Group",
  risk.table = TRUE,             # Include the risk table
  risk.table.title = "Number at Risk",
  xlab = "Time (days)",
  ylab = "Adjusted Survival Probability",
  title = "Cox Model Adjusted Survival Curves",
  palette = c("#A569BD", "#45B39D"),
  ggtheme = theme_minimal(base_size = 13) +
    theme(
      legend.position = "top",
      plot.title = element_text(hjust = 0.5, face = "bold"),
      panel.grid.major = element_line(color = "grey80", linewidth = 0.4),
      panel.grid.minor = element_line(color = "grey90", linewidth = 0.2)
    ),
  tables.theme = theme(
    axis.text.y = element_text(size = 10, margin = margin(r = 12), lineheight = 2)
  )
)
```

::: {.callout-tip title="Why use `ggadjustedcurves()` for this plot" collapse="true"}
For plotting survival curves adjusted for covariates from a Cox
Proportional Hazards model, the ggadjustedcurves() function from the
survminer package offers one of the most direct and statistically
appropriate approaches—particularly as model complexity increases.

-   Statistical validity: ggadjustedcurves() uses the fitted cox_model
    to compute adjusted survival probabilities for the variable of
    interest (in this case, trt). When method = "average" is specified
    (the default when covariates are present), it returns marginal
    estimates—appropriate for interpreting average treatment effects
    across the population. In our simpler model (\~ trt), it effectively
    plots the model-predicted survival curve for each treatment group.

-   Robustness: This function is purpose-built for adjusted curve
    plotting and avoids many of the pitfalls seen with manually passing
    newdata to survfit() and plotting with ggsurvplot(), such as
    misaligned strata or legend issues. It correctly associates the
    model, the data, and the plot elements, including the risk table.

-   Simplicity: It eliminates the need for manually constructing a
    new_trt data frame or calling survfit() separately, resulting in
    cleaner, more reproducible code—particularly useful in collaborative
    or long-term projects.
:::

**Interpretation of the Adjusted Survival Curves**

The Cox-adjusted survival curves for the Control and Intervention groups are nearly identical, indicating no meaningful difference in predicted survival probabilities between the treatment arms. This visual overlap aligns with the Cox model's hazard ratio, which was close to 1 and not statistically significant. Together, the model output and adjusted curves reinforce the conclusion that the intervention had no observable effect on survival in this cohort.

::: {.callout-note title="Why do the lines overlap?" collapse="true"}
The survival curves for the Control and Intervention groups overlap
almost entirely in this plot. This is because the Cox proportional
hazards model found no significant difference between the two groups in
terms of survival (hazard ratio ≈ 1, *p* = 0.922). As a result, the
model-predicted curves are nearly identical. The faint visibility of the
Control group line is due to it being directly behind the Intervention
group line.
:::

**Model Diagnostics**

Before interpreting the results of a Cox proportional hazards model, it’s important to ensure that the model’s assumptions are met. The two key assumptions are:

- The proportional hazards (PH) assumption: The effect of each covariate on the hazard is constant over time.

- Linearity of continuous covariates: Continuous covariates are assumed to have a linear relationship with the log hazard.

Violations of these assumptions can undermine the model’s predictive utility and lead to misleading conclusions.

In this section, we use the following diagnostic tools to assess model adequacy:

- Schoenfeld residuals to test the PH assumption.

- Martingale residuals to check for non-linearity and outliers.

- Functional form checks, particularly for continuous covariates like age.

- Where needed, we apply transformations (e.g., natural splines) to better capture nonlinear effects.

These checks help ensure that the model remains a sound and interpretable framework for estimating covariate effects on survival.

_Checking the Proportional Hazards Assumption_

A central assumption of the Cox proportional hazards model is that hazard ratios remain constant over time — meaning the relative risk between any two individuals does not change as time progresses. This is known as the proportional hazards (PH) assumption, and verifying it is crucial before interpreting model results.

**Schoenfeld Residuals**

To assess the PH assumption, we use Schoenfeld residuals, which evaluate whether covariate effects vary over time. If residuals show a systematic pattern against time, it may indicate a violation of the assumption.

We check this in two ways:

- Statistical tests: The cox.zph() function provides both a global test and tests for each covariate. A p-value < 0.05 suggests evidence against the PH assumption.

- Visual inspection: Plots of scaled Schoenfeld residuals include a smoothed line and 95% confidence bands. Systematic trends or large deviations from zero may indicate time-varying effects.

If either test indicates a violation, we may need to adjust the model using time-varying covariates, stratification, or alternative approaches.

```{r}
# Test the PH assumption
cox.zph(cox_model)
```

::: {.callout-note title="Understanding Chi-Squared and p-values in PH Assumption Testing" collapse="true"}
When checking the **proportional hazards (PH) assumption** in a Cox
model using `cox.zph()`, each covariate — and the model as a whole — is
tested for **time-dependent effects**.

-   **Chi-squared statistic**: Measures how much the residuals deviate
    from what would be expected if the covariate had a constant effect
    over time (i.e., proportional hazards). A larger value suggests a
    stronger departure from proportionality.

-   **Degrees of freedom (df)**: Typically 1 for individual covariates,
    and equal to the number of covariates in the global test.

-   **p-value**: Quantifies the probability that the observed
    chi-squared value (or more extreme) could occur **by chance** if the
    PH assumption holds.

**How to interpret**: - `p > 0.05`: No significant evidence of
time-varying effects — PH assumption likely holds. - `p < 0.05`:
Suggests the effect of the covariate may change over time — a potential
violation of the PH assumption.

Use both **statistical tests** and **residual plots** for a more robust
assessment. A borderline p-value (e.g., 0.06) should be interpreted with
caution and always viewed in conjunction with the corresponding plot.
:::

**Interpretation of PH Assumption Test (Schoenfeld residuals)**

- trt p = 0.06: Slightly above the conventional 0.05 threshold. This suggests no strong evidence that the effect of treatment changes over time.

- GLOBAL p = 0.06: The overall model does not show a statistically significant violation of the PH assumption.

Conclusion: We do not reject the PH assumption based on either the individual or global tests. The model is likely appropriate for interpretation without time-dependent adjustments.

**Schoenfeld Residuals Plot**

```{r}
plot(cox.zph(cox_model))
```

::: {.callout-note title="What the Visual Diagnostic Shows" collapse="true"}
*The y-axis:* Scaled Schoenfeld residuals for trt. If the proportional
hazards assumption holds, these residuals should fluctuate randomly
around zero.

*The x-axis:* Time (or transformed time).

*The line:* A smoothed spline fit (typically a loess curve) showing the
trend of residuals over time.

*The confidence band:* Helps you assess whether the trend significantly
deviates from horizontal (i.e., no time-dependent effect).

*A flat line (slope ≈ 0):* Suggests the hazard ratio for trt does not
change over time — good!

*A wavy or sloped line:* Suggests potential violation of the
proportional hazards assumption — this would back up a significant
p-value.

*x-axis (Time):* Time since baseline (likely in days, from your
dataset).

*y-axis (Beta(t) for trt):* Time-varying estimate of the log hazard
ratio for the treatment group (trt), based on Schoenfeld residuals.

*The solid black line:* A smoothed curve showing how the estimated
coefficient for trt might change over time.

*Dashed lines:* Approximate 95% confidence bands around that smoothed
estimate.

*Open circles:* Residuals for each individual event — they don't need to
follow a pattern, but their distribution can hint at model fit.

If the solid line is roughly flat and within the dashed lines, the
effect of treatment does not change with time, and the PH assumption is
reasonably upheld.

If the solid line drifts noticeably over time, or wanders outside the
dashed lines, that's evidence that the treatment effect may change — a
violation of the PH assumption.

In our case, the line is gently curved, but still well within the dashed
boundaries. There's no strong time trend — the log hazard ratio is
hovering around zero and doesn't appear to increase or decrease
dramatically.
:::

**Schoenfeld Residuals Interpretation**

The global test for proportional hazards yielded a p-value of 0.06, slightly above the conventional 0.05 threshold, suggesting no strong evidence of a violation. The smoothed plot of Schoenfeld residuals for the treatment variable shows mild curvature but remains well within the 95% confidence bands, with no clear time-dependent trend. Together, these results support the proportional hazards assumption for treatment and the suitability of the Cox model in this context.

**Martingale Residuals (for Functional Form of Covariates)**

Martingale residuals are used to evaluate whether continuous covariates have been correctly specified in the Cox model. By plotting these residuals against each covariate, we can detect departures from linearity on the log-hazard scale. Systematic patterns, such as curvature or clustering, may indicate the need for transformations (e.g., splines or categorization) to better capture the relationship.

::: {.callout-note title="Understanding Martingale Residuals" collapse="true"}
**Martingale residuals** are used to assess whether the **functional
form** of a continuous covariate is correctly specified in a Cox
proportional hazards model.

**What are they?**\
Martingale residuals represent the difference between the **observed
number of events** and the **expected number of events** under the
fitted model for each subject.\
They range from **–∞ to +1** and are **asymmetrically distributed**.

**Why do we use them?**\
To check the **linearity assumption**: the Cox model assumes that the
covariate has a **linear effect on the log hazard**. If this is not true
(e.g. the effect increases or decreases non-linearly), the model may
misestimate risk.

**What should you look for in the plot?**\
You typically plot Martingale residuals **against a continuous
covariate** (e.g. age). Then:

-   A **random scatter** around the horizontal line at 0 suggests
    linearity is reasonable.\
-   A **systematic curve** (e.g., U-shaped, S-shaped) suggests
    **non-linearity** and the need for transformations (e.g., log(age),
    age², splines).

**Practical Tip:**\
Martingale residuals are **not symmetric**, so smooth trends can be hard
to detect visually. It's a good idea to pair this check with
**`ggcoxfunctional()`** or use **splines** to model nonlinear effects
more flexibly.
:::

```{r}
# Generate Martingale residuals
martingale_resid <- residuals(cox_model, type = "martingale")

# Add residuals to your data
veteran$martingale <- martingale_resid

# Plot residuals against a continuous covariate (e.g., time or another variable)
# Here we plot against the linear predictor for simplicity
veteran$linear_pred <- predict(cox_model, type = "lp")

ggplot(veteran, aes(x = linear_pred, y = martingale, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Linear Predictor",
    x = "Linear Predictor (Log Hazard)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))


```

::: {.callout-note title="What do the Martingale Residuals Show" collapse="true"}
Martingale residuals help us check whether the variables in our Cox
model are specified correctly — particularly useful for continuous
covariates. However, when the variable in question is binary (like our
treatment group: Control vs Intervention), the Martingale residual plot
will display two vertical stacks of points — one for each group.

In the plot above:

-   The **x-axis** represents the linear predictor (log hazard). Because
    the treatment is binary, there are only two unique predictor values:
    one for the control group and one for the intervention group.
-   The **y-axis** shows the Martingale residuals for each individual.
-   The **dashed horizontal line at 0** acts as a visual guide to see if
    residuals are randomly scattered above and below it.

We want to see: - No clear patterns or trends in the residuals, - A
reasonably symmetric spread around zero in each group, - No outliers
clustering far from the bulk of points.

In this case, both groups show a spread of residuals centered roughly
around zero, with no curvature or skew that would suggest model
misspecification. That's a good sign — the model's structure seems to
handle the treatment variable well, and there's no obvious reason to
suspect we need to transform it or add interaction terms.
:::

**Interpretation of Martingale Residuals Plot**

The Martingale residual plot helps assess whether the model’s assumptions—particularly the functional form of covariates—are valid.

In this case, residuals were plotted against the linear predictor for a binary treatment variable. As expected, two vertical bands appear, corresponding to the Control and Intervention groups. The residuals are roughly symmetric around zero within each group, with no obvious curvature or systematic pattern. This supports the adequacy of the treatment variable’s specification in the model and suggests that no transformation is needed for this binary covariate.

::: {.callout-tip title="Why we haven't checked the functional form of covariates" collapse="true"}
The Cox model assumes a linear relationship between continuous
covariates and the log hazard. In our current model, only the treatment
group (`trt`) is included, which is a categorical variable.

If we were to build a more complex model including continuous predictors
like age, Karnofsky score (`karno`), or time since diagnosis
(`diagtime`), we would need to assess whether their relationship with
the log hazard is truly linear. This is typically done by plotting
Martingale residuals or using `ggcoxfunctional()`.

For now, since our model does not include continuous covariates, this
step is not necessary — but it's an important part of model diagnostics
when expanding our analysis.
:::

**Additional Covariates**

To improve the model and better account for variability in survival outcomes, we now include additional patient characteristics: age (a continuous variable) and prior treatment history (prior, a binary variable). Adding these covariates allows us to control for baseline differences between patients, assess their independent associations with survival, and explore whether they influence the effect of treatment.

::: {.callout-note title="Why Add Covariates to a Cox Model?" collapse="true"}
Including additional covariates in a Cox proportional hazards model serves several purposes:

- Control for confounding: Variables such as age or prior treatment may influence both the outcome and the likelihood of receiving a particular treatment. Adjusting for them helps isolate the effect of the treatment itself.

- Improve model accuracy: Predictive covariates enhance the model’s explanatory power and reduce residual variation.

- Estimate independent effects: Including multiple variables allows us to assess their individual contributions to survival outcomes, independent of each other.

- Explore effect modification: With covariates in the model, we can investigate interactions — for example, whether the effect of treatment varies by age or prior therapy.

When selecting covariates, decisions should be guided by clinical relevance and theoretical grounding, not just statistical significance.
:::

## Multivariable Models

To build on the univariable analysis of treatment effects, we next fit a
multivariable Cox proportional hazards model that incorporates
additional patient-level covariates. This allows us to estimate the
effect of treatment on survival **while adjusting for other potential
prognostic factors**, such as age and prior treatment history. By doing
so, we account for potential confounding and improve the model's
explanatory power.

::: {.callout-tip title="What is a multivariable Cox model and why are we using it?" collapse="true"}
In a univariable Cox model, we examine the effect of **one predictor at
a time** on the hazard of an event (e.g., death). While this can be
useful for preliminary exploration, it doesn't account for the
**influence of other variables** that might confound or modify the
relationship.

A **multivariable Cox proportional hazards model** includes **multiple
covariates** simultaneously. This allows us to: - Estimate the
**adjusted effect** of each variable while controlling for the others, -
Improve model accuracy and interpretability, - Detect whether the
apparent effect of one variable (e.g., treatment) persists after
adjusting for others (e.g., age, comorbidities), - Explore potential
**interactions** or **effect modification**.

In this section, we'll first add a **continuous covariate** (`age`) and
then a **binary variable** (`prior treatment`) to the model. We'll also
assess whether the functional form of continuous variables is
appropriate and ensure that model assumptions still hold.
:::

::: {.callout-tip title="Why not use logistic regression?" collapse="true"}
Logistic regression ignores time-to-event and censoring. When working
with survival data, Cox models are preferred as they preserve the time
structure and allow for censored cases.
:::

### Continuous Variable: Age

Age is a clinically relevant factor that may influence survival outcomes
in lung cancer. In this section, we assess whether age is independently
associated with survival time and verify that its relationship with the log hazard is appropriately linear.

::: {.callout-tip title="Modeling Continuous Covariates" collapse="true"}
Unlike categorical variables, continuous predictors like age require us
to check their **functional form** in the model. The Cox model assumes a
**linear relationship** between the covariate and the log hazard —
meaning each 1-year increase in age is assumed to have the same
proportional effect on the hazard.

If this assumption doesn't hold (e.g., if risk increases faster at older
ages), we may need to use transformations or splines to accurately model
the relationship.

We'll start by fitting a Cox model including age and then assess the
functional form using Martingale residuals.
:::

**Cox Model with Age**

```{r}
cox_age <- coxph(Surv(time, status) ~ trt + age, data = veteran)
summary(cox_age)
```

**Interpretation of the Cox Model with Age**

A Cox proportional hazards model was fitted to examine the relationship
between survival time and two predictors: treatment group (trt) and
patient age (age). The model included 137 patients, with 128 events
(deaths).

The hazard ratio for the intervention group was 0.996 (95% CI: 0.697 –
1.425; p = 0.984), suggesting virtually no difference in hazard compared
to the control group after adjusting for age. This effect was not
statistically significant.

For age, the hazard ratio was 1.008 (95% CI: 0.989 – 1.027; p = 0.436),
indicating that each additional year of age was associated with a 0.8%
increase in the hazard of death. However, this association was also not
statistically significant and the confidence interval included 1,
suggesting uncertainty about the true effect.

Overall, the model's ability to discriminate between patients with
longer vs shorter survival times was weak, with a concordance index of
0.514 (where 0.5 = random chance). The global model statistics —
including the likelihood ratio test, Wald test, and score test — were
all non-significant (p = 0.7), indicating that neither variable added
substantial predictive value to the model.

-   There is no evidence of a treatment effect after adjusting for age.

-   Age may have a small effect, but in this sample, it is not
    statistically or clinically meaningful.

-   The overall model performs poorly in predicting survival in this
    dataset.

::: {.callout-note title="Interpreting the Cox Model Output" collapse="true"}
Understanding the `coxph()` output is essential for interpreting
survival models. Here's a breakdown of the most common components you'll
encounter in both univariable and multivariable Cox regressions:

**🔹 *coef***\
This is the **log hazard ratio** — the raw estimate from the model. It
represents the change in the **log hazard** of the event (e.g., death)
associated with a **one-unit increase** in the covariate (for continuous
variables) or **relative to a reference group** (for categorical
variables).

-   Positive `coef` → increased hazard (i.e., higher risk)\
-   Negative `coef` → decreased hazard (i.e., protective)\
-   A `coef` close to 0 suggests little to no effect

🧠 *Note:* Log hazards are difficult to interpret directly, so we
usually exponentiate them.

**🔹 *exp(coef)***\
This is the **hazard ratio (HR)** — the exponentiated `coef`, which
expresses effect size more intuitively.

-   `exp(coef) > 1` → Increased hazard\
-   `exp(coef) < 1` → Decreased hazard\
-   `exp(coef) = 1` → No effect

🔎 Example: `exp(coef) = 1.20` means a 20% higher hazard.\
`exp(coef) = 0.80` means a 20% lower hazard.

**🔹 *se(coef)***\
The **standard error** of the coefficient — reflects how much
uncertainty surrounds the estimate.

-   Smaller values = more precise estimates\
-   Larger values = more uncertainty (due to small samples or noisy
    data)

**🔹 *z* and *Pr(\>\|z\|)***\
- `z` is the **Z-statistic**, calculated as `coef / se(coef)`\
- `Pr(>|z|)` is the **p-value** from the Wald test for each coefficient

This tests the **null hypothesis** that the covariate has **no effect**
(`coef = 0`).

-   `p < 0.05` → Statistically significant\
-   `p > 0.05` → Not statistically significant

**📊 *Model-Level Statistics***

**🔸 *Concordance***\
- Measures the model's ability to correctly rank survival times (similar
to AUC for classification)\
- Ranges from `0.5` (random) to `1.0` (perfect)\
- In clinical data, values of `0.6–0.75` are typical

**🔸 *Likelihood Ratio Test***\
- Compares the full model to a null (intercept-only) model\
- Tests overall model fit\
- `p < 0.05` = your model significantly improves fit

**🔸 *Wald Test***\
- Aggregates the individual Wald tests across all covariates\
- Tests whether any coefficients differ significantly from 0

**🔸 *Score (Logrank) Test***\
- Rank-based test comparing observed vs expected survival\
- Can be slightly more sensitive in small samples\
- Often used as an overall model check

✅ *In practice:* Use all three global tests and the concordance to
evaluate how well the model explains variation in survival. If all are
non-significant, the model may lack predictive utility.
:::

**Schoenfeld Residuals for Age**

```{r}
# Test proportional hazards assumption for the model including age
cox_zph_age <- cox.zph(cox_age)
print(cox_zph_age)
```

Proportional hazards were tested using Schoenfeld residuals. The
assumption was met for **age** (`p = 0.195`) but marginally violated for
**treatment group** (`p = 0.055`), which is slightly below the
conventional threshold of 0.05. The **global test** for the model
(`p = 0.045`) was also just below 0.05, indicating a **possible overall
violation** of the proportional hazards assumption.

While these results do not strongly indicate a breakdown of the PH
assumption, the borderline p-values warrant caution. A visual inspection
of the residual plots (`plot(cox.zph())`) can help determine whether the
trend is substantial or random variation.

::: {.callout-note title="Interpreting Schoenfeld Residuals" collapse="true"}
Schoenfeld residuals are used to test the **Proportional Hazards (PH)
assumption** in Cox models — that the effect of each covariate remains
**constant over time**.

**How to interpret the output:**

-   Each row (e.g., `trt`, `age`) gives a **test of proportionality**
    for that specific variable.
-   The `GLOBAL` row tests **all covariates together**.

**What the numbers mean:**

-   `chisq` is the test statistic (larger values = more evidence of
    time-varying effects).
-   `p` is the p-value testing **null hypothesis: proportional hazards
    holds**.

**Rule of thumb:** - `p < 0.05` → **PH assumption violated** (bad) -
`p > 0.05` → **PH assumption holds** (good)

🧠 *Note:* Slight violations (e.g., `p = 0.055`) might not be a problem
in practice — but major ones (e.g., `p = 0.01`) should be addressed.
:::

```{r}
# Visual check
plot(cox_zph_age)
```

The Schoenfeld residual plot for age, derived from the multivariable Cox
model, reveals no strong evidence of violation of the proportional
hazards assumption. The smoothed coefficient curve remains relatively
flat across time, with the 95% confidence band consistently overlapping
zero. This suggests that the effect of age on the hazard function does
not meaningfully vary with time. This visual impression aligns with the
result from the formal statistical test (χ² = 1.68, df = 1, p = 0.195),
providing no significant indication that the proportional hazards
assumption has been violated for age in this model.

::: {.callout-note title="How to Interpret Schoenfeld Residual Plots" collapse="true"}
This plot checks the **Proportional Hazards (PH) assumption** visually
for a covariate (e.g., `age`).

**What we are seeing:**

-   **X-axis:** Time\
-   **Y-axis:** Estimated **effect (beta)** of the covariate at each
    time point\
-   **Dots:** Residuals at each event time\
-   **Solid line:** Smoothed trend of the coefficient over time\
-   **Dashed lines:** 95% confidence interval of that trend

**What we want to see:**

✅ A **flat line near zero**, with the confidence band **containing
zero** throughout time → suggests **PH assumption holds**.

⚠️ A **wavy trend** or curve **drifting far from zero** → suggests
**non-proportional hazards** (i.e., the effect of the covariate changes
over time).

**Interpreting the plot:**

If the smoothed line is approximately flat and the confidence interval
consistently includes zero — with no strong patterns or directional
trends — then the PH assumption is likely satisfied for that covariate.
:::

**Martingale Residuals for Age to Assess Functional Form**

To assess whether age is appropriately modeled as a linear predictor in
the Cox model, we examine Martingale residuals plotted against age to
detect non-linearity in its functional form.

```{r}
# Generate Martingale residuals
veteran$martingale_age <- residuals(cox_age, type = "martingale")

# Generate linear predictor
veteran$linear_pred_age <- predict(cox_age, type = "lp")

# Plot residuals against age
ggplot(veteran, aes(x = age, y = martingale_age, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Age",
    x = "Age (years)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

Martingale residuals plotted against age showed no clear non-linear
trend, with residuals generally scattered randomly across the range of
age values. This suggests that modeling age as a linear term in the Cox
model is reasonable. There is no strong evidence to support the need for
transformations or spline-based modeling of age in this dataset.

::: {.callout-note title="How to Interpret Martingale Residuals for Functional Form" collapse="true"}
This plot helps assess whether the **relationship between a continuous
variable** (like *age*) and the **log hazard** is **linear**, which is a
core assumption of the Cox model.

-   **X-axis:** The covariate
-   **Y-axis:** Martingale residuals (bounded between -∞ and 1)
-   **Points:** One per individual, representing how well their survival
    time is predicted
-   **Dashed line at 0:** Reference line to help spot patterns

**What to look for:** - A **random scatter** of points across the X-axis
with **no clear curve** → linearity assumption is likely satisfied ✅ -
A **systematic curve**, bend, or wave → suggests the effect of age might
be **non-linear** ⚠️\
In that case, consider modeling age using splines or transformations
(e.g., `log()` or `^2`).

**Important:** These residuals are noisier near their bounds and should
be interpreted cautiously, especially when sample size is limited.
:::

**Functional Form Check**

```{r}
# Fit a univariable Cox model to assess functional form of age
cox_age_only <- coxph(Surv(time, status) ~ age, data = veteran)

# Check the functional form visually using Martingale residuals
ggcoxfunctional(cox_age_only, data = veteran)
```

::: {.callout-note title="Assessing Functional Form with Martingale Residuals" collapse="true"}
**What this plot shows:**\
This diagnostic visualizes whether a **continuous covariate** (e.g.,
*age*) has a **linear relationship** with the **log hazard**, as assumed
by the Cox model.

**X-axis:**\
The values of the continuous variable being assessed (e.g., age)

**Y-axis:**\
The **Martingale residuals** from a **null Cox model** (i.e., a model
with no covariates)

**Dots:**\
Each dot represents one observation

**Smooth black line:**\
A **non-parametric smoother** that helps reveal trends in the residuals
— this is the key to checking the shape of the relationship

**How to interpret it:**

-   ✅ If the **black line is roughly straight and flat**, this suggests
    a **linear relationship** between the variable and the log hazard —
    which is **what the Cox model assumes**.
-   ⚠️ If the line shows a **curved or non-linear pattern**, this
    suggests the variable may not be appropriately modeled with a simple
    linear term.
    -   This can indicate the need to use **transformations** (e.g.,
        `log(age)`, `age^2`) or **splines**.

**Tip:**\
Small wobbles are not a major concern — we are primarily looking for
clear trends, not minor bumps. Major **U-shapes, inflections, or long
curves** are the most concerning signs.
:::

The functional form check using Martingale residuals and a
non-parametric smoother suggests that age may not have a strictly linear
relationship with the log hazard. The smooth curve appears U-shaped,
implying that risk may not increase evenly with age across the full
range. This indicates that the assumption of linearity may be violated,
and that alternative modeling strategies (e.g., polynomial terms or
splines) could improve model fit.

While the Martingale residual plot suggests that the relationship
between age and the log hazard may not be strictly linear, age was not
statistically significant in the model. As such, no transformation or
spline was applied. This could be revisited in future modeling if age
becomes more influential.

### Binary Variable: Prior Treatment

We now evaluate the variable `prior`, which indicates whether a patient
received prior therapy before enrolment in the study. This binary
variable may capture important differences in disease severity,
treatment history, or prognosis that could influence survival outcomes.

::: {.callout-tip title="Why test binary covariates like 'prior'?" collapse="true"}
Even though `prior` is binary, its inclusion allows us to control for
historical differences in patient treatment, which may affect survival
independently of the trial treatment (`trt`).

For binary variables, we don't need to assess the functional form, but
we **do need to check the proportional hazards assumption** (as with any
covariate in the Cox model). We'll interpret the hazard ratio as the
relative difference in risk between those who received prior therapy and
those who did not.
:::

**Fit Cox Model with Prior Treatment**

```{r}
# Fit the Cox model including prior treatment
cox_prior <- coxph(Surv(time, status) ~ trt + prior, data = veteran)
summary(cox_prior)
```

A Cox proportional hazards model was fitted to evaluate the association
between survival and two predictors: treatment group (trt) and prior
therapy (prior). The model was based on 137 patients, with 128 observed
events.

The hazard ratio for the intervention group was 1.03 (95% CI: 0.72 –
1.46, p = 0.885), indicating no evidence of a treatment effect after
adjusting for prior therapy. Similarly, prior therapy was not associated
with survival, with a hazard ratio of 0.99 (95% CI: 0.95 – 1.03, p =
0.471).

Model discrimination was weak, with a concordance of 0.517, suggesting
the model has limited ability to rank survival times accurately. Global
model fit tests (likelihood ratio, Wald, and score) were all
non-significant (p = 0.8), indicating that the included variables do not
significantly improve model performance.

Neither treatment assignment nor prior therapy appear to be meaningful
predictors of survival in this dataset. There is no evidence of an
additive effect between these covariates.

**Schoenfeld Residuals for Prior Treatment**

```{r}
# Test proportional hazards assumption for the model including prior treatment
cox_zph_prior <- cox.zph(cox_prior)
print(cox_zph_prior)
```

Proportional hazards were assessed using Schoenfeld residuals for the
model including treatment group and prior therapy. The global test was
statistically significant (p = 0.048), suggesting potential violation of
the proportional hazards assumption overall.

Individually, both treatment (p = 0.066) and prior therapy (p = 0.081)
had p-values slightly above the conventional 0.05 threshold, indicating
borderline evidence of non-proportionality. These results do not
strongly confirm a violation, but they do warrant cautious
interpretation.

Given the marginal nature of these findings, visual inspection of the
Schoenfeld residual plots is recommended to assess whether any
systematic time-varying effects are present.

```{r}
# Visual check
plot(cox_zph_prior)
```

The Schoenfeld residual plot for prior treatment shows a smoothed line
that remains approximately flat over time, with the 95% confidence band
consistently including zero. This suggests that the effect of prior
treatment on the hazard is stable across the follow-up period, and
supports the assumption of proportional hazards for this variable.

In contrast, the plot for treatment group (trt) exhibits a noticeable
wavy pattern in the smoothed line, particularly in the early to
mid-study period. While the confidence bands still broadly include zero,
the shape of the curve raises some concern about potential time-varying
effects. Combined with a borderline p-value from the cox.zph() test (p =
0.060), this suggests the proportional hazards assumption for trt is not
clearly violated, but may warrant caution and further
consideration—especially in more complex models.

**Martingale Residuals for Prior Treatment**

```{r}
# Generate Martingale residuals
veteran$martingale_prior <- residuals(cox_prior, type = "martingale")
# Generate linear predictor
veteran$linear_pred_prior <- predict(cox_prior, type = "lp")
# Plot residuals against prior treatment
ggplot(veteran, aes(x = prior, y = martingale_prior, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Prior Treatment",
    x = "Prior Treatment (0 = No, 1 = Yes)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) # Adding jitter to avoid overlap

```

The Martingale residual plot for the prior variable shows two distinct
clusters of points at 0 and 1, corresponding to patients with and
without prior treatment, respectively. The residuals appear to be
roughly symmetric around zero in both groups, with no strong indication
of skew or heteroscedasticity. This supports the appropriateness of
modeling prior as a binary variable in the Cox regression model. No
transformation or alternative coding appears necessary.


### Multivariable Cox Model with Age and Prior Treatment

In this section, we fit a multivariable Cox proportional hazards model
that includes both age and prior treatment as covariates. This allows us
to assess the independent effects of these variables on survival while
controlling for treatment group.

```{r}
# Fit the multivariable Cox model including age and prior treatment
cox_multivariable <- coxph(Surv(time, status) ~ trt + age + prior, data = veteran)
summary(cox_multivariable)
```

::: {.callout-note title="Interpreting the Cox Model Output" collapse="true"}
Understanding the `coxph()` output is essential for interpreting
survival models. Here's a breakdown of the most common components you'll
encounter in both univariable and multivariable Cox regressions:

**🔹 *coef***\
This is the **log hazard ratio** — the raw estimate from the model. It
represents the change in the **log hazard** of the event (e.g., death)
associated with a **one-unit increase** in the covariate (for continuous
variables) or **relative to a reference group** (for categorical
variables).

-   Positive `coef` → increased hazard (i.e., higher risk)\
-   Negative `coef` → decreased hazard (i.e., protective)\
-   A `coef` close to 0 suggests little to no effect

🧠 *Note:* Log hazards are difficult to interpret directly, so we
usually exponentiate them.

**🔹 *exp(coef)***\
This is the **hazard ratio (HR)** — the exponentiated `coef`, which
expresses effect size more intuitively.

-   `exp(coef) > 1` → Increased hazard\
-   `exp(coef) < 1` → Decreased hazard\
-   `exp(coef) = 1` → No effect

🔎 Example: `exp(coef) = 1.20` means a 20% higher hazard.\
`exp(coef) = 0.80` means a 20% lower hazard.

**🔹 *se(coef)***\
The **standard error** of the coefficient — reflects how much
uncertainty surrounds the estimate.

-   Smaller values = more precise estimates\
-   Larger values = more uncertainty (due to small samples or noisy
    data)

**🔹 *z* and *Pr(\>\|z\|)***\
- `z` is the **Z-statistic**, calculated as `coef / se(coef)`\
- `Pr(>|z|)` is the **p-value** from the Wald test for each coefficient

This tests the **null hypothesis** that the covariate has **no effect**
(`coef = 0`).

-   `p < 0.05` → Statistically significant\
-   `p > 0.05` → Not statistically significant

**📊 *Model-Level Statistics***

**🔸 *Concordance***\
- Measures the model's ability to correctly rank survival times (similar
to AUC for classification)\
- Ranges from `0.5` (random) to `1.0` (perfect)\
- In clinical data, values of `0.6–0.75` are typical

**🔸 *Likelihood Ratio Test***\
- Compares the full model to a null (intercept-only) model\
- Tests overall model fit\
- `p < 0.05` = your model significantly improves fit

**🔸 *Wald Test***\
- Aggregates the individual Wald tests across all covariates\
- Tests whether any coefficients differ significantly from 0

**🔸 *Score (Logrank) Test***\
- Rank-based test comparing observed vs expected survival\
- Can be slightly more sensitive in small samples\
- Often used as an overall model check

✅ *In practice:* Use all three global tests and the concordance to
evaluate how well the model explains variation in survival. If all are
non-significant, the model may lack predictive utility.
:::

A multivariable Cox proportional hazards model was fitted to examine the
association between survival and three covariates: treatment group, age,
and prior therapy. After adjusting for all covariates:

-   The hazard ratio for the intervention group relative to control was
    1.00 (HR = 1.00; 95% CI: 0.70–1.44; p = 0.984), indicating no
    evidence of a treatment effect.

-   Age had a hazard ratio of 1.01 per year (p = 0.462), suggesting no
    significant association with survival in this model.

-   Prior therapy also showed no effect on hazard (HR = 0.99; p =
    0.500).

Model discrimination was minimal (concordance = 0.507), and all global
model tests were non-significant (p = 0.8), suggesting the model does
not meaningfully predict survival in this dataset.

**Schoenfeld Residuals for Multivariable Model**

We now assess whether the **proportional hazards assumption** holds for
each covariate in the multivariable Cox model. This step ensures that
the effect of each variable — treatment group, age, and prior therapy —
is consistent over time.

```{r}
# Test proportional hazards assumption for the multivariable model
cox_zph_multivariable <- cox.zph(cox_multivariable)
print(cox_zph_multivariable)
```

The proportional hazards assumption was assessed for the multivariable
Cox model using Schoenfeld residuals. The global test was statistically
significant (p = 0.032), suggesting that at least one covariate may have
a time-varying effect on the hazard.

At the individual variable level:

-   trt showed borderline evidence of non-proportionality (p = 0.060)

-   prior also approached significance (p = 0.074)

-   age did not show evidence of time-varying effects (p = 0.184)

While none of the individual tests crossed the 0.05 threshold, the
cumulative result (GLOBAL p \< 0.05) suggests some caution is warranted.
Visual inspection of the Schoenfeld residual plots is recommended to
evaluate whether these borderline findings correspond to meaningful
deviations over time.

```{r}
# Visual check
plot(cox_zph_multivariable)
```

Visual inspection of the Schoenfeld residual plots for treatment group
(trt), age, and prior treatment (prior) showed no substantial evidence
of non-proportionality. The smoothed lines for each covariate remained
relatively flat and were mostly contained within their respective 95%
confidence bands across the follow-up period.

-   For treatment group, the smoothed beta function showed some mild
    undulation, but it remained within the confidence bounds, indicating
    no major deviation from the proportional hazards assumption.

-   For age, the trend line was nearly horizontal with minimal
    fluctuation, strongly supporting time-invariant effects.

-   For prior treatment, the estimated beta also remained relatively
    stable over time, with the confidence interval consistently covering
    zero.

Together, these plots support the conclusion that the proportional
hazards assumption is reasonably satisfied for each covariate in the
multivariable model, aligning with the statistical test results where
none of the individual covariate tests were significant at the 0.05
level.

**Martingale Residuals for Multivariable Model**

We now examine the Martingale residuals from the multivariable Cox model
to evaluate the overall model fit and identify any patterns or potential
outliers in the covariate effects.

```{r}
# Generate Martingale residuals
veteran$martingale_multivariable <- residuals(cox_multivariable, type = "martingale")
# Generate linear predictor
veteran$linear_pred_multivariable <- predict(cox_multivariable, type = "lp")
# Plot residuals against linear predictor
ggplot(veteran, aes(x = linear_pred_multivariable, y = martingale_multivariable, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Linear Predictor\n(Multivariable Model)",
    x = "Linear Predictor (Log Hazard)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The Martingale residuals plotted against the linear predictor for the
multivariable model appear randomly scattered around the horizontal zero
line, with no clear pattern, curvature, or systematic deviation. This
suggests that the overall functional form of the covariates (treatment,
age, and prior treatment) is reasonably specified. There is no strong
evidence of model misspecification or non-linearity in the combined
effects. However, the narrow spread of the linear predictor values
(mostly between -0.2 and 0.2) reflects the modest effect sizes and
limited variation in the log hazard estimates, which aligns with the
model's lack of significant predictors.

::: {.callout-tip title="What This Plot Tells Us" collapse="true"}
Martingale residuals help assess how well the Cox model fits the data.

A good model will show residuals scattered symmetrically around zero
with no systematic pattern.

Since we are plotting residuals against the linear predictor (the
model's estimate of log hazard for each individual), this acts as a
catch-all for checking overall model specification.

In this case:

-   There's no strong non-linear trend.
-   No excessive clustering or heteroskedasticity.
-   Values are distributed fairly evenly above and below the zero line.

Conclusion: The model appears reasonably well-specified, and no major
violations are detected.
:::

**Functional Form Check for Continuous Covariates**

We now recheck the functional form of the continuous variable age within
the multivariable Cox model to ensure its effect on the log hazard
remains appropriately modeled as linear.

```{r}
# Refit a univariable model with age (again, for functional form check)
cox_age_for_form_check <- coxph(Surv(time, status) ~ age, data = veteran)

# Plot functional form using ggcoxfunctional
ggcoxfunctional(
  fit = cox_age_for_form_check,
  data = veteran,
  point.col = "#E74C3C",
  ggtheme = theme_minimal(base_size = 13),
  title = "Functional Form Check: Age"
)
```

The Martingale residual plot for age reveals a clear U-shaped curve,
suggesting that the assumption of linearity in the log hazard may not
hold. This implies that age may have a non-linear effect on survival,
and modeling it as a simple linear term may lead to misestimation of its
effect. A transformation (e.g., quadratic or restricted cubic splines)
may be warranted in further analysis.

::: {.callout-tip title="Why Check Functional Form Again in a Multivariable Model?" collapse="true"}
While functional form checks (e.g., using Martingale residuals) are
**only necessary for continuous covariates**, it's important to
**reassess them in the context of the full model** — even if they passed
in a univariable setting.

-   In a **univariable Cox model**, we check whether the continuous
    variable (e.g., age) has a linear relationship with the log hazard.
-   However, once we add other covariates (like treatment or prior
    therapy), the relationship can **change due to confounding or
    interactions**.
-   A variable might look linear on its own, but **show non-linearity
    when adjusted** for other predictors.

So we recheck **only the continuous variables** (e.g., age) in the
multivariable model to ensure the assumption still holds.
:::

### Multivariable Cox Model with Spline-Transformed Age

We refit the Cox model using a restricted cubic spline to model the
non-linear relationship between age and the log hazard. This allows for
greater flexibility in capturing age-related risk while still accounting
for treatment and prior therapy.

::: {.callout-note title="Understanding Spline and Non-Linear Transformations in Survival Models" collapse="true"}
**Why model transformations at all?**\
The Cox Proportional Hazards model assumes that each covariate has a
**linear relationship** with the **log hazard**. For **continuous
variables** like age, this means that each one-unit increase is expected
to have the same multiplicative effect on the hazard — *regardless of
whether someone is 40 or 80 years old*. This is often unrealistic.

To fix this, we apply **non-linear transformations** to more accurately
capture the shape of the relationship between a variable (like age) and
the outcome (e.g., death).

**What are spline transformations?**\
Splines are a **flexible way to model non-linear relationships** without
explicitly specifying the form of the curve. They break a variable into
*segments* using predefined cut points called **knots**, and fit
separate curves (usually polynomials) between them, ensuring smoothness
at the joins.

They allow the model to "bend" as needed — making them perfect for
capturing complex patterns in the data.

**Common types of transformations:**

**• *Restricted Cubic Splines (Natural Splines)*:**\
This is the most common spline in Cox models. It creates piecewise cubic
polynomials between knots, but constrains the ends to be linear — which
prevents overfitting at the tails.

-   ✅ Best for *flexible yet stable* curve fitting\
-   📌 Recommended when you suspect non-linearity but don't want to
    manually test transformations

**• *Quadratic or Polynomial Terms (e.g., age + I(age\^2))*:**\
Adds a squared (or cubic) term to capture curvature.

-   ✅ Simple to interpret
-   ⚠️ Risky at the extremes — can overfit or behave erratically

**• *Log Transformations (e.g., log(age))*:**\
Used when the effect *decelerates* — i.e., rapid effect early on that
flattens with increasing value.

-   ✅ Great when data is skewed
-   ⚠️ Doesn't work if the variable contains 0s or negatives

**• *Categorization (e.g., age groups)*:**\
Turns a continuous variable into bins (e.g., under 50, 50–65, over 65).

-   ✅ Easy to explain
-   ⚠️ Loses statistical power and may mislead if cutoffs are arbitrary

**When should you use each?**

| Transformation     | Use when...                                          |
|-----------------------|-------------------------------------------------|
| **Splines (ns())** | You want a smooth, data-driven relationship          |
| **Quadratic term** | You suspect a simple "U-shaped" or curvilinear trend |
| **Log transform**  | The variable increases rapidly then levels off       |
| **Categories**     | Clinical guidelines or reporting require it          |

**How many knots do I use for splines?**\
A good rule of thumb (Harrell, 2015):

-   3 knots: small datasets (n \< 100)
-   4–5 knots: medium datasets (n \~ 100–300)
-   5–7 knots: large datasets

Knots are often placed at **percentiles** of the variable — e.g., 5th,
35th, 65th, 95th percentile.

Final tip: Always re-check Schoenfeld residuals and Martingale residuals
after applying any transformation.
:::

**Model Output**

```{r}
# Fit a Cox model with restricted cubic splines for age
cox_spline <- coxph(Surv(time, status) ~ ns(age, df = 3) + trt + prior, data = veteran)
summary(cox_spline)
```

A multivariable Cox proportional hazards model incorporating a
restricted cubic spline transformation of age (df = 3) was fitted to
explore potential non-linear effects of age on survival. None of the
individual spline terms were statistically significant at the 0.05 level
(p = 0.906, 0.143, and 0.157), suggesting that while the spline captures
flexibility, the overall contribution of age remains statistically weak.

The treatment group (p = 0.626) and prior therapy (p = 0.640) also
remained non-significant. Model-level tests showed that the likelihood
ratio test (p = 0.1), Wald test (p = 0.1), and score test (p = 0.1) were
all non-significant, indicating that the model, as specified, may not
provide strong explanatory power.

However, the model's concordance improved slightly to 0.559 compared to
the linear age model, suggesting a modest increase in predictive
ability. While not conclusive, the use of splines helps satisfy the
functional form assumption and could be valuable if future models
identify stronger non-linear trends or interactions with age.

**Schoenfeld Residuals for Spline Model**

```{r}
# Test proportional hazards assumption for the spline model
cox.zph(cox_spline)
```

A Cox proportional hazards model incorporating a restricted cubic spline
transformation of age (df = 3) demonstrated acceptable fit to the
proportional hazards assumption across all covariates. The global
Schoenfeld residuals test was non-significant (p = 0.116), suggesting
that the model adequately satisfies the proportional hazards assumption.
Individually, spline-transformed age (p = 0.211), treatment group (p =
0.114), and prior treatment (p = 0.079) all showed no significant
time-dependent effects.

```{r}
# Visual check of Schoenfeld residuals for spline model
plot(cox.zph(cox_spline))
```

The Schoenfeld residuals were examined for the final multivariable Cox
model, which included a spline-transformed age term. Across all
covariates—age (using splines), treatment group (trt), and prior
treatment (prior)—the smoothed residual lines remained relatively flat
and within the 95% confidence intervals, suggesting no strong evidence
of time-varying effects.

-   Age (spline-transformed): Despite some visual variability, the three
    spline basis functions for age each produce smoothed lines that
    remain stable across time. Importantly, the confidence intervals
    broadly contain zero, indicating that the assumption of
    proportionality is not clearly violated despite the model’s
    non-linear specification.

-   Treatment (trt): The smoothed residual curve shows a mild wave-like
    pattern, but this fluctuation stays well within the confidence bands
    and does not indicate a strong trend over time. While a borderline
    global p-value was observed in earlier models, the residuals here
    provide no compelling evidence against the proportional hazards
    assumption for treatment.

-   Prior Treatment (prior): The smoothed residual line is relatively
    flat, and although the lower bound of the 95% confidence interval
    occasionally hovers just below zero, this does not strongly suggest
    non-proportionality. The overall trend appears stable, and the
    residuals provide visual reassurance that prior treatment's effect
    on hazard is consistent over time.

**Martingale Residuals for Spline Model**

```{r}
# Residuals vs linear predictor
veteran$martingale_spline <- residuals(cox_spline, type = "martingale")
veteran$linear_pred_spline <- predict(cox_spline, type = "lp")

ggplot(veteran, aes(x = linear_pred_spline, y = martingale_spline, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Linear Predictor\n(Spline Model)",
    x = "Linear Predictor (Log Hazard)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The plot displays Martingale residuals against the linear predictor (log
hazard) for the spline-transformed Cox model. The residuals appear
randomly scattered around the horizontal zero line, without a
discernible systematic pattern or curvature. This suggests that the
model has adequately captured the relationship between the covariates
and the outcome.

The use of restricted cubic splines for age appears to have addressed
any previously observed non-linearity, as there is no strong evidence of
model misfit. The distribution of residuals across treatment groups also
seems relatively balanced, with no obvious separation or
heteroscedasticity.

The model’s functional form is appropriate, and the spline
transformation has improved the model’s fit by flexibly accommodating
the non-linear effect of age.

### Model Comparison: Linear vs Spline Age Term

We now compare the original multivariable Cox model with age modeled
linearly to the updated model using a spline transformation for age.
This allows us to formally test whether the added flexibility of the
spline significantly improves model fit.

```{r}
# Compare linear age model vs spline-transformed age model
anova(cox_multivariable, cox_spline, test = "LRT")
```

A likelihood ratio test comparing a model with age modeled linearly to
one using restricted cubic splines (df = 3) showed a statistically
significant improvement in fit (χ² = 7.91, df = 2, p = 0.019). This
suggests that the effect of age on survival is non-linear, and that
modeling age with splines better captures its relationship with hazard.

**Visualizing the Effect of Spline-Transformed Age**

This section uses 'termplot()' to visualize how the hazard changes
across age when modeled as a spline. This provides an intuitive look at
the non-linear relationship.

::: {.callout-note title="Understanding ANOVA in Cox Regression" collapse="true"}
In the context of **Cox proportional hazards models**, ANOVA isn't used
in the traditional sense (like comparing means). Instead, it's a way of
**comparing nested models** using their **log-likelihoods**.

This is typically done using the **Likelihood Ratio Test (LRT)**, which
answers the question:

> *"Does the more complex model fit significantly better than the
> simpler one?"*

**🔍 What the Output Shows:**

-   **Model 1:** The **simpler model** (fewer terms or more constrained
    terms)
-   **Model 2:** The **more complex model** (extra variables or flexible
    terms, e.g., splines)
-   **Log-likelihoods:** How well each model fits the data (less
    negative = better)
-   **Chisq:** Computed as `2 × (LL₂ − LL₁)`\
-   **Degrees of freedom:** Difference in number of parameters
-   **p-value:** Significance of improvement in fit

**✅ How to Interpret:**

-   **p \< 0.05** → The complex model fits significantly better → **keep
    it**
-   **p ≥ 0.05** → No meaningful improvement → **stick with the simpler
    model**

This approach is especially useful for: - Deciding whether to model
continuous variables **non-linearly** (e.g., with splines) - Evaluating
whether new covariates **add predictive value** - Performing **stepwise
model refinement**

📦 **Takeaway:**\
Use ANOVA/LRT in Cox regression to guide decisions about **model
complexity vs parsimony**, balancing better fit with interpretability.
:::

```{r}
termplot(cox_spline, se = TRUE, col.term = "blue")
```

The term plots from the spline-based Cox proportional hazards model
illustrate the partial effect of each covariate on the log hazard:

-   Age (modeled using natural splines with 3 degrees of freedom)
    exhibits a non-linear association with the log hazard. The curve
    displays a shallow U-shape, with the lowest hazard around age 55,
    and a gradual increase in hazard at both younger and older ages.
    While the trend is suggestive, the wide confidence
    intervals—particularly at the distribution tails—indicate
    uncertainty in the precise shape of this relationship.

-   Treatment Group (trt) is represented as a binary variable with two
    flat lines, one for each group. The narrow confidence intervals
    overlap with zero, implying no statistically significant difference
    in log hazard between the intervention and control arms in this
    model.

-   Prior Treatment (prior) also shows a flat trend across the range of
    prior treatments, with confidence bands consistently containing
    zero. This suggests that the number of prior therapies does not
    contribute meaningfully to hazard estimation when adjusting for age
    and treatment.

These visualizations support the statistical findings from the model
summary. Age is the only covariate showing a potentially complex,
non-linear effect, while treatment and prior exposure show no clear
independent impact on survival in this cohort.

::: {.callout-note title="How to Interpret Term Plots in Cox Models" collapse="true"}
**Term plots** visualize the *partial effect* of each covariate in a
fitted model while holding other variables constant. They are especially
helpful for interpreting transformations like splines.

**🔹 What is a term plot?**\
A term plot shows how the **log hazard** (on the y-axis) changes with a
covariate (on the x-axis), after accounting for all other variables in
the model.

-   The **blue line** is the estimated partial effect of that variable.
-   The **dashed orange lines** show the 95% confidence interval.
-   The interpretation depends on the type of variable:

**🔸 Continuous Variables (e.g., Age):**

-   For **linear terms**, the blue line should be roughly straight.
-   For **splines** or transformed terms:
    -   A *curved* blue line indicates a non-linear relationship.
    -   Areas where the confidence band **does not include 0** suggest a
        statistically significant effect on the log hazard.
    -   Patterns like U-shapes, inflection points, or steep climbs
        reveal how risk evolves across the covariate range.

**🔸 Binary Variables (e.g., Treatment, Prior Therapy):**

-   These are shown as **two horizontal segments**:
    -   One for each group (e.g., Control = 0, Intervention = 1).
    -   The **difference between the two** shows the log hazard shift
        due to that variable.
-   If the confidence intervals **include zero**, there's no strong
    evidence that the variable affects the hazard.

**✅ Best Uses:**

-   Confirming that spline-transformed terms behave as expected.
-   Checking if a linear term should be transformed (e.g., age might be
    better modeled with splines).
-   Double-checking whether binary variables have any meaningful effect.

🧠 *Tip:* Term plots don't replace statistical tests, but they
**visualize model estimates** clearly and intuitively.
:::

# Interaction or Additive Effects?

In Cox regression, interaction terms allow us to test whether the effect
of one covariate on the hazard changes depending on the level of another
covariate. If significant, this suggests a departure from simple
additivity and may require stratified or more complex modelling.

::: {.callout-note title="Understanding Interaction Effects in Cox Models" collapse="true"}
In regression models—including **Cox proportional hazards models**—an
*interaction effect* tests whether the relationship between a covariate
and the outcome changes depending on the level of another covariate.

In practical terms, interaction terms allow us to ask:\
\> *“Does the effect of Variable A on survival depend on Variable B?”*

*Why Test for Interaction?*

-   To explore whether **treatment effects vary across subgroups**
    (e.g., age, gender, biomarker levels).
-   To identify **non-additive relationships** that a main-effects-only
    model would miss.
-   To **refine risk stratification** or support personalized approaches
    to care.

*What It Tells Us*

In the Cox model, including an interaction (e.g., `treatment * age`)
allows the **hazard ratio for treatment** to vary at different values of
age. If the interaction term is statistically significant, it suggests
the **effect of treatment is modified by age**.

This can reveal: - A treatment that works well in younger patients but
poorly in older ones. - A covariate that only has an impact in the
presence of another factor. - Complex survival dynamics that are
**hidden in simpler additive models**.

*Interpreting Output*

-   A significant interaction coefficient indicates a **departure from
    additivity**.
-   It's essential to **cautiously interpret main effects** in the
    presence of interactions, as their meaning may be conditional.
-   Visualization (e.g., stratified survival curves or effect plots) can
    aid interpretation.

*When to Use*

-   After fitting and interpreting a main-effects model.
-   When clinical theory or prior evidence suggests effect modification.
-   When residual plots or diagnostics hint at **unexplained
    heterogeneity**.
:::

## Interaction of Treatment and Age

```{r}
# Fit a cox model with interaction between treatment and age 
cox_interaction <- coxph(Surv(time, status) ~ trt * ns(age, df = 3) + prior, data = veteran)
summary(cox_interaction)
```

We tested for interaction between treatment group and age (modeled using
natural splines with 3 degrees of freedom) to explore whether the effect
of treatment on survival varied across age groups. The interaction model
revealed a statistically significant interaction for the second spline
basis term (p = 0.037), suggesting a possible age-dependent treatment
effect.

The overall Wald test for the model approached significance (p = 0.05),
while the likelihood ratio test was marginal (p = 0.07), indicating that
the model with interaction terms may provide a better fit than the
additive model. These findings suggest that the benefit (or lack
thereof) of treatment may differ depending on patient age, warranting
further exploration or stratified modelling.

**Schoenfeld Residuals for Interaction Model**

```{r}
# Test proportional hazards assumption for the interaction model
cox.zph(cox_interaction)
```

The Schoenfeld residual test was used to assess whether the proportional
hazards assumption holds for each covariate and interaction term in the
model.

-   trt: χ² = 1.39, p = 0.24 → No evidence of time-varying effects for
    treatment group.

-   ns(age, df = 3): χ² = 3.90, p = 0.27 → The spline-transformed age
    term shows no strong time dependency.

-   prior: χ² = 2.42, p = 0.12 → Also not statistically significant.

-   trt × ns(age, df = 3): χ² = 5.37, p = 0.15 → The interaction between
    age and treatment shows no significant violation of proportionality
    either.

-   GLOBAL test: χ² = 8.94 on 8 df, p = 0.35 → The model as a whole
    satisfies the PH assumption.

There is no strong evidence that the proportional hazards assumption is
violated in this interaction model. The covariate effects, including the
interaction, appear stable over time.

```{r}
# Visual check of Schoenfeld residuals for interaction model
plot(cox.zph(cox_interaction))
```

The global test using cox.zph() indicated no significant violation of
the PH assumption (χ² = 8.94, df = 8, p = 0.35). Individual covariate
tests also returned non-significant p-values, suggesting that the
estimated hazard ratios remain approximately constant over time.

Visual inspection of the Schoenfeld residual plots supported this
conclusion. The smoothed curves for treatment, age (spline terms), prior
treatment, and their interaction remained relatively flat and within
their 95% confidence bands. While a few outliers were
present—particularly in the treatment and age components—these did not
indicate strong evidence of non-proportionality.

Taken together, these results suggest that the proportional hazards
assumption is reasonably met for all terms in the interaction model.

**Martingale Residuals for Interaction Model**

The Martingale residuals from the interaction model were plotted against
the linear predictor (log hazard) to assess model fit and identify any
potential patterns or outliers.

```{r}
# Residuals vs linear predictor
veteran$martingale_interaction <- residuals(cox_interaction, type = "martingale")
veteran$linear_pred_interaction <- predict(cox_interaction, type = "lp")
ggplot(veteran, aes(x = linear_pred_interaction, y = martingale_interaction, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Linear Predictor\n(Interaction Model)",
    x = "Linear Predictor (Log Hazard)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The Martingale residuals plot from the interaction model reveals a
concentration of predicted log hazard values for the control group
between 0.5 and 1.0, whereas predictions for the intervention group span
a wider range. This may reflect the influence of the age-by-treatment
interaction term, which adds complexity primarily for patients receiving
the intervention. The residuals themselves are randomly scattered around
zero, suggesting that the functional form of the covariates is
appropriately specified and that no major nonlinearity remains
unaccounted for.

**Deviance Residuals for Interaction Model**

Deviance residuals were generated to further assess the model fit and
identify any potential outliers or influential observations.

::: {.callout-note title="Understanding Deviance Residuals in Cox Models" collapse="true"}
**Deviance residuals** are a type of diagnostic measure used to assess
how well a Cox proportional hazards model fits the observed survival
data. Each deviance residual reflects the discrepancy between the
observed outcome (event or censoring) and what the model predicts for an
individual, after accounting for all covariates.

🔍 Why we use them: - **Outlier detection**: Large residuals (positive
or negative) may indicate subjects whose survival times were poorly
predicted by the model. - **Model fit check**: If residuals cluster
systematically or display non-random patterns when plotted against the
linear predictor, it may suggest issues such as: - Mis-specification of
functional form (e.g., nonlinear effects not properly captured), -
Omitted interactions or variables, - Violation of model assumptions.

📊 How to interpret: - Deviance residuals are approximately normally
distributed around 0. - Values far from zero may indicate poor fit for
specific observations. - A balanced, symmetrical scatter around the
horizontal axis suggests a well-fitting model.

🧠 When to use: - After fitting a Cox model, especially when refining
the covariate structure. - In combination with **martingale** and
**Schoenfeld residuals** to build a comprehensive diagnostic view. -
Before finalizing your model, particularly if you are considering
transformations or interactions.

💡 Bonus tip: Unlike martingale residuals, deviance residuals are
**symmetrically distributed** and therefore better suited for
identifying both under- and over-predictions, making them particularly
useful for graphical diagnostics.
:::

```{r}
# Deviance residuals
veteran$deviance_interaction <- residuals(cox_interaction, type = "deviance")
# Plot deviance residuals
ggplot(veteran, aes(x = linear_pred_interaction, y = deviance_interaction, color = trt)) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) + # Adding jitter to avoid overlap
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Deviance Residuals vs Linear Predictor\n(Interaction Model)",
    x = "Linear Predictor (Log Hazard)",
    y = "Deviance Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The deviance residual plot provides a diagnostic check on the adequacy
of the Cox proportional hazards model that includes an interaction
between treatment and age. In this plot, residuals are symmetrically
distributed around zero with no discernible curvature or trend,
suggesting that the model fits the data reasonably well.

Notably, the intervention group shows a broader range of linear
predictor values (i.e., estimated log hazard), indicating greater
variability in predicted risk compared to the control group, which
clusters tightly between 0.5 and 1.0. Despite this asymmetry in
predictor spread, the vertical dispersion of residuals remains balanced,
and no strong outliers or patterns emerge. This supports the
appropriateness of the interaction model and suggests no serious
violations of model assumptions related to functional form or
influential observations.

Overall, the deviance residuals confirm that the model adequately
captures the observed survival data, with no clear evidence of misfit or
heteroscedasticity.

## Interaction of Treatment and Prior Therapy

Prior treatments might alter a patient’s response to new therapies
(e.g., due to resistance, immune suppression, or cumulative toxicity).

```{r}
# Fit a cox model with interaction between treatment and prior therapy
cox_interaction_prior <- coxph(Surv(time, status) ~ trt * prior + ns(age, df = 3), data = veteran)
summary(cox_interaction_prior)
```


A Cox proportional hazards model including an interaction between treatment and prior therapy was fitted to assess potential effect modification. The interaction term (trt × prior) yielded a hazard ratio of 0.94 (95% CI: 0.87–1.02, p = 0.135), suggesting that the effect of treatment may be modestly attenuated in patients who had received prior therapy. However, this association did not reach statistical significance. Model-level tests (Likelihood Ratio, Wald, and Score) were borderline (p = 0.07–0.08), indicating weak evidence that the inclusion of an interaction term may improve model fit. While these findings are not conclusive in this sample, they raise the possibility of effect modification that could be explored further in larger, more adequately powered studies.

**Schoenfeld Residuals for Interaction with Prior Therapy**

```{r}
# Test proportional hazards assumption for the interaction with prior therapy
cox.zph(cox_interaction_prior)
```

The proportional hazards assumption was assessed using Schoenfeld residuals. The global test did not indicate violation (p = 0.171), supporting the use of the Cox model. Most individual terms showed no evidence of non-proportionality, though the interaction between treatment and prior therapy (trt:prior) was borderline (p = 0.054). This may suggest some time-dependent behavior in how prior treatment modifies the treatment effect, though the evidence is not conclusive.

```{r}
# Visual check of Schoenfeld residuals for interaction with prior therapy
plot(cox.zph(cox_interaction_prior))
```

To assess whether the proportional hazards (PH) assumption holds in the model including an interaction between treatment and prior therapy, Schoenfeld residuals were examined both statistically and graphically.

The global test for the PH assumption was not significant (p = 0.17), and all individual covariates—including the interaction term—also showed non-significant test results (all p > 0.05), suggesting no strong evidence of time-varying effects.

Visual inspection of the residual plots supports this conclusion. For all variables, including trt, prior, age (splines), and the interaction term trt × prior, the smoothed lines remain relatively flat over time, and their 95% confidence bands generally include zero. While the plots for trt and prior show some scatter at the extremes—with several residuals appearing as outliers above and below the bands—these deviations do not appear systematic or directional.

In particular, the clusters of points outside the confidence bounds likely reflect the influence of a small number of high-leverage observations rather than a violation of the PH assumption. These artifacts are common in small-to-moderate datasets, especially when stratifying by interactions.

Conclusion: There is no strong statistical or visual evidence to reject the PH assumption for this interaction model. However, the presence of some outlier residuals may warrant further sensitivity analysis in larger samples.

**Martingale Residuals for Interaction with Prior Therapy**

```{r}
# Residuals vs linear predictor
veteran$martingale_interaction_prior <- residuals(cox_interaction_prior, type = "martingale")
veteran$linear_pred_interaction_prior <- predict(cox_interaction_prior, type = "lp")

ggplot(veteran, aes(x = linear_pred_interaction_prior, y = martingale_interaction_prior, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Linear Predictor\n(Interaction with Prior Therapy)",
    x = "Linear Predictor (Log Hazard)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The martingale residual plot shows the residuals plotted against the linear predictor (log hazard), grouped by treatment arm. In this context, the martingale residuals help assess the model's functional form — particularly whether the covariates are correctly specified (e.g., linearity assumptions, missing covariates, or poorly modeled interactions).
Key Observations:

_Vertical Spread (Residuals Axis):_
There is substantial vertical scatter in the residuals, with several points deviating far from the horizontal line at 0. This is expected in survival models, particularly with smaller datasets. However, there is no clear systematic pattern, such as curvature or directional trend, that would suggest strong non-linearity or misspecification.

_Horizontal Clustering (Linear Predictor Axis):_
As with previous models, a dense cluster of observations is seen between approximately 0 and 1 on the linear predictor scale, mostly corresponding to the control group. This may reflect limited variation in the covariates for this subgroup, or overlapping risk profiles. The intervention group displays more dispersion, which aligns with the interaction term — suggesting greater heterogeneity in risk when prior treatment is included in the model.

_Overall Interpretation:_
The lack of a strong non-random pattern supports the adequacy of the covariate specification, including the interaction term. However, the wide spread and presence of extreme residuals imply potential outliers or influential cases that might warrant further sensitivity analysis in larger samples.

**Deviance Residuals for Interaction with Prior Therapy**

```{r}
# Deviance residuals
veteran$deviance_interaction_prior <- residuals(cox_interaction_prior, type = "deviance")

# Plot deviance residuals
ggplot(veteran, aes(x = linear_pred_interaction_prior, y = deviance_interaction_prior, color = trt)) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) + # Adding jitter to avoid overlap
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Deviance Residuals vs Linear Predictor\n(Interaction with Prior Therapy)",
    x = "Linear Predictor (Log Hazard)",
    y = "Deviance Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The plot of deviance residuals against the linear predictor for the interaction model reveals no strong patterns or obvious deviations from randomness. Residuals are symmetrically scattered around the zero line, with similar vertical spread across the range of the linear predictor. This suggests that:

_Model Fit Appears Reasonable:_
There is no clear evidence of systematic bias, such as curvature, funneling, or clustering that might imply model misspecification, omitted variables, or heteroscedasticity.

_Symmetry Around Zero:_
The relatively balanced distribution of residuals above and below the zero line supports the assumption that the model is neither consistently under- nor over-estimating the risk.

_No Extreme Outliers:_
While a few points have larger residuals, there are no values that fall far outside the typical range (±3), indicating a lack of highly influential or poorly fit observations.

_Group Comparison:_
The intervention and control groups appear similarly distributed across the range of linear predictors, lending further support to the appropriateness of the model specification including the interaction term.

_Summary:_

The deviance residual plot offers no strong evidence of model misfit and supports the adequacy of the Cox model including the interaction between treatment and prior therapy. Residuals appear randomly and symmetrically distributed, suggesting that the model captures the underlying structure of the data well, at least within the limitations of this dataset.

# Model Comparison: Interaction vs Non-Interaction

Two extended models were fit to explore potential effect modification: one including an interaction between treatment and age (via natural splines), and another between treatment and prior therapy. While the treatment × age model showed a statistically significant interaction term, the overall model-level tests remained marginal (Wald p = 0.05, Score p = 0.03), with no compelling evidence of meaningful improvement in predictive performance. The treatment × prior therapy interaction suggested a potential moderating effect, but this did not reach statistical significance (p = 0.135), and proportional hazards assumptions were borderline (p = 0.054).

When comparing these interaction models to the simpler additive model (trt + prior + spline-transformed age), neither interaction offered substantial improvement in model fit or interpretability. Residual diagnostics also did not reveal meaningful deviations. As such, the additive model remains a strong candidate for final interpretation.

# Final Model Selection and Interpretation

After evaluating a series of Cox proportional hazards models—including univariable, multivariable, and interaction models, the additive multivariable model with treatment group, prior therapy, and age modeled using restricted cubic splines, was selected as the final model for interpretation.

Rationale for Model Selection:

Interaction terms (treatment × age, treatment × prior therapy) were explored but did not provide substantial or statistically significant improvement in model fit, as indicated by likelihood ratio tests and residual diagnostics.
Residual analyses (Martingale and deviance residuals) and proportional hazards tests showed no strong evidence of model misfit or violation of assumptions in the additive model.
Model parsimony and interpretability: The additive model is simpler, easier to interpret, and adequately captures the observed data structure without unnecessary complexity.
Interpretation of the Final Model:

Treatment group: There was no statistically significant difference in survival between the intervention and control arms after adjusting for age and prior therapy (hazard ratio ≈ 1, p > 0.6).
Age: Modeled flexibly with splines, age did not show a strong or consistent association with hazard, and the effect was not statistically significant.
Prior therapy: Similarly, prior treatment was not a significant predictor of survival in this cohort.

Overall, the final model suggests that, in this dataset, neither the experimental treatment, age, nor prior therapy were independently associated with improved or worsened survival among veterans with advanced lung cancer. The model fits the data well, satisfies key assumptions, and provides a robust basis for interpretation, though the lack of significant findings may reflect limited sample size or underlying clinical similarities between groups.

# Conclusion

# Limitations
