---
title: "Survival Analysis: Veteran Lung Cancer Patients"
author: "Julian Chung"
format:
  html:
    toc: true
    code-fold: true
    number-sections: true
    theme: cosmo
execute:
  echo: true
---
# 📘 Introduction

This dataset originates from a clinical trial conducted by the Veterans Administration in the United States to compare the effectiveness of two chemotherapy regimens in treating advanced, inoperable lung cancer. Patients were randomized to receive either a standard treatment or a test treatment (an experimental chemotherapy protocol), and followed until death or censoring. The primary outcome was overall survival, measured in days from treatment initiation.

The dataset contains information on:

    Treatment group (Standard vs Test),

    Cell type (squamous, small cell, adenocarcinoma, or large cell),

    Survival time and status,

    Karnofsky performance score (a proxy for baseline health),

    Time from diagnosis to study entry,

    Patient age, and

    Number of prior treatments received.

The data is widely used in survival analysis teaching due to its realistic structure, meaningful clinical context, and manageable size.

D Kalbfleisch and RL Prentice (1980), The Statistical Analysis of Failure Time Data. Wiley, New York.

# 📦 Load Packages

```{r}
library(survival)
library(survminer)
library(dplyr)
library(ggplot2)
library(broom)

```

# 🧬 Load & Inspect Data

```{r}
# Load the dataset
veteran <- survival::veteran %>%
  mutate(trt = factor(trt, levels = c(1, 2), labels = c("Control", "Intervention")))

# Peek at the structure and summary
glimpse(veteran)
summary(veteran)
```

# 📈 Kaplan-Meier Survival Curve

Firstly we estimate the **survival function** for patients using the Kaplan-Meier estimator. This provides a non-parametric view of survival over time, stratified by treatment group.

```{r}
# Create the survival object
surv_obj <- Surv(time = veteran$time, event = veteran$status)

# Fit Kaplan-Meier curve stratified by treatment
km_fit <- survfit(surv_obj ~ trt, data = veteran)

# View basic survival summary
summary(km_fit)

```
```{r}
# Name the treatment groups
veteran <- survival::veteran %>%
  mutate(trt = factor(trt, 
                      levels = c(1, 2), 
                      labels = c("Control", "Intervention")))

# Create survival variable
surv_obj <- Surv(time = veteran$time, event = veteran$status)

# Fit Kaplan-Meier model
km_fit <- survfit(surv_obj ~ trt, data = veteran)

# Plot the KM curve
ggsurvplot(
  km_fit,
  data = veteran,
  conf.int = TRUE,
  pval = TRUE,
  pval.size = 4,
  risk.table = TRUE,
  risk.table.title = "Number at Risk",
  risk.table.col = "strata",
  risk.table.y.text.col = TRUE,
  risk.table.fontsize = 4,
  legend.title = "Treatment Group",
  legend.labs = c("Control", "Intervention"),
  xlab = "Time (days)",
  ylab = "Survival Probability",
  palette = c("#A569BD", "#45B39D"),
  title = "Kaplan-Meier Survival Curve: VA Lung Cancer Trial",
  ggtheme = theme_minimal(base_size = 13) + 
    theme(
      legend.position = "top",
      plot.title = element_text(hjust = 0.5, face = "bold"),
      panel.grid.major = element_line(color = "grey80", linewidth = 0.4),
      panel.grid.minor = element_line(color = "grey90", linewidth = 0.2)
    ),
  tables.theme = theme(  # Trying to tweak the risk table
    axis.text.y = element_text(size = 10, margin = margin(r = 12), lineheight = 2)
  )
)
```
### 🧠 Interpretation

The Kaplan-Meier plot shows survival curves for patients receiving standard vs test treatment.

**X-Axis: Time (days)**

The X-axis shows the time since enrollment in days. The furthest value to the right (~1000 days) marks the maximum observed survival time in the dataset. Shorter durations towards the left reflect early mortality events. 

**Y-Axis: Survival Probability**

The Y-Axis is the Kaplan-Meier estimate of the probability that a patient survives beyond a certain timepoint. The axis starts at 1 (100% survival at time 0), and decreases with each event (death).

**Curves (Blue = Standard, Orange = Test)**

Each step downward represents a death. Tick marks would represent a censored individual (few in this dataset). 

**Shaded Area: 95% Confidence Interval** 

Around each curve is a shaded band indicating **uncertainty** in the estimate. Wider intervals to the right of the chart reflect fewer patients remaining and hence more variability in survival probability. 

**p = 0.93** 

The log-rank test result tests the null hypothesis: there is no difference in survival between the groups. The p-value of 0.93 provides strong evidence _against_ rejecting the H0. In other words, there is no statistically significant difference in survival between the Standard and Test treatments. If the curves had diverged strongly at some point, the P-value would be lower. 

**"Number at Risk" Table**

This shows how many participatns in each treatment group are still alive and uncensored at various timepoints (0, 250, 500, 750, 1000). As time progresses the number declines due to deaths or censoring. Few participants remain beyond 500 days, which is why the Confidence Intervals widen later in the graph. 

- The **p-value** is from a log-rank test (H₀: no difference in survival curves).
- The **shaded areas** represent 95% confidence intervals.
- The **risk table** shows the number of patients still at risk at each time point.

Visually, if the curves separate early and remain apart, it suggests a potential treatment effect.

# 📊 Cox Proportional Hazards Model

```{r}
# Fit the Cox Proportional Hazards model
cox_model <- coxph(surv_obj ~ trt, data = veteran) # Where trt is a binary treatment variable with levels Control and Intervention

# Summarize the model
summary(cox_model)

```

| Term          | Value          | What It Means                                                                                                                                                                         |
| ------------- | -------------- | ------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- |
| **coef**      | 0.01774        | This is the log hazard ratio (log(HR)) comparing the *Intervention* group to *Control*. Close to 0 means almost no effect.                                                            |
| **exp(coef)** | 1.01790        | This is the hazard ratio (HR). A value of 1.018 means the hazard (risk of death) is 1.8% higher in the Intervention group compared to Control. However at this level the value is insigificant. |
| **se(coef)**  | 0.18066        | The standard error. This gives us the variability around the estimate.                                                                                                                |
| **z**         | 0.098          | The Wald test statistic. The low value implies no stastical significange.                                                                                     |
| **p-value**   | 0.922          | Far above the conventional 0.05 threshold. This is very strong evidence *against* any meaningful treatment effect.                                                           |
| **95% CI**    | (0.7144, 1.45) | Wide confidence interval, spanning both below and above 1—so we can’t rule out either harm or benefit.                                                                                |
**Concordance** The C-index, or concordance statistic is the probability that a candomly hosen subject who died earlier had a higher predicted risk. At 0.525 it is random chance, suggested poor predictive power. 

**Interpretation of the Cox Model Results**

The Cox proportional hazards model found no statistically significant difference in survival between the Control and Intervention groups (HR = 1.02, 95% CI: 0.71–1.45, p = 0.922). The hazard ratio close to 1 suggests that the treatment had no appreciable effect on the risk of death. Additionally, the wide confidence interval includes both potential harm and benefit, indicating uncertainty in the estimate. The concordance statistic (0.525) implies the model has poor discriminative ability, performing only slightly better than random chance. Overall, the analysis provides no evidence that the intervention improves survival outcomes in this dataset.

```{r}
# Use ggadjustedcurves for plotting adjusted survival curves directly from the Cox model
ggadjustedcurves(
  cox_model,                     # The fitted Cox model
  data = veteran,                # The original data used to fit the model
  variable = "trt",              # The variable to plot adjusted curves for
  method = "average",            # Method for adjusting (average effect of other covariates)
  conf.int = TRUE,
  legend.labs = c("Control", "Intervention"),
  legend.title = "Treatment Group",
  risk.table = TRUE,             # Include the risk table
  risk.table.title = "Number at Risk",
  xlab = "Time (days)",
  ylab = "Adjusted Survival Probability",
  title = "Cox Model Adjusted Survival Curves",
  palette = c("#A569BD", "#45B39D"),
  ggtheme = theme_minimal(base_size = 13) +
    theme(
      legend.position = "top",
      plot.title = element_text(hjust = 0.5, face = "bold"),
      panel.grid.major = element_line(color = "grey80", linewidth = 0.4),
      panel.grid.minor = element_line(color = "grey90", linewidth = 0.2)
    ),
  tables.theme = theme(
    axis.text.y = element_text(size = 10, margin = margin(r = 12), lineheight = 2)
  )
)
```

::: {.callout-tip collapse="true" title="Why use `ggadjustedcurves()` for this plot"}
For plotting survival curves adjusted for covariates from a Cox Proportional Hazards model, the ggadjustedcurves() function from the survminer package offers one of the most direct and statistically appropriate approaches—particularly as model complexity increases.

    Statistical validity: ggadjustedcurves() uses the fitted cox_model to compute adjusted survival probabilities for the variable of interest (in this case, trt). When method = "average" is specified (the default when covariates are present), it returns marginal estimates—appropriate for interpreting average treatment effects across the population. In our simpler model (~ trt), it effectively plots the model-predicted survival curve for each treatment group.

    Robustness: This function is purpose-built for adjusted curve plotting and avoids many of the pitfalls seen with manually passing newdata to survfit() and plotting with ggsurvplot(), such as misaligned strata or legend issues. It correctly associates the model, the data, and the plot elements, including the risk table.

    Simplicity: It eliminates the need for manually constructing a new_trt data frame or calling survfit() separately, resulting in cleaner, more reproducible code—particularly useful in collaborative or long-term projects.
:::

The Cox-adjusted survival curves for the Control and Intervention groups are almost perfectly overlapping, indicating no meaningful difference in predicted survival probabilities between the two treatment arms. This is consistent with the hazard ratio from the Cox model, which was very close to 1, and not statistically significant. Visually, the indistinguishability of the curves reinforces the conclusion that the treatment had no observable impact on survival in this dataset.

::: {.tip title="Why do the lines overlap?"}
The survival curves for the Control and Intervention groups overlap almost entirely in this plot. This is because the Cox proportional hazards model found no significant difference between the two groups in terms of survival (hazard ratio ≈ 1, *p* = 0.922). As a result, the model-predicted curves are nearly identical. The faint visibility of the Control group line is due to it being directly behind the Intervention group line.
:::

# 🔎 Model Diagnostics

## Checking the Proportional Hazards Assumption

```{r}
# Test the PH assumption
cox.zph(cox_model)
```

**Interpretation of PH Assumption Test (Schoenfeld residuals)**

- trt p = 0.06: This is slightly above the conventional threshold of 0.05. It suggest no strong evidence of violation of the proportional hazards (PH) assumptions for _treatment_. 
- GLOBAL p = 0.06: This tests the model overall. Not statistically significant so we do not reject the PH assumptions. 

**Porportional Hazards Assumption**

The proportional hazards assumption was assessed using Schoenfeld residuals via the cox.zph() function. The test for the treatment variable (trt) yielded a chi-square statistic of 3.54 with a p-value of 0.06, and the global test returned the same values (χ² = 3.54, df = 1, p = 0.06).

Although the p-values are slightly above the conventional significance threshold of 0.05, they do not provide sufficient evidence to reject the null hypothesis that the proportional hazards assumption holds. As such, the Cox proportional hazards model is considered appropriate for this analysis, though the borderline result suggests caution and may warrant further inspection in more complex models.



## Schoenfeld Residuals Plot

```{r}
plot(cox.zph(cox_model))
```

**What the Visual Diagnostic Shows:**

The y-axis: Scaled Schoenfeld residuals for trt. If the proportional hazards assumption holds, these residuals should fluctuate randomly around zero.

The x-axis: Time (or transformed time).

The line: A smoothed spline fit (typically a loess curve) showing the trend of residuals over time.

The confidence band: Helps you assess whether the trend significantly deviates from horizontal (i.e., no time-dependent effect).

A flat line (slope ≈ 0): Suggests the hazard ratio for trt does not change over time — good!

A wavy or sloped line: Suggests potential violation of the proportional hazards assumption — this would back up a significant p-value.

x-axis (Time): Time since baseline (likely in days, from your dataset).

y-axis (Beta(t) for trt): Time-varying estimate of the log hazard ratio for the treatment group (trt), based on Schoenfeld residuals.

The solid black line: A smoothed curve showing how the estimated coefficient for trt might change over time.

Dashed lines: Approximate 95% confidence bands around that smoothed estimate.

Open circles: Residuals for each individual event — they don’t need to follow a pattern, but their distribution can hint at model fit.

If the solid line is roughly flat and within the dashed lines, the effect of treatment does not change with time, and the PH assumption is reasonably upheld.

If the solid line drifts noticeably over time, or wanders outside the dashed lines, that's evidence that the treatment effect may change — a violation of the PH assumption.

In our case, the line is gently curved, but still well within the dashed boundaries. There’s no strong time trend — the log hazard ratio is hovering around zero and doesn’t appear to increase or decrease dramatically.

**Schoenfeld Residuals Interpretation**

To assess whether the proportional hazards (PH) assumption holds for the treatment variable (trt), we examined Schoenfeld residuals using both statistical and graphical diagnostics. The global test for proportionality yielded a p-value of 0.06, indicating no statistically significant violation of the PH assumption at the conventional α = 0.05 level, though it approaches significance. The smoothed residual plot for trt shows a relatively stable trend over time, with the estimated log hazard ratio remaining within the 95% confidence bands across the follow-up period. While there is some minor curvature in the smoothed line, it does not suggest substantial deviation from proportionality. These findings support the use of the Cox proportional hazards model for analyzing the effect of treatment on survival in this dataset.

## Martingale Residuals (for Functional Form of Covariates)

```{r}
# Generate Martingale residuals
martingale_resid <- residuals(cox_model, type = "martingale")

# Add residuals to your data
veteran$martingale <- martingale_resid

# Plot residuals against a continuous covariate (e.g., time or another variable)
# Here we plot against the linear predictor for simplicity
veteran$linear_pred <- predict(cox_model, type = "lp")

ggplot(veteran, aes(x = linear_pred, y = martingale, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Linear Predictor",
    x = "Linear Predictor (Log Hazard)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))


```
<details>
<summary>More about Martingale Residuals</summary>

Martingale residuals help us check whether the variables in our Cox model are specified correctly — particularly useful for continuous covariates. However, when the variable in question is binary (like our treatment group: Control vs Intervention), the Martingale residual plot will display two vertical stacks of points — one for each group.

In the plot above:

- The **x-axis** represents the linear predictor (log hazard). Because the treatment is binary, there are only two unique predictor values: one for the control group and one for the intervention group.
- The **y-axis** shows the Martingale residuals for each individual.
- The **dashed horizontal line at 0** acts as a visual guide to see if residuals are randomly scattered above and below it.

We want to see:
- No clear patterns or trends in the residuals,
- A reasonably symmetric spread around zero in each group,
- No outliers clustering far from the bulk of points.

In this case, both groups show a spread of residuals centered roughly around zero, with no curvature or skew that would suggest model misspecification. That’s a good sign — the model’s structure seems to handle the treatment variable well, and there's no obvious reason to suspect we need to transform it or add interaction terms.

</details>

The Martingale Residuals vs Linear Predictor plot helps us check the fit of our Cox model—specifically, whether our model’s assumptions about the shape of relationships between predictors and survival are holding up.

Martingale residuals were plotted against the linear predictor to assess model fit. As expected with a binary treatment variable, the plot showed two vertical bands of points corresponding to the Control and Intervention groups. The residuals appeared symmetrically distributed around zero within each group, with no obvious non-linear patterns or systematic bias. This suggests the model’s functional form for the treatment variable is appropriate.

## Functional Form of Covariates

::: {.tip title="Why we haven't checked the functional form of covariates"}

The Cox model assumes a linear relationship between continuous covariates and the log hazard. In our current model, only the treatment group (`trt`) is included, which is a categorical variable. 

If we were to build a more complex model including continuous predictors like age, Karnofsky score (`karno`), or time since diagnosis (`diagtime`), we would need to assess whether their relationship with the log hazard is truly linear. This is typically done by plotting Martingale residuals or using `ggcoxfunctional()`.

For now, since our model does not include continuous covariates, this step is not necessary — but it's an important part of model diagnostics when expanding our analysis.

:::

## Additional Covariates

### Age as a Continuous Covariate

#### Model Fit
<!-- Fit a Cox model including 'age' and show summary -->

#### Linearity Assessment
<!-- Martingale residual plot for age (continuous) -->

#### Proportional Hazards Assumption
<!-- cox.zph test + plot for age -->

#### Interpretation
<!-- Interpret hazard ratio for age and statistical significance -->


### Prior Treatment as a Binary Covariate

#### Model Fit
<!-- Fit a Cox model with 'prior' and show summary -->

#### Proportional Hazards Assumption
<!-- cox.zph test + plot for prior -->

#### Interpretation
<!-- Interpret hazard ratio for prior treatment -->


### Combined Multivariable Cox Model

#### Model Summary
<!-- Fit full model with trt + age + prior -->

#### Assumption Checks
<!-- Global cox.zph, Martingale residuals (age), etc. -->

#### Adjusted Survival Curves
<!-- ggadjustedcurves for trt (adjusted for age and prior) -->

#### Interpretation
<!-- Explain each variable’s adjusted effect -->

::: {.tip title="Why not use logistic regression?"}
Logistic regression ignores time-to-event and censoring. When working with survival data, Cox models are preferred as they preserve the time structure and allow for censored cases.
:::

# 🔀 Multivariable Cox Regression

To build on the univariable analysis of treatment effects, we next fit a multivariable Cox proportional hazards model that incorporates additional patient-level covariates. This allows us to estimate the effect of treatment on survival **while adjusting for other potential prognostic factors**, such as age and prior treatment history. By doing so, we account for potential confounding and improve the model’s explanatory power.
::: {.tip title="What is a multivariable Cox model and why are we using it?" collapse=true}

In a univariable Cox model, we examine the effect of **one predictor at a time** on the hazard of an event (e.g., death). While this can be useful for preliminary exploration, it doesn't account for the **influence of other variables** that might confound or modify the relationship.

A **multivariable Cox proportional hazards model** includes **multiple covariates** simultaneously. This allows us to:
- Estimate the **adjusted effect** of each variable while controlling for the others,
- Improve model accuracy and interpretability,
- Detect whether the apparent effect of one variable (e.g., treatment) persists after adjusting for others (e.g., age, comorbidities),
- Explore potential **interactions** or **effect modification**.

In this section, we'll first add a **continuous covariate** (`age`) and then a **binary variable** (`prior treatment`) to the model. We'll also assess whether the functional form of continuous variables is appropriate and ensure that model assumptions still hold.

:::

## 📥 Add Covariates (Age, Prior Treatment)

To expand the model beyond treatment alone, we now include additional patient characteristics: `age` (a continuous variable) and `prior` (indicating whether the patient received prior therapy, a binary variable). Including these covariates allows us to better account for baseline differences between patients, assess their independent associations with survival, and evaluate how these variables may influence the effect of treatment.

::: {.tip title="Why include more covariates?" collapse=true}

Adding additional covariates serves several important purposes in survival analysis:

- **Control for confounding:** Variables like age and prior treatment may influence both survival and the likelihood of receiving a particular treatment. Including them helps isolate the true effect of treatment.
- **Improve model accuracy:** Covariates that predict survival can increase the model’s explanatory power and reduce residual variation.
- **Uncover independent predictors:** Some variables (like prior therapy) may have a meaningful relationship with survival, regardless of treatment.
- **Assess effect modification:** Once multiple variables are included, we can test for interactions — for example, whether the effect of treatment depends on age or prior treatment status.

We’ll begin by examining `age` as a continuous covariate (with extra attention to its functional form), followed by `prior` as a binary covariate. We'll then combine them in a multivariable Cox model.

:::


### 🔁 Continuous Variable: Age

Age is a clinically relevant factor that may influence survival outcomes in lung cancer. In this section, we assess whether age is independently associated with survival time and verify that its relationship with the hazard is appropriately modeled as linear on the log-hazard scale.

::: {.tip title="Why model age carefully?" collapse=true}

Unlike categorical variables, continuous predictors like age require us to check their **functional form** in the model. The Cox model assumes a **linear relationship** between the covariate and the log hazard — meaning each 1-year increase in age is assumed to have the same proportional effect on the hazard.

If this assumption doesn't hold (e.g., if risk increases faster at older ages), we may need to use transformations or splines to accurately model the relationship.

We'll start by fitting a Cox model including age and then assess the functional form using Martingale residuals.

:::


#### 📈 Martingale Residuals for Age
### 🔁 Binary Variable: Prior Treatment

### 🔁 Binary Variable: Prior Treatment

We now evaluate the variable `prior`, which indicates whether a patient received prior therapy before enrollment in the study. This binary variable may capture important differences in disease severity, treatment history, or prognosis that could influence survival outcomes.

::: {.tip title="Why test binary covariates like 'prior'?" collapse=true}

Even though `prior` is binary, its inclusion allows us to control for historical differences in patient treatment, which may affect survival independently of the trial treatment (`trt`). 

For binary variables, we don’t need to assess the functional form, but we **do need to check the proportional hazards assumption** (as with any covariate in the Cox model). We'll interpret the hazard ratio as the relative difference in risk between those who received prior therapy and those who did not.

:::


#### 📈 Interaction or Additive Effects?

# 🧾 Model Summary & Interpretation
## 📌 Hazard Ratios
## 🧠 Clinical Interpretation

# 🧰 Appendix
## 📎 Theory: Why Cox Models?
## 📎 Details: Code Tips, Reproducibility Notes
## 📎 Export for Report Publishing

# 🧾 Conclusion

# 📌 Limitations
Briefly reflect on sample size, generalizability, or assumptions.
