---
title: "Veteran Lung Cancer Survival Analysis: Overview"
format:
  html:
    toc: true
execute:
  warning: false
  message: false
---

```{r setup, include=FALSE}
library(survival)
library(survminer)
library(dplyr)
library(forcats)
library(broom)
library(purrr)
library(ggplot2)
library(splines)
library(glue)
library(ggfortify)

# Cleaned dataset with factor labels used throughout
veteran2 <- survival::veteran %>%
  mutate(
    trt      = factor(trt, levels = c(1, 2), labels = c("Standard", "Experimental")),
    celltype = factor(celltype, levels = c("squamous", "smallcell", "adeno", "large"),
                      labels = c("Squamous", "Small cell", "Adenocarcinoma", "Large cell")),
    prior    = factor(prior, levels = c(0, 10), labels = c("No", "Yes"))
  )
```

# Introduction

This project analyses survival outcomes in the Veterans Administration lung
cancer trial, a classic dataset used to demonstrate time-to-event modelling in
R. The goal is to compare standard and experimental chemotherapy while showing a
full Cox proportional hazards workflow: exploratory survival curves, adjusted
modelling, and model checks.

Survival analysis is useful when the outcome is not just whether an event
happened, but when it happened. In this case, the event is death, and the
methods account for censoring as well as differences in follow-up time.

The dataset includes 137 male patients with advanced, inoperable lung cancer
randomised to standard or experimental treatment. It is compact enough to work
through end to end, but rich enough to show the main decisions involved in a
real survival analysis.

The project is organised into three parts:

- **Overview** — this page; a brief, high-level summary of the key findings and visuals.
- **Full Report** — step-by-step analysis, model building, and diagnostics.
- **Teaching & Learning Reference** — statistical theory, code explanations,
  and a reusable reference for future survival analyses.

# Study Overview

```{r descriptive-stats}
# KM medians overall and by arm
km_all <- survfit(Surv(time, status) ~ 1,   data = veteran2)
km_by  <- survfit(Surv(time, status) ~ trt, data = veteran2)
overall <- surv_median(km_all)
byarm   <- surv_median(km_by)

med_all   <- round(overall$median)
ci_all_l  <- round(overall$lower);  ci_all_u  <- round(overall$upper)
med_std   <- round(byarm$median[byarm$strata == "trt=Standard"])
ci_std_l  <- round(byarm$lower [byarm$strata == "trt=Standard"])
ci_std_u  <- round(byarm$upper [byarm$strata == "trt=Standard"])
med_exp   <- round(byarm$median[byarm$strata == "trt=Experimental"])
ci_exp_l  <- round(byarm$lower [byarm$strata == "trt=Experimental"])
ci_exp_u  <- round(byarm$upper [byarm$strata == "trt=Experimental"])

# Baseline summaries
n_total   <- nrow(veteran2)
n_by_trt  <- veteran2 %>% count(trt, name = "n")
n_std_n   <- n_by_trt$n[n_by_trt$trt == "Standard"]
n_exp_n   <- n_by_trt$n[n_by_trt$trt == "Experimental"]

q_age   <- quantile(veteran2$age,   probs = c(0.25, 0.5, 0.75))
q_karno <- quantile(veteran2$karno, probs = c(0.25, 0.5, 0.75))
med_age   <- round(q_age[2]);   iqr_age_l <- round(q_age[1]);   iqr_age_u <- round(q_age[3])
med_karno <- round(q_karno[2]); iqr_kar_l <- round(q_karno[1]); iqr_kar_u <- round(q_karno[3])

cell_pct  <- round(100 * prop.table(table(veteran2$celltype)), 1)
prior_pct <- round(100 * prop.table(table(veteran2$prior)),    1)
prior_yes <- unname(prior_pct["Yes"])

lr   <- survdiff(Surv(time, status) ~ trt, data = veteran2)
lr_p <- signif(1 - pchisq(lr$chisq, length(lr$n) - 1), 3)

deaths      <- sum(veteran2$status == 1)
censored    <- sum(veteran2$status == 0)
deaths_pct  <- round(100 * deaths   / n_total, 1)
cens_pct    <- round(100 * censored / n_total, 1)

cat(glue("
**Study population**
- {n_total} male patients; Standard = **{n_std_n}**, Experimental = **{n_exp_n}**
- Median age: **{med_age}** years (IQR: {iqr_age_l}–{iqr_age_u})
- Median Karnofsky score: **{med_karno}** (IQR: {iqr_kar_l}–{iqr_kar_u})
- Cell type: Squamous {cell_pct['Squamous']}%, Small cell {cell_pct['Small cell']}%,
  Adenocarcinoma {cell_pct['Adenocarcinoma']}%, Large cell {cell_pct['Large cell']}%
- Prior therapy: **{prior_yes}%** yes

**Survival outcomes**
- Median survival (all): **{med_all}** days (95% CI: {ci_all_l}–{ci_all_u})
- Standard arm: **{med_std}** days (95% CI: {ci_std_l}–{ci_std_u})
- Experimental arm: **{med_exp}** days (95% CI: {ci_exp_l}–{ci_exp_u})
- Log-rank *p* = **{lr_p}**

**Events**
- Deaths: **{deaths}** ({deaths_pct}%) | Censored: **{censored}** ({cens_pct}%)
"))
```

# Plot 1 — Survival by Treatment Group

The primary question of the trial: does the experimental regimen improve
survival over standard chemotherapy?

```{r km-treatment, fig.width=8, fig.height=6}
ggsurvplot(
  km_by,
  data        = veteran2,
  risk.table  = TRUE,
  pval        = TRUE,
  conf.int    = TRUE,
  palette     = c("#2E86AB", "#E84855"),
  legend.title = "Treatment",
  xlab        = "Time (days)",
  ylab        = "Survival probability",
  title       = "Kaplan–Meier Survival Curves by Treatment Arm",
  ggtheme     = theme_minimal(base_size = 13)
)
```

The survival curves for the two treatment arms overlap almost entirely
throughout follow-up. The log-rank test confirms no statistically significant
difference (p = `r lr_p`). The experimental chemotherapy did not improve
survival in this cohort.

# Plot 2 — Survival by Karnofsky Performance Score

Karnofsky score is a 0-100 measure of functional status. In this dataset, it is
one of the clearest predictors of survival.

```{r km-karnofsky, fig.width=8, fig.height=6}
veteran2$karno_group <- cut(
  veteran2$karno,
  breaks = c(0, 50, 70, 100),
  labels = c("Low (≤50)", "Medium (51–70)", "High (>70)"),
  right  = TRUE
)

km_karno <- survfit(Surv(time, status) ~ karno_group, data = veteran2)

ggsurvplot(
  km_karno,
  data         = veteran2,
  risk.table   = TRUE,
  pval         = TRUE,
  conf.int     = FALSE,
  palette      = c("#E84855", "#F4A259", "#2E86AB"),
  legend.title = "Karnofsky group",
  legend.labs  = c("Low (≤50)", "Medium (51–70)", "High (>70)"),
  xlab         = "Time (days)",
  ylab         = "Survival probability",
  title        = "Kaplan–Meier Survival Curves by Karnofsky Score Group",
  ggtheme      = theme_minimal(base_size = 13)
)
```

The separation between the groups is clear: patients with higher functional
status survived much longer than those with lower scores. That pattern is
strong, consistent, and carries through to the adjusted Cox model, where
Karnofsky score is the most important predictor in the analysis.

# Plot 3 — Final Cox Model: Adjusted Hazard Ratios

This plot shows the final adjusted Cox model, including treatment group, prior
therapy, age, Karnofsky score, and cell type.

```{r cox-forest, fig.width=9, fig.height=6}
cox_final <- coxph(
  Surv(time, status) ~ trt + prior + ns(age, df = 3) + karno + celltype,
  data = veteran2
)

ggforest(
  cox_final,
  data      = veteran2,
  main      = "Adjusted Hazard Ratios — Final Cox PH Model",
  fontsize  = 0.9,
  noDigits  = 2,
  refLabel  = "Reference"
)
```

The clearest signals in the model are Karnofsky score and tumour cell type.
Higher Karnofsky scores are associated with lower risk, while adenocarcinoma and
small cell carcinoma show substantially higher hazard than squamous cell
carcinoma. Even after adjustment, the treatment effect remains modest and
uncertain (HR ≈ 1.4, *p* = 0.097). Overall, the model discriminates much better
than a treatment-only model, with concordance improving from about 0.51 to
0.74.

# Plot 4 — Aalen Additive Model: Time-Varying Effects

This plot checks the main findings using an additive survival model rather than
the Cox model.

```{r aalen, fig.width=10, fig.height=8}
aa_fit <- aareg(
  Surv(time, status) ~ trt + celltype + karno + age + prior,
  data = veteran2
)

autoplot(aa_fit) +
  theme_minimal(base_size = 12) +
  labs(title = "Aalen Additive Model: Cumulative Regression Functions")
```

The result is broadly the same. Karnofsky score remains the strongest and most
stable predictor of survival, while treatment group and prior therapy contribute
little on their own. Seeing the same pattern across two different modelling
approaches adds confidence to the overall interpretation.
