---
title: "Survival Analysis of Veterans with Advanced Lung Cancer: A Cox Proportional Hazards Approach"
author: "Julian Chung"
format:
  html:
    toc: true
    code-fold: true
    number-sections: true
execute:
  echo: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

# Introduction

This report presents a survival analysis of patients with advanced,
inoperable lung cancer who participated in a randomized clinical trial
conducted by the Veterans Administration in the United States. The
objective of the study was to compare the effectiveness of a standard
chemotherapy protocol versus a test (experimental) treatment in
prolonging patient survival.

Participants were followed until death or censoring, with overall
survival time (in days from treatment initiation) as the primary
outcome. This dataset, originally introduced in The Statistical Analysis
of Failure Time Data by Kalbfleisch and Prentice (1980), has become a
foundational paper in the teaching survival analysis due to its manageable size,
clear structure, and strong clinical relevance.

This analysis explores survival outcomes using both Kaplan-Meier
estimation and Cox proportional hazards modeling, followed by diagnostic
checks to assess model assumptions and potential improvements. 

# Setup and Data Import

We begin by loading the necessary R packages and importing the veteran
lung cancer dataset from the `survival` package.

```{r}
# Load required libraries
library(survival) # Core survival analysis tools
library(survminer)  # Visualization and diagnostics
library(dplyr)  # Data wrangling
library(ggplot2)  # Plotting
library(broom)  # Tidying model output
library(splines) # Splines for flexible modeling

```


Next, the dataset is loaded and a quick transformation applied to label
the treatment and control groups.

```{r}
# Load the dataset
veteran <- survival::veteran %>%
  mutate(trt = factor(trt,
                      levels = c(1, 2),
                      labels = c("Control", "Intervention")))

# Peek at the structure and summary
# glimpse(veteran)
# summary(veteran)
```


**Variable Summary:**

Below is a summary of the key variable in the dataset.

| Variable | Type | Description | Notes |
|------------|------------|-------------------------|------------------------|
| `trt` | Factor | Treatment group: 1 = Standard, 2 = Test | Relabeled to `"Control"` and `"Intervention"` |
| `celltype` | Factor | Histologic type of lung cancer: squamous, small cell, adeno, large | Included in final model; squamous is the reference category |
| `time` | Numeric | Survival time in days | Outcome variable in the `Surv()` object |
| `status` | Numeric | Event indicator: 1 = died, 0 = censored | Used to define survival endpoint |
| `karno` | Numeric | Karnofsky performance score (proxy for baseline health status) | Included in final model as a continuous linear predictor |
| `diagtime` | Numeric | Time from diagnosis to study entry (in months) | Possibly useful in extended time-dependent models |
| `age` | Numeric | Patient age (in years) | Treated as **continuous**; **checked for linearity and splines** |
| `prior` | Numeric | Number of prior therapies (0 or 10) | Binary variable: 0 = none, 10 = prior treatment |

# Exploratory Survival Analysis

Before proceeding with formal regression modeling, we begin with a non-parametric exploration of the survival data using the Kaplan-Meier estimator. This allows us to visualize survival patterns and group differences without making assumptions about the underlying hazard function. 


## Kaplan-Meier Survival Curve

Firstly we estimate the **survival function** for patients using the
Kaplan-Meier estimator. This provides a non-parametric view of survival
over time, stratified by treatment group.

```{r}
# Create the survival object
surv_obj <- Surv(time = veteran$time, event = veteran$status)

# Fit Kaplan-Meier curve stratified by treatment
km_fit <- survfit(surv_obj ~ trt, data = veteran)

# View basic survival summary
# summary(km_fit)

```



```{r}
# Plot the KM curve
ggsurvplot(
  km_fit,
  data = veteran,
  conf.int = TRUE,
  pval = TRUE,
  pval.size = 4,
  risk.table = TRUE,
  risk.table.title = "Number at Risk",
  risk.table.col = "strata",
  risk.table.y.text.col = TRUE,
  risk.table.fontsize = 4,
  legend.title = "Treatment Group",
  legend.labs = c("Control", "Intervention"),
  xlab = "Time (days)",
  ylab = "Survival Probability",
  palette = c("#A569BD", "#45B39D"),
  title = "Kaplan-Meier Survival Curve: VA Lung Cancer Trial",
  ggtheme = theme_minimal(base_size = 13) + 
    theme(
      legend.position = "top",
      plot.title = element_text(hjust = 0.5, face = "bold"),
      panel.grid.major = element_line(color = "grey80", linewidth = 0.4),
      panel.grid.minor = element_line(color = "grey90", linewidth = 0.2)
    ),
  tables.theme = theme(  # Trying to tweak the risk table
    axis.text.y = element_text(size = 10, margin = margin(r = 12), lineheight = 2)
  )
)
```

**Interpretation of the Kaplan-Meier Plot**

The Kaplan-Meier plot shows the survival curves for patients in the
intervention and control arms. For both groups there is a steep decline
in survival probability during the first 200 days, indicating rapid
disease progression. The shaded areas represent the 95% confidence
intervals, which largely overlap throughout the follow-up period,
suggesting no significant difference in survival between treatment arms.
The log-rank test yields a p-value of 0.93, providing no evidence to
reject the null hypothesis of equal survival. Therefore, we conclude
that there is no statistically significant difference in survival
between the two chemotherapy regimens.


## Log-Rank Test

The log-rank test is a statistical test used to compare the survival
distributions of two or more groups. It assesses whether there are
significant differences in survival times between the groups, based on
the observed and expected number of events (deaths) at each time point.


```{r}
# Perform the log-rank test
log_rank_test <- survdiff(surv_obj ~ trt, data = veteran)
log_rank_test
```

```{r}
# Extract the p-value
1 - pchisq(log_rank_test$chisq, df = length(log_rank_test$n) - 1)
```


**Interpretation of the Log-Rank Test Results**

A log-rank test was performed to compare survival distributions between
the control and intervention treatment groups. The test yielded a
chi-square statistic of 0.008 with 1 degree of freedom, resulting in a
p-value of 0.93. This indicates no statistically significant difference
in survival between the two treatment arms. The observed and expected
number of events (deaths) were nearly identical in both groups, further
supporting the conclusion that survival outcomes did not differ
meaningfully between treatments in this sample.

However, the log-rank test is a univariate method — it does not account for covariates that may confound or modify treatment effects (e.g., age, prior therapy). Additionally, it assumes proportional hazards and constant hazard differences over time. These limitations justify the use of a Cox proportional hazards model, which allows for multivariable adjustment and more nuanced interpretation of the treatment effect in the context of patient characteristics.


## Why Proceed with the Cox Model?

While the Kaplan-Meier plot and log-rank test indicate no statistically significant difference in survival between treatment arms, these methods do not account for other variables that may affect outcomes. The Cox proportional hazards model allows us to:

- Adjust for confounding covariates (e.g., age, prior treatment)
- Estimate hazard ratios for each variable
- Explore possible interactions or time-varying effects

Thus, we proceed with a Cox model to assess whether treatment effects become more apparent after adjusting for relevant clinical factors.

# Cox Regression Models

The Cox proportional hazards model is a regression method for analysing time-to-event data. Unlike the Kaplan-Meier estimator, which provides a non-parametric, unadjusted view of survival, the Cox model allows us to examine the effect of multiple covariates on the hazard of death.

This model assumes that covariates have a multiplicative and time-constant effect on the hazard (the proportional hazards assumption). By estimating hazard ratios, we can assess whether factors such as treatment type, age, or prior therapy are independently associated with better or worse survival outcomes.

We begin with a simple model including only treatment group, and progressively build toward a multivariable model incorporating continuous and categorical variables. Each model is followed by diagnostic checks to assess assumptions and guide model refinement.


## Univariable Models

In this section, we fit a series of univariable Cox models to assess the effect of individual covariates on survival. This step helps us explore the independent contribution of each variable before building more complex multivariable models.

### Treatment Group Only

We begin with a univariable Cox model that includes only the treatment
variable (`trt`). This allows us to assess the effect of treatment on
survival without adjusting for any other covariates.

```{r}
# Fit the Cox Proportional Hazards model
cox_model <- coxph(surv_obj ~ trt, data = veteran) # Where trt is a binary treatment variable with levels Control and Intervention

# Summarize the model
summary(cox_model)

```


**Interpretation of the Cox Model Results**

The univariable Cox proportional hazards model found no statistically significant difference in survival between the control and intervention groups (HR = 1.02, 95% CI: 0.71–1.45, p = 0.92). The hazard ratio near 1 suggests little to no effect of the treatment on mortality risk. The wide confidence interval spans both potential benefit and harm, highlighting uncertainty in the estimate. The concordance statistic (0.525) indicates the model’s discriminative ability is only marginally better than chance. Overall, this analysis offers no evidence that the intervention improves survival in this sample.

```{r}
# Use ggadjustedcurves for plotting adjusted survival curves directly from the Cox model
ggadjustedcurves(
  cox_model,                     # The fitted Cox model
  data = veteran,                # The original data used to fit the model
  variable = "trt",              # The variable to plot adjusted curves for
  method = "average",            # Method for adjusting (average effect of other covariates)
  conf.int = TRUE,
  legend.labs = c("Control", "Intervention"),
  legend.title = "Treatment Group",
  risk.table = TRUE,             # Include the risk table
  risk.table.title = "Number at Risk",
  xlab = "Time (days)",
  ylab = "Adjusted Survival Probability",
  title = "Cox Model Adjusted Survival Curves",
  palette = c("#A569BD", "#45B39D"),
  ggtheme = theme_minimal(base_size = 13) +
    theme(
      legend.position = "top",
      plot.title = element_text(hjust = 0.5, face = "bold"),
      panel.grid.major = element_line(color = "grey80", linewidth = 0.4),
      panel.grid.minor = element_line(color = "grey90", linewidth = 0.2)
    ),
  tables.theme = theme(
    axis.text.y = element_text(size = 10, margin = margin(r = 12), lineheight = 2)
  )
)
```


**Interpretation of the Adjusted Survival Curves**

The Cox-adjusted survival curves for the Control and Intervention groups are nearly identical, indicating no meaningful difference in predicted survival probabilities between the treatment arms. This visual overlap aligns with the Cox model's hazard ratio, which was close to 1 and not statistically significant. Together, the model output and adjusted curves reinforce the conclusion that the intervention had no observable effect on survival in this cohort.


**Model Diagnostics**

Before interpreting the results of a Cox proportional hazards model, it’s important to ensure that the model’s assumptions are met. The two key assumptions are:

- The proportional hazards (PH) assumption: The effect of each covariate on the hazard is constant over time.

- Linearity of continuous covariates: Continuous covariates are assumed to have a linear relationship with the log hazard.

Violations of these assumptions can undermine the model’s predictive utility and lead to misleading conclusions.

In this section, we use the following diagnostic tools to assess model adequacy:

- Schoenfeld residuals to test the PH assumption.

- Martingale residuals to check for non-linearity and outliers.

- Functional form checks, particularly for continuous covariates like age.

- Where needed, we apply transformations (e.g., natural splines) to better capture nonlinear effects.

These checks help ensure that the model remains a sound and interpretable framework for estimating covariate effects on survival.

_Checking the Proportional Hazards Assumption_

A central assumption of the Cox proportional hazards model is that hazard ratios remain constant over time — meaning the relative risk between any two individuals does not change as time progresses. This is known as the proportional hazards (PH) assumption, and verifying it is crucial before interpreting model results.

**Schoenfeld Residuals**

To assess the PH assumption, we use Schoenfeld residuals, which evaluate whether covariate effects vary over time. If residuals show a systematic pattern against time, it may indicate a violation of the assumption.

We check this in two ways:

- Statistical tests: The cox.zph() function provides both a global test and tests for each covariate. A p-value < 0.05 suggests evidence against the PH assumption.

- Visual inspection: Plots of scaled Schoenfeld residuals include a smoothed line and 95% confidence bands. Systematic trends or large deviations from zero may indicate time-varying effects.

If either test indicates a violation, we may need to adjust the model using time-varying covariates, stratification, or alternative approaches.

```{r}
# Test the PH assumption
cox.zph(cox_model)
```


**Interpretation of PH Assumption Test (Schoenfeld residuals)**

- trt p = 0.06: Slightly above the conventional 0.05 threshold. This suggests no strong evidence that the effect of treatment changes over time.

- GLOBAL p = 0.06: The overall model does not show a statistically significant violation of the PH assumption.

Conclusion: We do not reject the PH assumption based on either the individual or global tests. The model is likely appropriate for interpretation without time-dependent adjustments.

**Schoenfeld Residuals Plot**

```{r}
plot(cox.zph(cox_model))
```


**Schoenfeld Residuals Interpretation**

The global test for proportional hazards yielded a p-value of 0.06, slightly above the conventional 0.05 threshold, suggesting no strong evidence of a violation. The smoothed plot of Schoenfeld residuals for the treatment variable shows mild curvature but remains well within the 95% confidence bands, with no clear time-dependent trend. Together, these results support the proportional hazards assumption for treatment and the suitability of the Cox model in this context.

**Martingale Residuals (for Functional Form of Covariates)**

Martingale residuals are used to evaluate whether continuous covariates have been correctly specified in the Cox model. By plotting these residuals against each covariate, we can detect departures from linearity on the log-hazard scale. Systematic patterns, such as curvature or clustering, may indicate the need for transformations (e.g., splines or categorization) to better capture the relationship.


```{r}
# Generate Martingale residuals
martingale_resid <- residuals(cox_model, type = "martingale")

# Add residuals to your data
veteran$martingale <- martingale_resid

# Plot residuals against a continuous covariate (e.g., time or another variable)
# Here we plot against the linear predictor for simplicity
veteran$linear_pred <- predict(cox_model, type = "lp")

ggplot(veteran, aes(x = linear_pred, y = martingale, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Linear Predictor",
    x = "Linear Predictor (Log Hazard)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))


```


**Interpretation of Martingale Residuals Plot**

The Martingale residual plot helps assess whether the model’s assumptions—particularly the functional form of covariates—are valid.

In this case, residuals were plotted against the linear predictor for a binary treatment variable. As expected, two vertical bands appear, corresponding to the Control and Intervention groups. The residuals are roughly symmetric around zero within each group, with no obvious curvature or systematic pattern. This supports the adequacy of the treatment variable’s specification in the model and suggests that no transformation is needed for this binary covariate.


**Additional Covariates**

To improve the model and better account for variability in survival outcomes, we now include additional patient characteristics: age (a continuous variable) and prior treatment history (prior, a binary variable). Adding these covariates allows us to control for baseline differences between patients, assess their independent associations with survival, and explore whether they influence the effect of treatment.


## Multivariable Models

To build on the univariable analysis of treatment effects, we next fit a
multivariable Cox proportional hazards model that incorporates
additional patient-level covariates. This allows us to estimate the
effect of treatment on survival **while adjusting for other potential
prognostic factors**, such as age and prior treatment history. By doing
so, we account for potential confounding and improve the model's
explanatory power.



### Continuous Variable: Age

Age is a clinically relevant factor that may influence survival outcomes
in lung cancer. In this section, we assess whether age is independently
associated with survival time and verify that its relationship with the log hazard is appropriately linear.


**Cox Model with Age**

```{r}
cox_age <- coxph(Surv(time, status) ~ trt + age, data = veteran)
summary(cox_age)
```

**Interpretation of the Cox Model with Age**

A Cox proportional hazards model was fitted to examine the relationship
between survival time and two predictors: treatment group (trt) and
patient age (age). The model included 137 patients, with 128 events
(deaths).

The hazard ratio for the intervention group was 0.996 (95% CI: 0.697 –
1.425; p = 0.984), suggesting virtually no difference in hazard compared
to the control group after adjusting for age. This effect was not
statistically significant.

For age, the hazard ratio was 1.008 (95% CI: 0.989 – 1.027; p = 0.436),
indicating that each additional year of age was associated with a 0.8%
increase in the hazard of death. However, this association was also not
statistically significant and the confidence interval included 1,
suggesting uncertainty about the true effect.

Overall, the model's ability to discriminate between patients with
longer vs shorter survival times was weak, with a concordance index of
0.514 (where 0.5 = random chance). The global model statistics —
including the likelihood ratio test, Wald test, and score test — were
all non-significant (p = 0.7), indicating that neither variable added
substantial predictive value to the model.

-   There is no evidence of a treatment effect after adjusting for age.

-   Age may have a small effect, but in this sample, it is not
    statistically or clinically meaningful.

-   The overall model performs poorly in predicting survival in this
    dataset.


**Schoenfeld Residuals for Age**

```{r}
# Test proportional hazards assumption for the model including age
cox_zph_age <- cox.zph(cox_age)
print(cox_zph_age)
```

Proportional hazards were tested using Schoenfeld residuals. The
assumption was met for **age** (`p = 0.195`) but marginally violated for
**treatment group** (`p = 0.055`), which is slightly below the
conventional threshold of 0.05. The **global test** for the model
(`p = 0.045`) was also just below 0.05, indicating a **possible overall
violation** of the proportional hazards assumption.

While these results do not strongly indicate a breakdown of the PH
assumption, the borderline p-values warrant caution. A visual inspection
of the residual plots (`plot(cox.zph())`) can help determine whether the
trend is substantial or random variation.


```{r}
# Visual check
plot(cox_zph_age)
```

The Schoenfeld residual plot for age, derived from the multivariable Cox
model, reveals no strong evidence of violation of the proportional
hazards assumption. The smoothed coefficient curve remains relatively
flat across time, with the 95% confidence band consistently overlapping
zero. This suggests that the effect of age on the hazard function does
not meaningfully vary with time. This visual impression aligns with the
result from the formal statistical test (χ² = 1.68, df = 1, p = 0.195),
providing no significant indication that the proportional hazards
assumption has been violated for age in this model.


**Martingale Residuals for Age to Assess Functional Form**

To assess whether age is appropriately modeled as a linear predictor in
the Cox model, we examine Martingale residuals plotted against age to
detect non-linearity in its functional form.

```{r}
# Generate Martingale residuals
veteran$martingale_age <- residuals(cox_age, type = "martingale")

# Generate linear predictor
veteran$linear_pred_age <- predict(cox_age, type = "lp")

# Plot residuals against age
ggplot(veteran, aes(x = age, y = martingale_age, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Age",
    x = "Age (years)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

Martingale residuals plotted against age showed no clear non-linear
trend, with residuals generally scattered randomly across the range of
age values. This suggests that modeling age as a linear term in the Cox
model is reasonable. There is no strong evidence to support the need for
transformations or spline-based modeling of age in this dataset.


**Functional Form Check**

```{r}
# Fit a univariable Cox model to assess functional form of age
cox_age_only <- coxph(Surv(time, status) ~ age, data = veteran)

# Check the functional form visually using Martingale residuals
ggcoxfunctional(cox_age_only, data = veteran)
```


The functional form check using Martingale residuals and a
non-parametric smoother suggests that age may not have a strictly linear
relationship with the log hazard. The smooth curve appears U-shaped,
implying that risk may not increase evenly with age across the full
range. This indicates that the assumption of linearity may be violated,
and that alternative modeling strategies (e.g., polynomial terms or
splines) could improve model fit.

While the Martingale residual plot suggests that the relationship
between age and the log hazard may not be strictly linear, age was not
statistically significant in the model. As such, no transformation or
spline was applied. This could be revisited in future modeling if age
becomes more influential.

### Binary Variable: Prior Treatment

We now evaluate the variable `prior`, which indicates whether a patient
received prior therapy before enrolment in the study. This binary
variable may capture important differences in disease severity,
treatment history, or prognosis that could influence survival outcomes.


**Fit Cox Model with Prior Treatment**

```{r}
# Fit the Cox model including prior treatment
cox_prior <- coxph(Surv(time, status) ~ trt + prior, data = veteran)
summary(cox_prior)
```

A Cox proportional hazards model was fitted to evaluate the association
between survival and two predictors: treatment group (trt) and prior
therapy (prior). The model was based on 137 patients, with 128 observed
events.

The hazard ratio for the intervention group was 1.03 (95% CI: 0.72 –
1.46, p = 0.885), indicating no evidence of a treatment effect after
adjusting for prior therapy. Similarly, prior therapy was not associated
with survival, with a hazard ratio of 0.99 (95% CI: 0.95 – 1.03, p =
0.471).

Model discrimination was weak, with a concordance of 0.517, suggesting
the model has limited ability to rank survival times accurately. Global
model fit tests (likelihood ratio, Wald, and score) were all
non-significant (p = 0.8), indicating that the included variables do not
significantly improve model performance.

Neither treatment assignment nor prior therapy appear to be meaningful
predictors of survival in this dataset. There is no evidence of an
additive effect between these covariates.

**Schoenfeld Residuals for Prior Treatment**

```{r}
# Test proportional hazards assumption for the model including prior treatment
cox_zph_prior <- cox.zph(cox_prior)
print(cox_zph_prior)
```

Proportional hazards were assessed using Schoenfeld residuals for the
model including treatment group and prior therapy. The global test was
statistically significant (p = 0.048), suggesting potential violation of
the proportional hazards assumption overall.

Individually, both treatment (p = 0.066) and prior therapy (p = 0.081)
had p-values slightly above the conventional 0.05 threshold, indicating
borderline evidence of non-proportionality. These results do not
strongly confirm a violation, but they do warrant cautious
interpretation.

Given the marginal nature of these findings, visual inspection of the
Schoenfeld residual plots is recommended to assess whether any
systematic time-varying effects are present.

```{r}
# Visual check
plot(cox_zph_prior)
```

The Schoenfeld residual plot for prior treatment shows a smoothed line
that remains approximately flat over time, with the 95% confidence band
consistently including zero. This suggests that the effect of prior
treatment on the hazard is stable across the follow-up period, and
supports the assumption of proportional hazards for this variable.

In contrast, the plot for treatment group (trt) exhibits a noticeable
wavy pattern in the smoothed line, particularly in the early to
mid-study period. While the confidence bands still broadly include zero,
the shape of the curve raises some concern about potential time-varying
effects. Combined with a borderline p-value from the cox.zph() test (p =
0.060), this suggests the proportional hazards assumption for trt is not
clearly violated, but may warrant caution and further
consideration—especially in more complex models.

**Martingale Residuals for Prior Treatment**

```{r}
# Generate Martingale residuals
veteran$martingale_prior <- residuals(cox_prior, type = "martingale")
# Generate linear predictor
veteran$linear_pred_prior <- predict(cox_prior, type = "lp")
# Plot residuals against prior treatment
ggplot(veteran, aes(x = prior, y = martingale_prior, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Prior Treatment",
    x = "Prior Treatment (0 = No, 1 = Yes)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) # Adding jitter to avoid overlap

```

The Martingale residual plot for the prior variable shows two distinct
clusters of points at 0 and 1, corresponding to patients with and
without prior treatment, respectively. The residuals appear to be
roughly symmetric around zero in both groups, with no strong indication
of skew or heteroscedasticity. This supports the appropriateness of
modeling prior as a binary variable in the Cox regression model. No
transformation or alternative coding appears necessary.


### Multivariable Cox Model with Age and Prior Treatment

In this section, we fit a multivariable Cox proportional hazards model
that includes both age and prior treatment as covariates. This allows us
to assess the independent effects of these variables on survival while
controlling for treatment group.

```{r}
# Fit the multivariable Cox model including age and prior treatment
cox_multivariable <- coxph(Surv(time, status) ~ trt + age + prior, data = veteran)
summary(cox_multivariable)
```


A multivariable Cox proportional hazards model was fitted to examine the
association between survival and three covariates: treatment group, age,
and prior therapy. After adjusting for all covariates:

-   The hazard ratio for the intervention group relative to control was
    1.00 (HR = 1.00; 95% CI: 0.70–1.44; p = 0.984), indicating no
    evidence of a treatment effect.

-   Age had a hazard ratio of 1.01 per year (p = 0.462), suggesting no
    significant association with survival in this model.

-   Prior therapy also showed no effect on hazard (HR = 0.99; p =
    0.500).

Model discrimination was minimal (concordance = 0.507), and all global
model tests were non-significant (p = 0.8), suggesting the model does
not meaningfully predict survival in this dataset.

**Schoenfeld Residuals for Multivariable Model**

We now assess whether the **proportional hazards assumption** holds for
each covariate in the multivariable Cox model. This step ensures that
the effect of each variable — treatment group, age, and prior therapy —
is consistent over time.

```{r}
# Test proportional hazards assumption for the multivariable model
cox_zph_multivariable <- cox.zph(cox_multivariable)
print(cox_zph_multivariable)
```

The proportional hazards assumption was assessed for the multivariable
Cox model using Schoenfeld residuals. The global test was statistically
significant (p = 0.032), suggesting that at least one covariate may have
a time-varying effect on the hazard.

At the individual variable level:

-   trt showed borderline evidence of non-proportionality (p = 0.060)

-   prior also approached significance (p = 0.074)

-   age did not show evidence of time-varying effects (p = 0.184)

While none of the individual tests crossed the 0.05 threshold, the
cumulative result (GLOBAL p \< 0.05) suggests some caution is warranted.
Visual inspection of the Schoenfeld residual plots is recommended to
evaluate whether these borderline findings correspond to meaningful
deviations over time.

```{r}
# Visual check
plot(cox_zph_multivariable)
```

Visual inspection of the Schoenfeld residual plots for treatment group
(trt), age, and prior treatment (prior) showed no substantial evidence
of non-proportionality. The smoothed lines for each covariate remained
relatively flat and were mostly contained within their respective 95%
confidence bands across the follow-up period.

-   For treatment group, the smoothed beta function showed some mild
    undulation, but it remained within the confidence bounds, indicating
    no major deviation from the proportional hazards assumption.

-   For age, the trend line was nearly horizontal with minimal
    fluctuation, strongly supporting time-invariant effects.

-   For prior treatment, the estimated beta also remained relatively
    stable over time, with the confidence interval consistently covering
    zero.

Together, these plots support the conclusion that the proportional
hazards assumption is reasonably satisfied for each covariate in the
multivariable model, aligning with the statistical test results where
none of the individual covariate tests were significant at the 0.05
level.

**Martingale Residuals for Multivariable Model**

We now examine the Martingale residuals from the multivariable Cox model
to evaluate the overall model fit and identify any patterns or potential
outliers in the covariate effects.

```{r}
# Generate Martingale residuals
veteran$martingale_multivariable <- residuals(cox_multivariable, type = "martingale")
# Generate linear predictor
veteran$linear_pred_multivariable <- predict(cox_multivariable, type = "lp")
# Plot residuals against linear predictor
ggplot(veteran, aes(x = linear_pred_multivariable, y = martingale_multivariable, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Linear Predictor\n(Multivariable Model)",
    x = "Linear Predictor (Log Hazard)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The Martingale residuals plotted against the linear predictor for the
multivariable model appear randomly scattered around the horizontal zero
line, with no clear pattern, curvature, or systematic deviation. This
suggests that the overall functional form of the covariates (treatment,
age, and prior treatment) is reasonably specified. There is no strong
evidence of model misspecification or non-linearity in the combined
effects. However, the narrow spread of the linear predictor values
(mostly between -0.2 and 0.2) reflects the modest effect sizes and
limited variation in the log hazard estimates, which aligns with the
model's lack of significant predictors.


**Functional Form Check for Continuous Covariates**

We now recheck the functional form of the continuous variable age within
the multivariable Cox model to ensure its effect on the log hazard
remains appropriately modeled as linear.

```{r}
# Refit a univariable model with age (again, for functional form check)
cox_age_for_form_check <- coxph(Surv(time, status) ~ age, data = veteran)

# Plot functional form using ggcoxfunctional
ggcoxfunctional(
  fit = cox_age_for_form_check,
  data = veteran,
  point.col = "#E74C3C",
  ggtheme = theme_minimal(base_size = 13),
  title = "Functional Form Check: Age"
)
```

The Martingale residual plot for age reveals a clear U-shaped curve,
suggesting that the assumption of linearity in the log hazard may not
hold. This implies that age may have a non-linear effect on survival,
and modeling it as a simple linear term may lead to misestimation of its
effect. A transformation (e.g., quadratic or restricted cubic splines)
may be warranted in further analysis.


### Multivariable Cox Model with Spline-Transformed Age

We refit the Cox model using a restricted cubic spline to model the
non-linear relationship between age and the log hazard. This allows for
greater flexibility in capturing age-related risk while still accounting
for treatment and prior therapy.


**Model Output**

```{r}
# Fit a Cox model with restricted cubic splines for age
cox_spline <- coxph(Surv(time, status) ~ ns(age, df = 3) + trt + prior, data = veteran)
summary(cox_spline)
```

A multivariable Cox proportional hazards model incorporating a
restricted cubic spline transformation of age (df = 3) was fitted to
explore potential non-linear effects of age on survival. None of the
individual spline terms were statistically significant at the 0.05 level
(p = 0.906, 0.143, and 0.157), suggesting that while the spline captures
flexibility, the overall contribution of age remains statistically weak.

The treatment group (p = 0.626) and prior therapy (p = 0.640) also
remained non-significant. Model-level tests showed that the likelihood
ratio test (p = 0.1), Wald test (p = 0.1), and score test (p = 0.1) were
all non-significant, indicating that the model, as specified, may not
provide strong explanatory power.

However, the model's concordance improved slightly to 0.559 compared to
the linear age model, suggesting a modest increase in predictive
ability. While not conclusive, the use of splines helps satisfy the
functional form assumption and could be valuable if future models
identify stronger non-linear trends or interactions with age.

**Schoenfeld Residuals for Spline Model**

```{r}
# Test proportional hazards assumption for the spline model
cox.zph(cox_spline)
```

A Cox proportional hazards model incorporating a restricted cubic spline
transformation of age (df = 3) demonstrated acceptable fit to the
proportional hazards assumption across all covariates. The global
Schoenfeld residuals test was non-significant (p = 0.116), suggesting
that the model adequately satisfies the proportional hazards assumption.
Individually, spline-transformed age (p = 0.211), treatment group (p =
0.114), and prior treatment (p = 0.079) all showed no significant
time-dependent effects.

```{r}
# Visual check of Schoenfeld residuals for spline model
plot(cox.zph(cox_spline))
```

The Schoenfeld residuals were examined for the final multivariable Cox
model, which included a spline-transformed age term. Across all
covariates—age (using splines), treatment group (trt), and prior
treatment (prior)—the smoothed residual lines remained relatively flat
and within the 95% confidence intervals, suggesting no strong evidence
of time-varying effects.

-   Age (spline-transformed): Despite some visual variability, the three
    spline basis functions for age each produce smoothed lines that
    remain stable across time. Importantly, the confidence intervals
    broadly contain zero, indicating that the assumption of
    proportionality is not clearly violated despite the model’s
    non-linear specification.

-   Treatment (trt): The smoothed residual curve shows a mild wave-like
    pattern, but this fluctuation stays well within the confidence bands
    and does not indicate a strong trend over time. While a borderline
    global p-value was observed in earlier models, the residuals here
    provide no compelling evidence against the proportional hazards
    assumption for treatment.

-   Prior Treatment (prior): The smoothed residual line is relatively
    flat, and although the lower bound of the 95% confidence interval
    occasionally hovers just below zero, this does not strongly suggest
    non-proportionality. The overall trend appears stable, and the
    residuals provide visual reassurance that prior treatment's effect
    on hazard is consistent over time.

**Martingale Residuals for Spline Model**

```{r}
# Residuals vs linear predictor
veteran$martingale_spline <- residuals(cox_spline, type = "martingale")
veteran$linear_pred_spline <- predict(cox_spline, type = "lp")

ggplot(veteran, aes(x = linear_pred_spline, y = martingale_spline, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Linear Predictor\n(Spline Model)",
    x = "Linear Predictor (Log Hazard)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The plot displays Martingale residuals against the linear predictor (log
hazard) for the spline-transformed Cox model. The residuals appear
randomly scattered around the horizontal zero line, without a
discernible systematic pattern or curvature. This suggests that the
model has adequately captured the relationship between the covariates
and the outcome.

The use of restricted cubic splines for age appears to have addressed
any previously observed non-linearity, as there is no strong evidence of
model misfit. The distribution of residuals across treatment groups also
seems relatively balanced, with no obvious separation or
heteroscedasticity.

The model’s functional form is appropriate, and the spline
transformation has improved the model’s fit by flexibly accommodating
the non-linear effect of age.

### Model Comparison: Linear vs Spline Age Term

We now compare the original multivariable Cox model with age modeled
linearly to the updated model using a spline transformation for age.
This allows us to formally test whether the added flexibility of the
spline significantly improves model fit.

```{r}
# Compare linear age model vs spline-transformed age model
anova(cox_multivariable, cox_spline, test = "LRT")
```

A likelihood ratio test comparing a model with age modeled linearly to
one using restricted cubic splines (df = 3) showed a statistically
significant improvement in fit (χ² = 7.91, df = 2, p = 0.019). This
suggests that the effect of age on survival is non-linear, and that
modeling age with splines better captures its relationship with hazard.

**Visualizing the Effect of Spline-Transformed Age**

This section uses 'termplot()' to visualize how the hazard changes
across age when modeled as a spline. This provides an intuitive look at
the non-linear relationship.


```{r}
termplot(cox_spline, se = TRUE, col.term = "blue")
```

The term plots from the spline-based Cox proportional hazards model
illustrate the partial effect of each covariate on the log hazard:

-   Age (modeled using natural splines with 3 degrees of freedom)
    exhibits a non-linear association with the log hazard. The curve
    displays a shallow U-shape, with the lowest hazard around age 55,
    and a gradual increase in hazard at both younger and older ages.
    While the trend is suggestive, the wide confidence
    intervals—particularly at the distribution tails—indicate
    uncertainty in the precise shape of this relationship.

-   Treatment Group (trt) is represented as a binary variable with two
    flat lines, one for each group. The narrow confidence intervals
    overlap with zero, implying no statistically significant difference
    in log hazard between the intervention and control arms in this
    model.

-   Prior Treatment (prior) also shows a flat trend across the range of
    prior treatments, with confidence bands consistently containing
    zero. This suggests that the number of prior therapies does not
    contribute meaningfully to hazard estimation when adjusting for age
    and treatment.

These visualizations support the statistical findings from the model
summary. Age is the only covariate showing a potentially complex,
non-linear effect, while treatment and prior exposure show no clear
independent impact on survival in this cohort.


### Multivariable Cox Model with Karnofsky Score and Cell Type

Having established that spline-transformed age improves model fit, we now
extend the model to include Karnofsky performance score and tumour cell type
— two clinically important variables that were not included in earlier models.
Karnofsky score is a widely used measure of functional status and is one of
the strongest prognostic factors in oncology. Cell type captures the
histological character of each patient's tumour, which is known to influence
prognosis in lung cancer independently of treatment.

```{r}
# Fit a multivariable Cox model including Karnofsky score and Cell Type
cox_karno_and_cell <- coxph(
  Surv(time, status) ~ ns(age, df = 3) + trt + prior + karno + celltype,
  data = veteran)

summary(cox_karno_and_cell)
```

A multivariable Cox proportional hazards model was fitted including
spline-transformed age (df = 3), treatment group, prior therapy, Karnofsky
performance score, and tumour cell type. The model included 137 patients and
128 events.

-   **Karnofsky score** was the strongest independent predictor of survival.
    Each 1-point increase in score was associated with a 3.4% reduction in
    the hazard of death (HR = 0.966, 95% CI: 0.955–0.977, p < 0.001).

-   **Cell type** was significantly associated with survival. Compared to
    squamous cell carcinoma (reference), patients with small cell carcinoma
    had a 2.2-fold increased hazard (HR = 2.21, 95% CI: 1.31–3.75, p = 0.003)
    and adenocarcinoma patients had a 3.3-fold increased hazard
    (HR = 3.30, 95% CI: 1.82–5.96, p < 0.001). Large cell carcinoma was not
    significantly different from squamous (HR = 1.40, p = 0.233).

-   **Treatment group** was not statistically significant after adjusting for
    functional status and cell type (HR = 1.40, 95% CI: 0.94–2.09, p = 0.097),
    though the hazard ratio trended toward harm in the experimental arm.

-   **Prior therapy** was not associated with survival (HR = 1.01, p = 0.658).

-   **Spline-transformed age** showed evidence of a non-linear association with
    survival. The second spline term reached statistical significance
    (HR = 0.098, p = 0.025), consistent with the earlier functional form
    analysis.

Overall model performance was substantially improved: concordance = 0.735,
compared to 0.559 for the model without Karnofsky and cell type. The likelihood
ratio, Wald, and score tests were all highly significant (p < 0.001).

**Forest Plot**

```{r}
library(forestmodel)
forest_model(cox_karno_and_cell)
```

**Survival by Karnofsky Score**

```{r}
veteran$karno_group <- cut(veteran$karno,
                           breaks = c(0, 50, 70, 100),
                           labels = c("Low (≤50)", "Medium (51–70)", "High (>70)"),
                           right = TRUE)

km_karno <- survfit(Surv(time, status) ~ karno_group, data = veteran)

ggsurvplot(km_karno,
           data = veteran,
           risk.table = TRUE,
           pval = TRUE,
           conf.int = FALSE,
           legend.title = "Karnofsky Group",
           legend.labs = c("Low (≤50)", "Medium (51–70)", "High (>70)"),
           palette = c("firebrick", "steelblue", "darkgreen"),
           xlab = "Time (days)",
           ylab = "Survival Probability",
           title = "Kaplan-Meier Survival by Karnofsky Score Group")
```

The Kaplan-Meier curves stratified by Karnofsky score group confirm the strong
prognostic gradient seen in the Cox model. Patients with high functional status
(Karnofsky > 70) had substantially longer median survival than those in the low
group (Karnofsky ≤ 50), with the log-rank test confirming a highly significant
difference across groups.

**Survival by Cell Type**

```{r}
km_cell <- survfit(Surv(time, status) ~ celltype, data = veteran)

ggsurvplot(km_cell,
           data = veteran,
           risk.table = TRUE,
           pval = TRUE,
           conf.int = FALSE,
           legend.title = "Cell Type",
           legend.labs = levels(factor(veteran$celltype)),
           palette = "Dark2",
           xlab = "Time (days)",
           ylab = "Survival Probability",
           title = "Kaplan-Meier Survival by Cell Type")
```

Survival curves by cell type show clear separation, particularly between
squamous cell carcinoma and the adenocarcinoma and small cell groups. The
log-rank test confirms a significant difference across histological subtypes,
consistent with the hazard ratios estimated in the Cox model.

**Proportional Hazards Assumption**

```{r}
cox_zph_final <- cox.zph(cox_karno_and_cell)
print(cox_zph_final)
```

```{r}
plot(cox_zph_final)
```

The proportional hazards assumption was assessed via Schoenfeld residuals.
None of the individual covariates reached statistical significance, and the
global test was non-significant, providing no compelling evidence against the
proportional hazards assumption for any term in the final model. Visual
inspection of the Schoenfeld residual plots showed broadly flat smoothed trends
with no systematic drift over time.

**Martingale Residuals**

```{r}
veteran$martingale_final <- residuals(cox_karno_and_cell, type = "martingale")
veteran$linear_pred_final <- predict(cox_karno_and_cell, type = "lp")

ggplot(veteran, aes(x = linear_pred_final, y = martingale_final, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Linear Predictor (Final Model)",
    x = "Linear Predictor (Log Hazard)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

Martingale residuals plotted against the linear predictor show random scatter
around zero with no systematic curvature, supporting adequate overall model
specification. The linear predictor now spans a wider range than the earlier
model, reflecting the stronger prognostic signal contributed by Karnofsky score
and cell type. No major outliers or influential observations are apparent.

**Functional Form Check: Karnofsky Score**

```{r}
cox_karno_for_form_check <- coxph(Surv(time, status) ~ karno, data = veteran)

ggcoxfunctional(
  fit = cox_karno_for_form_check,
  data = veteran,
  point.col = "#E74C3C",
  ggtheme = theme_minimal(base_size = 13),
  title = "Functional Form Check: Karnofsky Score"
)
```

The Martingale residual plot for Karnofsky score shows a broadly linear
decreasing trend, supporting its inclusion as a continuous linear predictor.
No threshold or non-monotonic pattern is apparent that would require a spline
transformation, in contrast to age.

# Interaction or Additive Effects?

In Cox regression, interaction terms allow us to test whether the effect
of one covariate on the hazard changes depending on the level of another
covariate. If significant, this suggests a departure from simple
additivity and may require stratified or more complex modelling.


## Interaction of Treatment and Age

```{r}
# Fit a cox model with interaction between treatment and age 
cox_interaction <- coxph(Surv(time, status) ~ trt * ns(age, df = 3) + prior, data = veteran)
summary(cox_interaction)
```

We tested for interaction between treatment group and age (modeled using
natural splines with 3 degrees of freedom) to explore whether the effect
of treatment on survival varied across age groups. The interaction model
revealed a statistically significant interaction for the second spline
basis term (p = 0.037), suggesting a possible age-dependent treatment
effect.

The overall Wald test for the model approached significance (p = 0.05),
while the likelihood ratio test was marginal (p = 0.07), indicating that
the model with interaction terms may provide a better fit than the
additive model. These findings suggest that the benefit (or lack
thereof) of treatment may differ depending on patient age, warranting
further exploration or stratified modelling.

**Schoenfeld Residuals for Interaction Model**

```{r}
# Test proportional hazards assumption for the interaction model
cox.zph(cox_interaction)
```

The Schoenfeld residual test was used to assess whether the proportional
hazards assumption holds for each covariate and interaction term in the
model.

-   trt: χ² = 1.39, p = 0.24 → No evidence of time-varying effects for
    treatment group.

-   ns(age, df = 3): χ² = 3.90, p = 0.27 → The spline-transformed age
    term shows no strong time dependency.

-   prior: χ² = 2.42, p = 0.12 → Also not statistically significant.

-   trt × ns(age, df = 3): χ² = 5.37, p = 0.15 → The interaction between
    age and treatment shows no significant violation of proportionality
    either.

-   GLOBAL test: χ² = 8.94 on 8 df, p = 0.35 → The model as a whole
    satisfies the PH assumption.

There is no strong evidence that the proportional hazards assumption is
violated in this interaction model. The covariate effects, including the
interaction, appear stable over time.

```{r}
# Visual check of Schoenfeld residuals for interaction model
plot(cox.zph(cox_interaction))
```

The global test using cox.zph() indicated no significant violation of
the PH assumption (χ² = 8.94, df = 8, p = 0.35). Individual covariate
tests also returned non-significant p-values, suggesting that the
estimated hazard ratios remain approximately constant over time.

Visual inspection of the Schoenfeld residual plots supported this
conclusion. The smoothed curves for treatment, age (spline terms), prior
treatment, and their interaction remained relatively flat and within
their 95% confidence bands. While a few outliers were
present—particularly in the treatment and age components—these did not
indicate strong evidence of non-proportionality.

Taken together, these results suggest that the proportional hazards
assumption is reasonably met for all terms in the interaction model.

**Martingale Residuals for Interaction Model**

The Martingale residuals from the interaction model were plotted against
the linear predictor (log hazard) to assess model fit and identify any
potential patterns or outliers.

```{r}
# Residuals vs linear predictor
veteran$martingale_interaction <- residuals(cox_interaction, type = "martingale")
veteran$linear_pred_interaction <- predict(cox_interaction, type = "lp")
ggplot(veteran, aes(x = linear_pred_interaction, y = martingale_interaction, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Linear Predictor\n(Interaction Model)",
    x = "Linear Predictor (Log Hazard)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The Martingale residuals plot from the interaction model reveals a
concentration of predicted log hazard values for the control group
between 0.5 and 1.0, whereas predictions for the intervention group span
a wider range. This may reflect the influence of the age-by-treatment
interaction term, which adds complexity primarily for patients receiving
the intervention. The residuals themselves are randomly scattered around
zero, suggesting that the functional form of the covariates is
appropriately specified and that no major nonlinearity remains
unaccounted for.

**Deviance Residuals for Interaction Model**

Deviance residuals were generated to further assess the model fit and
identify any potential outliers or influential observations.


```{r}
# Deviance residuals
veteran$deviance_interaction <- residuals(cox_interaction, type = "deviance")
# Plot deviance residuals
ggplot(veteran, aes(x = linear_pred_interaction, y = deviance_interaction, color = trt)) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) + # Adding jitter to avoid overlap
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Deviance Residuals vs Linear Predictor\n(Interaction Model)",
    x = "Linear Predictor (Log Hazard)",
    y = "Deviance Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The deviance residual plot provides a diagnostic check on the adequacy
of the Cox proportional hazards model that includes an interaction
between treatment and age. In this plot, residuals are symmetrically
distributed around zero with no discernible curvature or trend,
suggesting that the model fits the data reasonably well.

Notably, the intervention group shows a broader range of linear
predictor values (i.e., estimated log hazard), indicating greater
variability in predicted risk compared to the control group, which
clusters tightly between 0.5 and 1.0. Despite this asymmetry in
predictor spread, the vertical dispersion of residuals remains balanced,
and no strong outliers or patterns emerge. This supports the
appropriateness of the interaction model and suggests no serious
violations of model assumptions related to functional form or
influential observations.

Overall, the deviance residuals confirm that the model adequately
captures the observed survival data, with no clear evidence of misfit or
heteroscedasticity.

## Interaction of Treatment and Prior Therapy

Prior treatments might alter a patient’s response to new therapies
(e.g., due to resistance, immune suppression, or cumulative toxicity).

```{r}
# Fit a cox model with interaction between treatment and prior therapy
cox_interaction_prior <- coxph(Surv(time, status) ~ trt * prior + ns(age, df = 3), data = veteran)
summary(cox_interaction_prior)
```


A Cox proportional hazards model including an interaction between treatment and prior therapy was fitted to assess potential effect modification. The interaction term (trt × prior) yielded a hazard ratio of 0.94 (95% CI: 0.87–1.02, p = 0.135), suggesting that the effect of treatment may be modestly attenuated in patients who had received prior therapy. However, this association did not reach statistical significance. Model-level tests (Likelihood Ratio, Wald, and Score) were borderline (p = 0.07–0.08), indicating weak evidence that the inclusion of an interaction term may improve model fit. While these findings are not conclusive in this sample, they raise the possibility of effect modification that could be explored further in larger, more adequately powered studies.

**Schoenfeld Residuals for Interaction with Prior Therapy**

```{r}
# Test proportional hazards assumption for the interaction with prior therapy
cox.zph(cox_interaction_prior)
```

The proportional hazards assumption was assessed using Schoenfeld residuals. The global test did not indicate violation (p = 0.171), supporting the use of the Cox model. Most individual terms showed no evidence of non-proportionality, though the interaction between treatment and prior therapy (trt:prior) was borderline (p = 0.054). This may suggest some time-dependent behavior in how prior treatment modifies the treatment effect, though the evidence is not conclusive.

```{r}
# Visual check of Schoenfeld residuals for interaction with prior therapy
plot(cox.zph(cox_interaction_prior))
```

To assess whether the proportional hazards (PH) assumption holds in the model including an interaction between treatment and prior therapy, Schoenfeld residuals were examined both statistically and graphically.

The global test for the PH assumption was not significant (p = 0.17), and all individual covariates—including the interaction term—also showed non-significant test results (all p > 0.05), suggesting no strong evidence of time-varying effects.

Visual inspection of the residual plots supports this conclusion. For all variables, including trt, prior, age (splines), and the interaction term trt × prior, the smoothed lines remain relatively flat over time, and their 95% confidence bands generally include zero. While the plots for trt and prior show some scatter at the extremes—with several residuals appearing as outliers above and below the bands—these deviations do not appear systematic or directional.

In particular, the clusters of points outside the confidence bounds likely reflect the influence of a small number of high-leverage observations rather than a violation of the PH assumption. These artifacts are common in small-to-moderate datasets, especially when stratifying by interactions.

Conclusion: There is no strong statistical or visual evidence to reject the PH assumption for this interaction model. However, the presence of some outlier residuals may warrant further sensitivity analysis in larger samples.

**Martingale Residuals for Interaction with Prior Therapy**

```{r}
# Residuals vs linear predictor
veteran$martingale_interaction_prior <- residuals(cox_interaction_prior, type = "martingale")
veteran$linear_pred_interaction_prior <- predict(cox_interaction_prior, type = "lp")

ggplot(veteran, aes(x = linear_pred_interaction_prior, y = martingale_interaction_prior, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Linear Predictor\n(Interaction with Prior Therapy)",
    x = "Linear Predictor (Log Hazard)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The martingale residual plot shows the residuals plotted against the linear predictor (log hazard), grouped by treatment arm. In this context, the martingale residuals help assess the model's functional form — particularly whether the covariates are correctly specified (e.g., linearity assumptions, missing covariates, or poorly modeled interactions).
Key Observations:

_Vertical Spread (Residuals Axis):_
There is substantial vertical scatter in the residuals, with several points deviating far from the horizontal line at 0. This is expected in survival models, particularly with smaller datasets. However, there is no clear systematic pattern, such as curvature or directional trend, that would suggest strong non-linearity or misspecification.

_Horizontal Clustering (Linear Predictor Axis):_
As with previous models, a dense cluster of observations is seen between approximately 0 and 1 on the linear predictor scale, mostly corresponding to the control group. This may reflect limited variation in the covariates for this subgroup, or overlapping risk profiles. The intervention group displays more dispersion, which aligns with the interaction term — suggesting greater heterogeneity in risk when prior treatment is included in the model.

_Overall Interpretation:_
The lack of a strong non-random pattern supports the adequacy of the covariate specification, including the interaction term. However, the wide spread and presence of extreme residuals imply potential outliers or influential cases that might warrant further sensitivity analysis in larger samples.

**Deviance Residuals for Interaction with Prior Therapy**

```{r}
# Deviance residuals
veteran$deviance_interaction_prior <- residuals(cox_interaction_prior, type = "deviance")

# Plot deviance residuals
ggplot(veteran, aes(x = linear_pred_interaction_prior, y = deviance_interaction_prior, color = trt)) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) + # Adding jitter to avoid overlap
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Deviance Residuals vs Linear Predictor\n(Interaction with Prior Therapy)",
    x = "Linear Predictor (Log Hazard)",
    y = "Deviance Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The plot of deviance residuals against the linear predictor for the interaction model reveals no strong patterns or obvious deviations from randomness. Residuals are symmetrically scattered around the zero line, with similar vertical spread across the range of the linear predictor. This suggests that:

_Model Fit Appears Reasonable:_
There is no clear evidence of systematic bias, such as curvature, funneling, or clustering that might imply model misspecification, omitted variables, or heteroscedasticity.

_Symmetry Around Zero:_
The relatively balanced distribution of residuals above and below the zero line supports the assumption that the model is neither consistently under- nor over-estimating the risk.

_No Extreme Outliers:_
While a few points have larger residuals, there are no values that fall far outside the typical range (±3), indicating a lack of highly influential or poorly fit observations.

_Group Comparison:_
The intervention and control groups appear similarly distributed across the range of linear predictors, lending further support to the appropriateness of the model specification including the interaction term.

_Summary:_

The deviance residual plot offers no strong evidence of model misfit and supports the adequacy of the Cox model including the interaction between treatment and prior therapy. Residuals appear randomly and symmetrically distributed, suggesting that the model captures the underlying structure of the data well, at least within the limitations of this dataset.

# Model Comparison: Interaction vs Non-Interaction

Two extended models were fit to explore potential effect modification: one
including an interaction between treatment and age (via natural splines), and
another between treatment and prior therapy. While the treatment × age model
showed a statistically significant interaction term, the overall model-level
tests remained marginal (Wald p = 0.05, Score p = 0.03), with no compelling
evidence of meaningful improvement in predictive performance. The treatment ×
prior therapy interaction suggested a potential moderating effect, but this did
not reach statistical significance (p = 0.135), and proportional hazards
assumptions were borderline (p = 0.054).

When comparing these interaction models to the full additive model
(trt + prior + ns(age, df=3) + karno + celltype), neither interaction offered
substantial improvement in model fit or interpretability. Residual diagnostics
also did not reveal meaningful deviations. The additive model was therefore
selected as the final model.

# Final Model Selection and Interpretation

After evaluating a sequence of Cox proportional hazards models — from
univariable models for each covariate individually, through multivariable
models of increasing complexity, to interaction models — the final selected
model is:

$$\text{trt} + \text{prior} + ns(\text{age}, df=3) + \text{karno} + \text{celltype}$$

**Rationale for Model Selection**

-   **Karnofsky score and cell type** were included on clinical grounds.
    Karnofsky performance score is the dominant prognostic factor in oncology;
    cell type defines biologically distinct lung cancer subtypes with very
    different natural histories. Both were confirmed as strong independent
    predictors in the multivariable model.

-   **Age** was retained and modeled using a natural cubic spline (df = 3)
    following a functional form check that revealed a non-linear relationship
    with the log hazard. A likelihood ratio test confirmed the spline
    specification provided a significantly better fit than a linear age term
    (χ² = 7.91, df = 2, p = 0.019).

-   **Prior therapy** was retained as a standard adjustment covariate in the
    RCT context. It was not statistically significant (HR = 1.01, p = 0.658)
    but its inclusion is clinically justified and does not meaningfully alter
    other estimates.

-   **Interaction terms** (treatment × age, treatment × prior therapy) were
    explored but neither provided a meaningful or significant improvement in
    model fit. Neither was retained in the final model.

-   **`diagtime`** was not included. It was non-significant and its
    interpretation as a predictor is clinically ambiguous in this context.

**Interpretation of the Final Model**

-   **Treatment group:** The experimental chemotherapy regimen was not
    associated with improved survival compared to standard treatment after
    adjusting for all other covariates (HR = 1.40, 95% CI: 0.94–2.09,
    p = 0.097). This null finding is consistent with the log-rank test
    (p = 0.928).

-   **Karnofsky score:** Each 1-point increase in performance status was
    associated with a 3.4% reduction in the hazard of death (HR = 0.966,
    95% CI: 0.955–0.977, p < 0.001). This is the strongest predictor in the
    model by a considerable margin.

-   **Cell type:** Compared to squamous cell carcinoma, patients with small
    cell carcinoma faced a 2.2-fold increased hazard (HR = 2.21, p = 0.003)
    and adenocarcinoma patients faced a 3.3-fold increased hazard
    (HR = 3.30, p < 0.001). Large cell carcinoma was not significantly
    different from squamous (HR = 1.40, p = 0.233).

-   **Age:** The spline-transformed age term showed a non-linear association
    with survival, with the second spline term reaching significance
    (HR = 0.098, p = 0.025). The overall age effect was modest.

-   **Prior therapy:** Not a significant predictor in the adjusted model
    (HR = 1.01, p = 0.658).

**Overall Model Performance**

The final model achieved a concordance index of 0.735, a substantial
improvement over the earlier model using only treatment, age, and prior therapy
(concordance ≈ 0.55). The likelihood ratio, Wald, and score tests were all
highly significant (p < 0.001). Proportional hazards diagnostics showed no
evidence of assumption violations.

# Conclusion

This analysis examined survival outcomes among 137 male veterans with advanced,
inoperable lung cancer enrolled in a randomised controlled trial comparing
standard chemotherapy to an experimental regimen. A series of Cox proportional
hazards models were fitted, with progressive inclusion of clinical covariates,
functional form testing, and diagnostic checking.

The primary finding is that the experimental treatment did not improve survival
relative to standard chemotherapy (HR = 1.40, 95% CI: 0.94–2.09, p = 0.097),
consistent with the unadjusted log-rank test (p = 0.928). This null result is
robust across model specifications and is not an artefact of inadequate
adjustment.

The two dominant prognostic factors in this cohort were Karnofsky performance
score and tumour cell type. A 10-point increase in Karnofsky score was
associated with a roughly 29% reduction in the hazard of death; adenocarcinoma
and small cell carcinoma conferred substantially higher hazard than squamous
cell carcinoma. These findings align with decades of oncological literature and
underscore the primacy of functional status and histology in lung cancer
prognosis.

Age, modeled with a natural cubic spline, showed some evidence of a
non-monotonic association with survival. Prior therapy was not independently
associated with survival in the adjusted model.

The final model demonstrated good discrimination (concordance = 0.735),
satisfied proportional hazards assumptions, and showed no major evidence of
misspecification in residual diagnostics.

# Limitations

**Sample size and generalisability.** The dataset comprises 137 patients from
a single trial, which constrains statistical power for detecting modest
treatment effects and limits the precision of subgroup estimates. The small
cell carcinoma and adenocarcinoma groups in particular have relatively few
patients, making their hazard ratio estimates less stable.

**All-male cohort.** All participants were male veterans. Lung cancer biology,
treatment response, and prognosis differ between sexes, so findings cannot be
assumed to generalise to female patients.

**Historical data.** The trial was conducted in the 1970s–80s. Neither the
experimental treatment nor the standard regimen reflects current clinical
practice, and the trial predates targeted therapy, immunotherapy, and modern
platinum-based regimens.

**Event rate and censoring.** With 128 deaths among 137 patients (93.4%),
this dataset has very limited censoring. Survival estimates at longer time
points are based on very few at-risk individuals.

**No external validation.** Model performance (concordance = 0.735) was
assessed on the same data used to fit the model. Without an independent
validation cohort, this estimate is likely optimistic.

**Residual confounding.** Unmeasured factors — such as comorbidities, disease
stage, smoking history, or treatment compliance — may confound the observed
associations.

**Interaction exploration was limited.** Only two interaction terms were
tested. Clinically meaningful interactions between other covariates (e.g.,
treatment × cell type) were not explored.
