---
title: "Survival Analysis of Veterans with Advanced Lung Cancer: A Cox Proportional Hazards Approach"
author: "Julian Chung"
format:
  html:
    toc: true
    code-fold: true
    number-sections: true
    theme: cosmo
execute:
  echo: true
  warning: false
  message: false
---
# Introduction

This report presents a survival analysis of patients with advanced, inoperable lung cancer who participated in a randomized clinical trial conducted by the Veterans Administration in the United States. The objective of the study was to compare the effectiveness of a standard chemotherapy protocol versus a test (experimental) treatment in prolonging patient survival.

Participants were followed until death or censoring, with overall survival time (in days from treatment initiation) as the primary outcome. This dataset, originally introduced in The Statistical Analysis of Failure Time Data by Kalbfleisch and Prentice (1980), has become a cornerstone in teaching survival analysis due to its manageable size, clear structure, and strong clinical relevance.

Our analysis explores survival outcomes using both Kaplan-Meier estimation and Cox proportional hazards modeling, followed by diagnostic checks to assess model assumptions and potential improvements.
The dataset contains information on:

# Setup and Data Import

We begin by loading the necessary R packages and importing the veteran lung cancer dataset from the `survival` package.

```{r}
# Load required libraries
library(survival) # Core survival analysis tools
library(survminer)  # Visualization and diagnostics
library(dplyr)  # Data wrangling
library(ggplot2)  # Plotting
library(broom)  # Tidying model output
library(splines) # Splines for flexible modeling

```

::: {.callout-tip title="Why These Packages?" collapse=true}

- **`survival`**: Core package for survival analysis in R. Provides `Surv()` to construct survival objects and `coxph()` for Cox models.

- **`survminer`**: High-level visualization tools for survival curves and diagnostics (e.g., `ggsurvplot()`, `ggcoxdiagnostics()`).

- **`splines`**: Adds support for modeling non-linear relationships in covariates using **restricted cubic splines** (e.g., `ns()`).

- **`ggplot2`**: Used to build layered, customizable plots — often integrated with `survminer`.

- **`dplyr`**: Enables clean and readable data manipulation (e.g., `mutate()`, `filter()`).

- **`broom`**: Tidies model outputs into data frames (useful for summaries, tables, and downstream analysis).
:::

Next, the dataset is loaded and a quick transformation applied to label the treatment and control groups.

```{r}
# Load the dataset
veteran <- survival::veteran %>%
  mutate(trt = factor(trt,
                      levels = c(1, 2),
                      labels = c("Control", "Intervention")))

# Peek at the structure and summary
# glimpse(veteran)
# summary(veteran)
```

::: {.callout-note title="Dataset Preview" collapse=true}
We show the first few rows to get a sense of the data structure.
```{r}
head(veteran)
```
:::

**Variable Summary:**

Below is a summary of the key variable in the dataset.

| Variable   | Type    | Description                                                        | Notes                                                            |
| ---------- | ------- | ------------------------------------------------------------------ | ---------------------------------------------------------------- |
| `trt`      | Factor  | Treatment group: 1 = Standard, 2 = Test                            | Relabeled to `"Control"` and `"Intervention"`                    |
| `celltype` | Factor  | Histologic type of lung cancer: squamous, small cell, adeno, large | Not used in this model, but useful in extended analyses          |
| `time`     | Numeric | Survival time in days                                              | Outcome variable in the `Surv()` object                          |
| `status`   | Numeric | Event indicator: 1 = died, 0 = censored                            | Used to define survival endpoint                                 |
| `karno`    | Numeric | Karnofsky performance score (proxy for baseline health status)     | May affect prognosis; not modeled in current analysis            |
| `diagtime` | Numeric | Time from diagnosis to study entry (in months)                     | Possibly useful in extended time-dependent models                |
| `age`      | Numeric | Patient age (in years)                                             | Treated as **continuous**; **checked for linearity and splines** |
| `prior`    | Numeric | Number of prior therapies (0 or 10)                                | Binary variable: 0 = none, 10 = prior treatment                  |

# Exploratory Survival Analysis

## Kaplan-Meier Survival Curve

Firstly we estimate the **survival function** for patients using the Kaplan-Meier estimator. This provides a non-parametric view of survival over time, stratified by treatment group.

```{r}
# Create the survival object
surv_obj <- Surv(time = veteran$time, event = veteran$status)

# Fit Kaplan-Meier curve stratified by treatment
km_fit <- survfit(surv_obj ~ trt, data = veteran)

# View basic survival summary
# summary(km_fit)

```

::: {.callout-note title="Survival Summary Truncated for Clarity" collapse=true}
The full Kaplan-Meier summary includes hundreds of time points. For readability, we display only the first few rows per treatment group.
```{r}
# View basic survival summary
summary(km_fit)
```
:::

::: {.callout-note title="Why Start with a Kaplan-Meier Curve?" collapse=true}

The **Kaplan-Meier (KM) curve** is a fundamental starting point in survival analysis because it allows us to:

- **Visualize survival patterns** over time,
- Compare groups (e.g., treatment vs control) using log-rank tests,
- And understand the *overall shape* and drop-off of survival before making model-based assumptions.

The KM method is **non-parametric**, which means it does **not assume any specific distribution** for survival times. It simply calculates the probability of surviving past a given time point based on the observed data.

This makes it ideal as a **descriptive tool** and as a way to **check for potential group differences** *before* moving on to more complex models.

By contrast, the **Cox proportional hazards model** is a **semi-parametric** model. It assumes:
- A baseline hazard function (unspecified),
- And that covariates have a *multiplicative effect* on the hazard over time.

This adds **power and flexibility**, but also introduces assumptions — like the **proportional hazards** assumption — that must be checked.

✅ So we start with KM to understand the data and get a rough idea of group differences — without assuming too much — before fitting a Cox model.
:::


```{r}
# Plot the KM curve
ggsurvplot(
  km_fit,
  data = veteran,
  conf.int = TRUE,
  pval = TRUE,
  pval.size = 4,
  risk.table = TRUE,
  risk.table.title = "Number at Risk",
  risk.table.col = "strata",
  risk.table.y.text.col = TRUE,
  risk.table.fontsize = 4,
  legend.title = "Treatment Group",
  legend.labs = c("Control", "Intervention"),
  xlab = "Time (days)",
  ylab = "Survival Probability",
  palette = c("#A569BD", "#45B39D"),
  title = "Kaplan-Meier Survival Curve: VA Lung Cancer Trial",
  ggtheme = theme_minimal(base_size = 13) + 
    theme(
      legend.position = "top",
      plot.title = element_text(hjust = 0.5, face = "bold"),
      panel.grid.major = element_line(color = "grey80", linewidth = 0.4),
      panel.grid.minor = element_line(color = "grey90", linewidth = 0.2)
    ),
  tables.theme = theme(  # Trying to tweak the risk table
    axis.text.y = element_text(size = 10, margin = margin(r = 12), lineheight = 2)
  )
)
```

**Interpretation**

The Kaplan-Meier plot shows the survival curves for patients in the intervention and control arms. For both groups there is a steep decline in survival probability during the first 200 days, indicating rapid disease progression. The shaded areas represent the 95% confidence intervals, which largely overlap throughout the follow-up period, suggesting no significant difference in survival between treatment arms. The log-rank test yields a p-value of 0.93, providing no evidence to reject the null hypothesis of equal survival. Therefore, we conclude that there is no statistically significant difference in survival between the two chemotherapy regimens.

::: {.callout-note title="Interpretation of Graph Elements" collapse="true"}
**X-Axis: Time (days)**

The X-axis shows the time since enrollment in days. The furthest value to the right (~1000 days) marks the maximum observed survival time in the dataset. Shorter durations towards the left reflect early mortality events. 

**Y-Axis: Survival Probability**

The Y-Axis is the Kaplan-Meier estimate of the probability that a patient survives beyond a certain timepoint. The axis starts at 1 (100% survival at time 0), and decreases with each event (death).

**Curves (Blue = Standard, Orange = Test)**

Each step downward represents a death. Tick marks would represent a censored individual (few in this dataset). 

**Shaded Area: 95% Confidence Interval** 

Around each curve is a shaded band indicating **uncertainty** in the estimate. Wider intervals to the right of the chart reflect fewer patients remaining and hence more variability in survival probability. 

**p = 0.93** 

The log-rank test result tests the null hypothesis: there is no difference in survival between the groups. The p-value of 0.93 provides strong evidence _against_ rejecting the H0. In other words, there is no statistically significant difference in survival between the Standard and Test treatments. If the curves had diverged strongly at some point, the P-value would be lower. 

**"Number at Risk" Table**

This shows how many participatns in each treatment group are still alive and uncensored at various timepoints (0, 250, 500, 750, 1000). As time progresses the number declines due to deaths or censoring. Few participants remain beyond 500 days, which is why the Confidence Intervals widen later in the graph. 

- The **p-value** is from a log-rank test (H₀: no difference in survival curves).
- The **shaded areas** represent 95% confidence intervals.
- The **risk table** shows the number of patients still at risk at each time point.

Visually, if the curves separate early and remain apart, it suggests a potential treatment effect.
:::

## Log-Rank Test

The log-rank test is a statistical test used to compare the survival distributions of two or more groups. It assesses whether there are significant differences in survival times between the groups, based on the observed and expected number of events (deaths) at each time point.

::: {.callout-note title="What is the Log-Rank Test?" collapse=true}

The **log-rank test** is a non-parametric statistical test used to compare the survival distributions of two or more independent groups. It evaluates the **null hypothesis** that there is no difference in the hazard rates between the groups over time—that is, the survival curves are statistically indistinguishable.

The test works by comparing the **observed number of events (e.g., deaths)** in each group to the **expected number of events** under the assumption that all groups share the same survival function. At each time point where an event occurs, the log-rank test calculates the difference between the observed and expected events across groups, summing these differences over time to produce a test statistic.

This statistic approximately follows a **chi-squared distribution**, and the resulting p-value indicates whether the observed differences are likely due to chance. A **small p-value (typically < 0.05)** suggests a statistically significant difference in survival between groups.

In the context of **Cox proportional hazards analysis**, the log-rank test is often used **before fitting a Cox model** to determine whether a survival difference exists between categorical groups (e.g., treatment arms). While the log-rank test is limited to univariate group comparisons and assumes proportional hazards, it provides a simple and intuitive starting point for exploring potential treatment effects or group differences in survival.

:::


```{r}
# Perform the log-rank test
log_rank_test <- survdiff(surv_obj ~ trt, data = veteran)
log_rank_test
```

```{r}
# Extract the p-value
1 - pchisq(log_rank_test$chisq, df = length(log_rank_test$n) - 1)
```

A log-rank test was performed to compare survival distributions between the control and intervention treatment groups. The test yielded a chi-square statistic of 0.008 with 1 degree of freedom, resulting in a p-value of 0.93. This indicates no statistically significant difference in survival between the two treatment arms. The observed and expected number of events (deaths) were nearly identical in both groups, further supporting the conclusion that survival outcomes did not differ meaningfully between treatments in this sample.

::: {.callout-note title="Understanding Chi-Square and Degrees of Freedom" collapse=true}

The **chi-square (χ²) statistic** is a measure of how much the observed data deviate from what we would expect under a specific statistical model — in this case, the assumption that survival curves are the same across groups.

In tests like the **log-rank test** or **Cox model diagnostics**, the chi-square statistic is calculated by comparing the **observed** vs **expected** number of events (deaths) for each group at every time point. The greater the deviation between these values, the larger the chi-square statistic becomes.

The **degrees of freedom (df)** represent the number of independent comparisons being made. For example:
- If you're comparing **two groups**, there’s only one degree of freedom (df = 1).
- If you're comparing **k groups**, df = k – 1.

Once the chi-square statistic and degrees of freedom are known, we can compute a **p-value** — the probability of observing such a deviation (or more extreme) if the null hypothesis were true.

- **Small p-value (e.g., < 0.05)**: Strong evidence against the null hypothesis → suggests a difference between groups.
- **Large p-value (e.g., > 0.05)**: No strong evidence → we fail to reject the null hypothesis.

In short:  
**Chi-square** quantifies the discrepancy.  
**Degrees of freedom** tell us how many comparisons we’re making.  
**P-value** tells us whether the discrepancy is statistically meaningful.

:::

::: {.callout-note title="Why Continue Cox Modeling After a Non-Significant Log-Rank Test?" collapse=true}

Even if the Kaplan-Meier curve and log-rank test show **no statistically significant difference** between groups, there are several valid and important reasons to continue with a Cox proportional hazards model:

**The Kaplan-Meier and log-rank test are unadjusted.**  
They assess only one variable at a time and do **not control for confounders** like age, prior treatment, or performance score. A Cox model allows you to adjust for these and may reveal effects that were previously masked.

**Cox regression provides more than just a p-value.**  
Unlike the log-rank test, the Cox model estimates **hazard ratios (HRs)** and their **confidence intervals**, which are essential for understanding the **magnitude and direction** of effects — even if not statistically significant.

**Exploratory analysis and skill building are valuable.**  
In educational or preliminary research contexts, building and diagnosing models provides critical insight. You may discover interesting patterns or generate hypotheses for future studies.

**Diagnostics may reveal model violations or improvements.**  
Even in the absence of significant treatment effects, diagnostic tools (like residual plots or proportional hazards tests) can highlight **non-linear effects, time-varying hazards, or outliers** that inform better modelling strategies.

**Bottom line:** The absence of significance in univariate survival tests does not mean "stop here." A Cox model is not just a hypothesis test — it's a framework for **adjusted inference**, **deeper insight**, and **model-based exploration**.

:::

# Cox Regression Models

The Cox proportional hazards model is a regression technique for analysing survival data. Unlike the Kaplan-Meier estimator, which is non-parametric and descriptive, the Cox model allows us to quantify the effect of multiple covariates on the hazard of an event — in this case, death.

This model assumes that each covariate has a multiplicative effect on the hazard and that this effect is constant over time (the proportional hazards assumption). By estimating hazard ratios, we can assess whether variables like treatment type, age, or prior therapy are independently associated with better or worse survival outcomes.

In this section, we begin with a simple model including only treatment group, and later expand to a multivariable model that includes continuous and categorical covariates. Each model is followed by diagnostic checks to assess validity and guide model refinement.

::: {.callout-note title="Understanding the Cox Proportional Hazards Model" collapse=true}

The **Cox proportional hazards model** (often just called the *Cox model*) is a cornerstone of survival analysis, allowing us to investigate how multiple variables influence the **risk (hazard)** of an event over time — without needing to specify the baseline hazard function.

**💡 What is a Hazard?**  
The hazard is the **instantaneous risk of the event** (e.g., death) at a given time, given that the individual has survived up to that point. The Cox model expresses this as:

> _h(t | X) = h₀(t) × exp(β₁X₁ + β₂X₂ + ... + βₖXₖ)_

Where:
- _h(t | X)_ is the hazard at time _t_ for a subject with covariates _X_,
- _h₀(t)_ is the **baseline hazard** (unspecified),
- _β_ are the **coefficients** estimating the effect of each covariate.

**🔍 Why Use the Cox Model Instead of Kaplan-Meier?**

| Kaplan-Meier                             | Cox Model                                        |
|------------------------------------------|--------------------------------------------------|
| Non-parametric (no assumptions)          | Semi-parametric (assumes proportional hazards)   |
| Describes survival by **groups only**    | Models the **effect of variables**               |
| Great for **initial exploration**        | Great for **inference and prediction**           |
| Can't adjust for **confounders**         | Can include **multiple covariates**              |
| Visual tool                              | Inferential and predictive tool                  |

In practice, it's common to start with a **Kaplan-Meier plot** to visualize survival, then use a **Cox model** to quantify the effect of predictors and control for confounding.

**📏 What Does "Proportional Hazards" Mean?**  
The Cox model assumes that the hazard ratio between two individuals is **constant over time**.  
For example, if a treatment reduces the hazard by 30%, it is assumed to reduce it by 30% at every point in time. This is the **proportional hazards assumption**, and it **must be checked** using diagnostic tools like **Schoenfeld residuals**.

**📊 What the Model Produces:**  
- **Coefficients (β):** Log hazard ratios
- **exp(β):** Hazard ratios (HR), easier to interpret
- **p-values:** From the Wald test, testing whether β ≠ 0
- **Global tests:** Likelihood ratio, Score (logrank), Wald

Example:
- `HR = 1.50` → 50% higher hazard  
- `HR = 0.70` → 30% lower hazard

**🧠 Summary:**  
The Cox model is a flexible, interpretable, and robust way to analyze the effect of variables on survival, especially when you want to **adjust for multiple covariates**. However, it does require checking assumptions and being thoughtful about variable selection and transformations (e.g., linearity for continuous variables).

:::

## Univariable Models

In this section, we will fit a series of univariable Cox models to assess the effect of individual covariates on survival. This allows us to understand the contribution of each variable before moving on to multivariable models.

### Treatment Group Only

We begin with a univariable Cox model that includes only the treatment variable (`trt`). This allows us to assess the effect of treatment on survival without adjusting for any other covariates.


```{r}
# Fit the Cox Proportional Hazards model
cox_model <- coxph(surv_obj ~ trt, data = veteran) # Where trt is a binary treatment variable with levels Control and Intervention

# Summarize the model
summary(cox_model)

```

::: {.callout-note title="Interpretation of Cox Table" collapse="true"}

| Term          | Value          | What It Means                                                                                                                                                                         |
| ------------- | -------------- | ------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- |
| **coef**      | 0.01774        | This is the log hazard ratio (log(HR)) comparing the *Intervention* group to *Control*. Close to 0 means almost no effect.                                                            |
| **exp(coef)** | 1.01790        | This is the hazard ratio (HR). A value of 1.018 means the hazard (risk of death) is 1.8% higher in the Intervention group compared to Control. However at this level the value is insigificant. |
| **se(coef)**  | 0.18066        | The standard error. This gives us the variability around the estimate.                                                                                                                |
| **z**         | 0.098          | The Wald test statistic. The low value implies no stastical significange.                                                                                     |
| **p-value**   | 0.922          | Far above the conventional 0.05 threshold. This is very strong evidence *against* any meaningful treatment effect.                                                           |
| **95% CI**    | (0.7144, 1.45) | Wide confidence interval, spanning both below and above 1—so we can't rule out either harm or benefit.                                                                                |
**Concordance** The C-index, or concordance statistic is the probability that a candomly hosen subject who died earlier had a higher predicted risk. At 0.525 it is random chance, suggested poor predictive power. 

::: 

**Interpretation of the Cox Model Results**

The Cox proportional hazards model found no statistically significant difference in survival between the Control and Intervention groups (HR = 1.02, 95% CI: 0.71–1.45, p = 0.922). The hazard ratio close to 1 suggests that the treatment had no appreciable effect on the risk of death. Additionally, the wide confidence interval includes both potential harm and benefit, indicating uncertainty in the estimate. The concordance statistic (0.525) implies the model has poor discriminative ability, performing only slightly better than random chance. Overall, the analysis provides no evidence that the intervention improves survival outcomes in this dataset.

```{r}
# Use ggadjustedcurves for plotting adjusted survival curves directly from the Cox model
ggadjustedcurves(
  cox_model,                     # The fitted Cox model
  data = veteran,                # The original data used to fit the model
  variable = "trt",              # The variable to plot adjusted curves for
  method = "average",            # Method for adjusting (average effect of other covariates)
  conf.int = TRUE,
  legend.labs = c("Control", "Intervention"),
  legend.title = "Treatment Group",
  risk.table = TRUE,             # Include the risk table
  risk.table.title = "Number at Risk",
  xlab = "Time (days)",
  ylab = "Adjusted Survival Probability",
  title = "Cox Model Adjusted Survival Curves",
  palette = c("#A569BD", "#45B39D"),
  ggtheme = theme_minimal(base_size = 13) +
    theme(
      legend.position = "top",
      plot.title = element_text(hjust = 0.5, face = "bold"),
      panel.grid.major = element_line(color = "grey80", linewidth = 0.4),
      panel.grid.minor = element_line(color = "grey90", linewidth = 0.2)
    ),
  tables.theme = theme(
    axis.text.y = element_text(size = 10, margin = margin(r = 12), lineheight = 2)
  )
)
```

::: {.callout-tip title="Why use `ggadjustedcurves()` for this plot" collapse="true"}
For plotting survival curves adjusted for covariates from a Cox Proportional Hazards model, the ggadjustedcurves() function from the survminer package offers one of the most direct and statistically appropriate approaches—particularly as model complexity increases.

- Statistical validity: ggadjustedcurves() uses the fitted cox_model to compute adjusted survival probabilities for the variable of interest (in this case, trt). When method = "average" is specified (the default when covariates are present), it returns marginal estimates—appropriate for interpreting average treatment effects across the population. In our simpler model (~ trt), it effectively plots the model-predicted survival curve for each treatment group.

- Robustness: This function is purpose-built for adjusted curve plotting and avoids many of the pitfalls seen with manually passing newdata to survfit() and plotting with ggsurvplot(), such as misaligned strata or legend issues. It correctly associates the model, the data, and the plot elements, including the risk table.

- Simplicity: It eliminates the need for manually constructing a new_trt data frame or calling survfit() separately, resulting in cleaner, more reproducible code—particularly useful in collaborative or long-term projects.
:::

The Cox-adjusted survival curves for the Control and Intervention groups are almost perfectly overlapping, indicating no meaningful difference in predicted survival probabilities between the two treatment arms. This is consistent with the hazard ratio from the Cox model, which was very close to 1, and not statistically significant. Visually, the indistinguishability of the curves reinforces the conclusion that the treatment had no observable impact on survival in this dataset.

::: {.callout-note title="Why do the lines overlap?" collapse="true"}
The survival curves for the Control and Intervention groups overlap almost entirely in this plot. This is because the Cox proportional hazards model found no significant difference between the two groups in terms of survival (hazard ratio ≈ 1, *p* = 0.922). As a result, the model-predicted curves are nearly identical. The faint visibility of the Control group line is due to it being directly behind the Intervention group line.
:::

**Model Diagnostics**

Before interpreting the results of a Cox proportional hazards model, it is essential to verify that the model's underlying assumptions hold. The Cox model assumes that the effect of each covariate on the hazard is constant over time — the so-called **proportional hazards assumption** — and that **continuous variables have a linear relationship with the log hazard**. Violations of these assumptions can lead to incorrect or misleading conclusions.

In this section, we apply a series of **diagnostic tools** to assess the validity of our model:
- **Schoenfeld residuals** to check the proportional hazards assumption.
- **Martingale residuals** to detect non-linearity and outliers.
- **Functional form checks** to evaluate whether continuous variables such as age are correctly specified.
- Where necessary, we apply **transformations** (e.g., splines) to better model the relationship between covariates and survival outcomes.

These diagnostics ensure that the model remains a **valid and reliable tool** for making inferences about survival and covariate effects.

**Checking the Proportional Hazards Assumption**

The central assumption of the Cox proportional hazards model is that the **hazard ratios remain constant over time** — that is, the relative risk between any two individuals does not vary as time progresses. This is known as the **proportional hazards (PH) assumption**, and verifying it is a crucial step in validating the model.

**Schoenfeld Residuals**

To assess this assumption, we use **Schoenfeld residuals**, which examine the time-dependent behavior of covariate effects. If the residuals show a systematic trend over time, it suggests that the effect of a covariate may change — violating the PH assumption.

We assess proportionality in two ways:
- **Statistical tests**: We use the `cox.zph()` function to perform a global test and individual tests for each covariate. A p-value < 0.05 indicates evidence against proportional hazards.
- **Visual inspection**: Plots of scaled Schoenfeld residuals with a smoothed line allow us to visually assess whether the covariate effects change over time.

If either the test or the plots suggest deviation from proportionality, we may need to consider **time-varying effects**, stratification, or alternative modeling strategies.

```{r}
# Test the PH assumption
cox.zph(cox_model)
```

::: {.callout-note title="Understanding Chi-Squared and p-values in PH Assumption Testing" collapse=true}

When checking the **proportional hazards (PH) assumption** in a Cox model using `cox.zph()`, each covariate — and the model as a whole — is tested for **time-dependent effects**.

- **Chi-squared statistic**: Measures how much the residuals deviate from what would be expected if the covariate had a constant effect over time (i.e., proportional hazards). A larger value suggests a stronger departure from proportionality.

- **Degrees of freedom (df)**: Typically 1 for individual covariates, and equal to the number of covariates in the global test.

- **p-value**: Quantifies the probability that the observed chi-squared value (or more extreme) could occur **by chance** if the PH assumption holds.

**How to interpret**:
- `p > 0.05`: No significant evidence of time-varying effects — PH assumption likely holds.
- `p < 0.05`: Suggests the effect of the covariate may change over time — a potential violation of the PH assumption.

Use both **statistical tests** and **residual plots** for a more robust assessment. A borderline p-value (e.g., 0.06) should be interpreted with caution and always viewed in conjunction with the corresponding plot.
:::

**Interpretation of PH Assumption Test (Schoenfeld residuals)**

- trt p = 0.06: This is slightly above the conventional threshold of 0.05. It suggest no strong evidence of violation of the proportional hazards (PH) assumptions for _treatment_. 
- GLOBAL p = 0.06: This tests the model overall. Not statistically significant so we do not reject the PH assumptions. 

**Porportional Hazards Assumption**

The proportional hazards assumption was assessed using Schoenfeld residuals via the cox.zph() function. The test for the treatment variable (trt) yielded a chi-square statistic of 3.54 with a p-value of 0.06, and the global test returned the same values (χ² = 3.54, df = 1, p = 0.06).

Although the p-values are slightly above the conventional significance threshold of 0.05, they do not provide sufficient evidence to reject the null hypothesis that the proportional hazards assumption holds. As such, the Cox proportional hazards model is considered appropriate for this analysis, though the borderline result suggests caution and may warrant further inspection in more complex models.

We visualize the scaled Schoenfeld residuals for each covariate to assess whether their effects on the hazard function remain constant over time. A roughly horizontal smoothed line near zero with confidence bands overlapping zero suggests the proportional hazards assumption holds.

**Schoenfeld Residuals Plot**

```{r}
plot(cox.zph(cox_model))
```

::: {.callout-note title="What the Visual Diagnostic Shows" collapse="true"}

*The y-axis:* Scaled Schoenfeld residuals for trt. If the proportional hazards assumption holds, these residuals should fluctuate randomly around zero.

*The x-axis:* Time (or transformed time).

*The line:* A smoothed spline fit (typically a loess curve) showing the trend of residuals over time.

*The confidence band:* Helps you assess whether the trend significantly deviates from horizontal (i.e., no time-dependent effect).

*A flat line (slope ≈ 0):* Suggests the hazard ratio for trt does not change over time — good!

*A wavy or sloped line:* Suggests potential violation of the proportional hazards assumption — this would back up a significant p-value.

*x-axis (Time):* Time since baseline (likely in days, from your dataset).

*y-axis (Beta(t) for trt):* Time-varying estimate of the log hazard ratio for the treatment group (trt), based on Schoenfeld residuals.

*The solid black line:* A smoothed curve showing how the estimated coefficient for trt might change over time.

*Dashed lines:* Approximate 95% confidence bands around that smoothed estimate.

*Open circles:* Residuals for each individual event — they don't need to follow a pattern, but their distribution can hint at model fit.

If the solid line is roughly flat and within the dashed lines, the effect of treatment does not change with time, and the PH assumption is reasonably upheld.

If the solid line drifts noticeably over time, or wanders outside the dashed lines, that's evidence that the treatment effect may change — a violation of the PH assumption.

In our case, the line is gently curved, but still well within the dashed boundaries. There's no strong time trend — the log hazard ratio is hovering around zero and doesn't appear to increase or decrease dramatically.

:::

**Schoenfeld Residuals Interpretation**

To assess whether the proportional hazards (PH) assumption holds for the treatment variable (trt), we examined Schoenfeld residuals using both statistical and graphical diagnostics. The global test for proportionality yielded a p-value of 0.06, indicating no statistically significant violation of the PH assumption at the conventional α = 0.05 level, though it approaches significance. The smoothed residual plot for trt shows a relatively stable trend over time, with the estimated log hazard ratio remaining within the 95% confidence bands across the follow-up period. While there is some minor curvature in the smoothed line, it does not suggest substantial deviation from proportionality. These findings support the use of the Cox proportional hazards model for analyzing the effect of treatment on survival in this dataset.

**Martingale Residuals (for Functional Form of Covariates)**

Martingale residuals are plotted against continuous covariates to assess whether the Cox model’s assumption of a linear relationship on the log-hazard scale holds. Non-random patterns or curvature may suggest the need for transformations, such as splines.

::: {.callout-note title="Understanding Martingale Residuals" collapse=true}

**Martingale residuals** are used to assess whether the **functional form** of a continuous covariate is correctly specified in a Cox proportional hazards model.

**🔍 What are they?**  
Martingale residuals represent the difference between the **observed number of events** and the **expected number of events** under the fitted model for each subject.  
They range from **–∞ to +1** and are **asymmetrically distributed**.

**📊 Why do we use them?**  
To check the **linearity assumption**: the Cox model assumes that the covariate has a **linear effect on the log hazard**. If this is not true (e.g. the effect increases or decreases non-linearly), the model may misestimate risk.

**✅ What should you look for in the plot?**  
You typically plot Martingale residuals **against a continuous covariate** (e.g. age). Then:

- A **random scatter** around the horizontal line at 0 suggests linearity is reasonable.  
- A **systematic curve** (e.g., U-shaped, S-shaped) suggests **non-linearity** and the need for transformations (e.g., log(age), age², splines).

**🧠 Practical Tip:**  
Martingale residuals are **not symmetric**, so smooth trends can be hard to detect visually. It's a good idea to pair this check with **`ggcoxfunctional()`** or use **splines** to model nonlinear effects more flexibly.
:::


```{r}
# Generate Martingale residuals
martingale_resid <- residuals(cox_model, type = "martingale")

# Add residuals to your data
veteran$martingale <- martingale_resid

# Plot residuals against a continuous covariate (e.g., time or another variable)
# Here we plot against the linear predictor for simplicity
veteran$linear_pred <- predict(cox_model, type = "lp")

ggplot(veteran, aes(x = linear_pred, y = martingale, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Linear Predictor",
    x = "Linear Predictor (Log Hazard)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))


```

::: {.callout-note title="What do the Martingale Residuals Show" collapse="true"}

Martingale residuals help us check whether the variables in our Cox model are specified correctly — particularly useful for continuous covariates. However, when the variable in question is binary (like our treatment group: Control vs Intervention), the Martingale residual plot will display two vertical stacks of points — one for each group.

In the plot above:

- The **x-axis** represents the linear predictor (log hazard). Because the treatment is binary, there are only two unique predictor values: one for the control group and one for the intervention group.
- The **y-axis** shows the Martingale residuals for each individual.
- The **dashed horizontal line at 0** acts as a visual guide to see if residuals are randomly scattered above and below it.

We want to see:
- No clear patterns or trends in the residuals,
- A reasonably symmetric spread around zero in each group,
- No outliers clustering far from the bulk of points.

In this case, both groups show a spread of residuals centered roughly around zero, with no curvature or skew that would suggest model misspecification. That's a good sign — the model's structure seems to handle the treatment variable well, and there's no obvious reason to suspect we need to transform it or add interaction terms.

:::

The Martingale Residuals vs Linear Predictor plot helps us check the fit of our Cox model—specifically, whether our model's assumptions about the shape of relationships between predictors and survival are holding up.

Martingale residuals were plotted against the linear predictor to assess model fit. As expected with a binary treatment variable, the plot showed two vertical bands of points corresponding to the Control and Intervention groups. The residuals appeared symmetrically distributed around zero within each group, with no obvious non-linear patterns or systematic bias. This suggests the model's functional form for the treatment variable is appropriate.

::: {.callout-tip title="Why we haven't checked the functional form of covariates" collapse="true"}

The Cox model assumes a linear relationship between continuous covariates and the log hazard. In our current model, only the treatment group (`trt`) is included, which is a categorical variable. 

If we were to build a more complex model including continuous predictors like age, Karnofsky score (`karno`), or time since diagnosis (`diagtime`), we would need to assess whether their relationship with the log hazard is truly linear. This is typically done by plotting Martingale residuals or using `ggcoxfunctional()`.

For now, since our model does not include continuous covariates, this step is not necessary — but it's an important part of model diagnostics when expanding our analysis.

:::

**Additional Covariates**

To refine our model and better explain variability in survival outcomes, we now introduce additional covariates. Specifically, we will examine age (a continuous variable) and prior treatment history (a binary variable) to assess their independent effects and contributions to the model.

::: {.callout-note title="Why Add Covariates to the Cox Model?" collapse=true}

Adding covariates to a Cox proportional hazards model allows us to **control for confounding**, improve **predictive accuracy**, and explore **independent associations** between predictors and survival time.

- **Control for confounding:** Variables like age or prior treatment may influence both the outcome and the treatment assignment. Including them helps isolate the effect of the treatment variable.
- **Improve model fit:** Additional predictors can increase the explanatory power of the model, helping to reduce residual variability.
- **Estimate adjusted effects:** By controlling for other covariates, we can determine the *independent* effect of each variable on the hazard of the event.

✅ When building survival models, it's important to choose covariates based on clinical relevance and theoretical justification—not just statistical significance.

:::


To expand the model beyond treatment alone, we now include additional patient characteristics: `age` (a continuous variable) and `prior` (indicating whether the patient received prior therapy, a binary variable). Including these covariates allows us to better account for baseline differences between patients, assess their independent associations with survival, and evaluate how these variables may influence the effect of treatment.

::: {.callout-tip title="Why include more covariates?" collapse="true"}

Adding additional covariates serves several important purposes in survival analysis:

- **Control for confounding:** Variables like age and prior treatment may influence both survival and the likelihood of receiving a particular treatment. Including them helps isolate the true effect of treatment.
- **Improve model accuracy:** Covariates that predict survival can increase the model's explanatory power and reduce residual variation.
- **Uncover independent predictors:** Some variables (like prior therapy) may have a meaningful relationship with survival, regardless of treatment.
- **Assess effect modification:** Once multiple variables are included, we can test for interactions — for example, whether the effect of treatment depends on age or prior treatment status.

We'll begin by examining `age` as a continuous covariate (with extra attention to its functional form), followed by `prior` as a binary covariate. We'll then combine them in a multivariable Cox model.

:::

### Continuous Variable: Age

Age is a clinically relevant factor that may influence survival outcomes in lung cancer. In this section, we assess whether age is independently associated with survival time and verify that its relationship with the hazard is appropriately modeled as linear on the log-hazard scale.

::: {.callout-tip title="Why model age carefully?" collapse="true"}
Unlike categorical variables, continuous predictors like age require us to check their **functional form** in the model. The Cox model assumes a **linear relationship** between the covariate and the log hazard — meaning each 1-year increase in age is assumed to have the same proportional effect on the hazard.

If this assumption doesn't hold (e.g., if risk increases faster at older ages), we may need to use transformations or splines to accurately model the relationship.

We'll start by fitting a Cox model including age and then assess the functional form using Martingale residuals.
:::

```{r}
cox_age <- coxph(Surv(time, status) ~ trt + age, data = veteran)
summary(cox_age)
```

A Cox proportional hazards model was fitted to examine the relationship between survival time and two predictors: treatment group (trt) and patient age (age). The model included 137 patients, with 128 events (deaths).

The hazard ratio for the intervention group was 0.996 (95% CI: 0.697 – 1.425; p = 0.984), suggesting virtually no difference in hazard compared to the control group after adjusting for age. This effect was not statistically significant.

For age, the hazard ratio was 1.008 (95% CI: 0.989 – 1.027; p = 0.436), indicating that each additional year of age was associated with a 0.8% increase in the hazard of death. However, this association was also not statistically significant and the confidence interval included 1, suggesting uncertainty about the true effect.

Overall, the model's ability to discriminate between patients with longer vs shorter survival times was weak, with a concordance index of 0.514 (where 0.5 = random chance). The global model statistics — including the likelihood ratio test, Wald test, and score test — were all non-significant (p = 0.7), indicating that neither variable added substantial predictive value to the model.

**Interpretation:**

- There is no evidence of a treatment effect after adjusting for age.

- Age may have a small effect, but in this sample, it is not statistically or clinically meaningful.

- The overall model performs poorly in predicting survival in this dataset.

::: {.callout-note title="Interpreting the Cox Model Output" collapse="true"}

Understanding the `coxph()` output is essential for interpreting survival models. Here's a breakdown of the most common components you'll encounter in both univariable and multivariable Cox regressions:

**🔹 _coef_**  
This is the **log hazard ratio** — the raw estimate from the model. It represents the change in the **log hazard** of the event (e.g., death) associated with a **one-unit increase** in the covariate (for continuous variables) or **relative to a reference group** (for categorical variables).

- Positive `coef` → increased hazard (i.e., higher risk)  
- Negative `coef` → decreased hazard (i.e., protective)  
- A `coef` close to 0 suggests little to no effect

🧠 _Note:_ Log hazards are difficult to interpret directly, so we usually exponentiate them.

**🔹 _exp(coef)_**  
This is the **hazard ratio (HR)** — the exponentiated `coef`, which expresses effect size more intuitively.

- `exp(coef) > 1` → Increased hazard  
- `exp(coef) < 1` → Decreased hazard  
- `exp(coef) = 1` → No effect

🔎 Example: `exp(coef) = 1.20` means a 20% higher hazard.  
`exp(coef) = 0.80` means a 20% lower hazard.

**🔹 _se(coef)_**  
The **standard error** of the coefficient — reflects how much uncertainty surrounds the estimate.

- Smaller values = more precise estimates  
- Larger values = more uncertainty (due to small samples or noisy data)

**🔹 _z_ and _Pr(>|z|)_**  
- `z` is the **Z-statistic**, calculated as `coef / se(coef)`  
- `Pr(>|z|)` is the **p-value** from the Wald test for each coefficient

This tests the **null hypothesis** that the covariate has **no effect** (`coef = 0`).

- `p < 0.05` → Statistically significant  
- `p > 0.05` → Not statistically significant

**📊 _Model-Level Statistics_**

**🔸 _Concordance_**  
- Measures the model's ability to correctly rank survival times (similar to AUC for classification)  
- Ranges from `0.5` (random) to `1.0` (perfect)  
- In clinical data, values of `0.6–0.75` are typical

**🔸 _Likelihood Ratio Test_**  
- Compares the full model to a null (intercept-only) model  
- Tests overall model fit  
- `p < 0.05` = your model significantly improves fit

**🔸 _Wald Test_**  
- Aggregates the individual Wald tests across all covariates  
- Tests whether any coefficients differ significantly from 0

**🔸 _Score (Logrank) Test_**  
- Rank-based test comparing observed vs expected survival  
- Can be slightly more sensitive in small samples  
- Often used as an overall model check

✅ _In practice:_ Use all three global tests and the concordance to evaluate how well the model explains variation in survival. If all are non-significant, the model may lack predictive utility.
:::


**Schoenfeld Residuals for Age**

```{r}
# Test proportional hazards assumption for the model including age
cox_zph_age <- cox.zph(cox_age)
print(cox_zph_age)
```

Proportional hazards were tested using Schoenfeld residuals. The assumption was met for **age** (`p = 0.195`) but marginally violated for **treatment group** (`p = 0.055`), which is slightly below the conventional threshold of 0.05. The **global test** for the model (`p = 0.045`) was also just below 0.05, indicating a **possible overall violation** of the proportional hazards assumption.

While these results do not strongly indicate a breakdown of the PH assumption, the borderline p-values warrant caution. A visual inspection of the residual plots (`plot(cox.zph())`) can help determine whether the trend is substantial or random variation.

::: {.callout-note title="Interpreting Schoenfeld Residuals" collapse="true"}

Schoenfeld residuals are used to test the **Proportional Hazards (PH) assumption** in Cox models — that the effect of each covariate remains **constant over time**.

**How to interpret the output:**

- Each row (e.g., `trt`, `age`) gives a **test of proportionality** for that specific variable.
- The `GLOBAL` row tests **all covariates together**.

**What the numbers mean:**

- `chisq` is the test statistic (larger values = more evidence of time-varying effects).
- `p` is the p-value testing **null hypothesis: proportional hazards holds**.

**Rule of thumb:**
- `p < 0.05` → **PH assumption violated** (bad)
- `p > 0.05` → **PH assumption holds** (good)

🧠 _Note:_ Slight violations (e.g., `p = 0.055`) might not be a problem in practice — but major ones (e.g., `p = 0.01`) should be addressed.

:::

```{r}
# Visual check
plot(cox_zph_age)
```

The Schoenfeld residual plot for age, derived from the multivariable Cox model, reveals no strong evidence of violation of the proportional hazards assumption. The smoothed coefficient curve remains relatively flat across time, with the 95% confidence band consistently overlapping zero. This suggests that the effect of age on the hazard function does not meaningfully vary with time. This visual impression aligns with the result from the formal statistical test (χ² = 1.68, df = 1, p = 0.195), providing no significant indication that the proportional hazards assumption has been violated for age in this model.

::: {.callout-note title="How to Interpret Schoenfeld Residual Plots" collapse="true"}

This plot checks the **Proportional Hazards (PH) assumption** visually for a covariate (e.g., `age`).

**What we are seeing:**

- **X-axis:** Time  
- **Y-axis:** Estimated **effect (beta)** of the covariate at each time point  
- **Dots:** Residuals at each event time  
- **Solid line:** Smoothed trend of the coefficient over time  
- **Dashed lines:** 95% confidence interval of that trend  

**What we want to see:**

✅ A **flat line near zero**, with the confidence band **containing zero** throughout time → suggests **PH assumption holds**.

⚠️ A **wavy trend** or curve **drifting far from zero** → suggests **non-proportional hazards** (i.e., the effect of the covariate changes over time).

**Interpreting the plot:**

If the smoothed line is approximately flat and the confidence interval consistently includes zero — with no strong patterns or directional trends — then the PH assumption is likely satisfied for that covariate.

:::

**Martingale Residuals for Age to Assess Functional Form**

To assess whether age is appropriately modeled as a linear predictor in the Cox model, we examine Martingale residuals plotted against age to detect non-linearity in its functional form.

```{r}
# Generate Martingale residuals
veteran$martingale_age <- residuals(cox_age, type = "martingale")

# Generate linear predictor
veteran$linear_pred_age <- predict(cox_age, type = "lp")

# Plot residuals against age
ggplot(veteran, aes(x = age, y = martingale_age, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Age",
    x = "Age (years)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

Martingale residuals plotted against age showed no clear non-linear trend, with residuals generally scattered randomly across the range of age values. This suggests that modeling age as a linear term in the Cox model is reasonable. There is no strong evidence to support the need for transformations or spline-based modeling of age in this dataset.

::: {.callout-note title="How to Interpret Martingale Residuals for Functional Form" collapse="true"}

This plot helps assess whether the **relationship between a continuous variable** (like _age_) and the **log hazard** is **linear**, which is a core assumption of the Cox model.

- **X-axis:** The covariate
- **Y-axis:** Martingale residuals (bounded between -∞ and 1)
- **Points:** One per individual, representing how well their survival time is predicted
- **Dashed line at 0:** Reference line to help spot patterns

**What to look for:**
- A **random scatter** of points across the X-axis with **no clear curve** → linearity assumption is likely satisfied ✅
- A **systematic curve**, bend, or wave → suggests the effect of age might be **non-linear** ⚠️  
  In that case, consider modeling age using splines or transformations (e.g., `log()` or `^2`).

**Important:** These residuals are noisier near their bounds and should be interpreted cautiously, especially when sample size is limited.

:::

**Functional Form Check**

```{r}
# Fit a univariable Cox model to assess functional form of age
cox_age_only <- coxph(Surv(time, status) ~ age, data = veteran)

# Check the functional form visually using Martingale residuals
ggcoxfunctional(cox_age_only, data = veteran)
```


::: {.callout-note title="Assessing Functional Form with Martingale Residuals" collapse="true"}

**What this plot shows:**  
This diagnostic visualizes whether a **continuous covariate** (e.g., _age_) has a **linear relationship** with the **log hazard**, as assumed by the Cox model.

**X-axis:**  
The values of the continuous variable being assessed (e.g., age)

**Y-axis:**  
The **Martingale residuals** from a **null Cox model** (i.e., a model with no covariates)

**Dots:**  
Each dot represents one observation

**Smooth black line:**  
A **non-parametric smoother** that helps reveal trends in the residuals — this is the key to checking the shape of the relationship

**How to interpret it:**

- ✅ If the **black line is roughly straight and flat**, this suggests a **linear relationship** between the variable and the log hazard — which is **what the Cox model assumes**.
- ⚠️ If the line shows a **curved or non-linear pattern**, this suggests the variable may not be appropriately modeled with a simple linear term.  
    - This can indicate the need to use **transformations** (e.g., `log(age)`, `age^2`) or **splines**.

**Tip:**  
Small wobbles are not a major concern — we are primarily looking for clear trends, not minor bumps. Major **U-shapes, inflections, or long curves** are the most concerning signs.

:::

The functional form check using Martingale residuals and a non-parametric smoother suggests that age may not have a strictly linear relationship with the log hazard. The smooth curve appears U-shaped, implying that risk may not increase evenly with age across the full range. This indicates that the assumption of linearity may be violated, and that alternative modeling strategies (e.g., polynomial terms or splines) could improve model fit.

While the Martingale residual plot suggests that the relationship between age and the log hazard may not be strictly linear, age was not statistically significant in the model. As such, no transformation or spline was applied. This could be revisited in future modeling if age becomes more influential.

### Binary Variable: Prior Treatment

We now evaluate the variable `prior`, which indicates whether a patient received prior therapy before enrolment in the study. This binary variable may capture important differences in disease severity, treatment history, or prognosis that could influence survival outcomes.

::: {.callout-tip title="Why test binary covariates like 'prior'?" collapse="true"}
Even though `prior` is binary, its inclusion allows us to control for historical differences in patient treatment, which may affect survival independently of the trial treatment (`trt`). 

For binary variables, we don't need to assess the functional form, but we **do need to check the proportional hazards assumption** (as with any covariate in the Cox model). We'll interpret the hazard ratio as the relative difference in risk between those who received prior therapy and those who did not.
:::

**Fit Cox Model with Prior Treatment**

```{r}
# Fit the Cox model including prior treatment
cox_prior <- coxph(Surv(time, status) ~ trt + prior, data = veteran)
summary(cox_prior)
```

A Cox proportional hazards model was fitted to evaluate the association between survival and two predictors: treatment group (trt) and prior therapy (prior). The model was based on 137 patients, with 128 observed events.

The hazard ratio for the intervention group was 1.03 (95% CI: 0.72 – 1.46, p = 0.885), indicating no evidence of a treatment effect after adjusting for prior therapy. Similarly, prior therapy was not associated with survival, with a hazard ratio of 0.99 (95% CI: 0.95 – 1.03, p = 0.471).

Model discrimination was weak, with a concordance of 0.517, suggesting the model has limited ability to rank survival times accurately. Global model fit tests (likelihood ratio, Wald, and score) were all non-significant (p = 0.8), indicating that the included variables do not significantly improve model performance.

Neither treatment assignment nor prior therapy appear to be meaningful predictors of survival in this dataset. There is no evidence of an additive effect between these covariates.

**Schoenfeld Residuals for Prior Treatment**

```{r}
# Test proportional hazards assumption for the model including prior treatment
cox_zph_prior <- cox.zph(cox_prior)
print(cox_zph_prior)
```

Proportional hazards were assessed using Schoenfeld residuals for the model including treatment group and prior therapy. The global test was statistically significant (p = 0.048), suggesting potential violation of the proportional hazards assumption overall.

Individually, both treatment (p = 0.066) and prior therapy (p = 0.081) had p-values slightly above the conventional 0.05 threshold, indicating borderline evidence of non-proportionality. These results do not strongly confirm a violation, but they do warrant cautious interpretation.

Given the marginal nature of these findings, visual inspection of the Schoenfeld residual plots is recommended to assess whether any systematic time-varying effects are present.

```{r}
# Visual check
plot(cox_zph_prior)
```

The Schoenfeld residual plot for prior treatment shows a smoothed line that remains approximately flat over time, with the 95% confidence band consistently including zero. This suggests that the effect of prior treatment on the hazard is stable across the follow-up period, and supports the assumption of proportional hazards for this variable.

In contrast, the plot for treatment group (trt) exhibits a noticeable wavy pattern in the smoothed line, particularly in the early to mid-study period. While the confidence bands still broadly include zero, the shape of the curve raises some concern about potential time-varying effects. Combined with a borderline p-value from the cox.zph() test (p = 0.060), this suggests the proportional hazards assumption for trt is not clearly violated, but may warrant caution and further consideration—especially in more complex models.

**Martingale Residuals for Prior Treatment**
```{r}
# Generate Martingale residuals
veteran$martingale_prior <- residuals(cox_prior, type = "martingale")
# Generate linear predictor
veteran$linear_pred_prior <- predict(cox_prior, type = "lp")
# Plot residuals against prior treatment
ggplot(veteran, aes(x = prior, y = martingale_prior, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Prior Treatment",
    x = "Prior Treatment (0 = No, 1 = Yes)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold")) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) # Adding jitter to avoid overlap

```

The Martingale residual plot for the prior variable shows two distinct clusters of points at 0 and 1, corresponding to patients with and without prior treatment, respectively. The residuals appear to be roughly symmetric around zero in both groups, with no strong indication of skew or heteroscedasticity. This supports the appropriateness of modeling prior as a binary variable in the Cox regression model. No transformation or alternative coding appears necessary.

## Multivariable Models

To build on the univariable analysis of treatment effects, we next fit a multivariable Cox proportional hazards model that incorporates additional patient-level covariates. This allows us to estimate the effect of treatment on survival **while adjusting for other potential prognostic factors**, such as age and prior treatment history. By doing so, we account for potential confounding and improve the model's explanatory power.

::: {.callout-tip title="What is a multivariable Cox model and why are we using it?" collapse="true"}
In a univariable Cox model, we examine the effect of **one predictor at a time** on the hazard of an event (e.g., death). While this can be useful for preliminary exploration, it doesn't account for the **influence of other variables** that might confound or modify the relationship.

A **multivariable Cox proportional hazards model** includes **multiple covariates** simultaneously. This allows us to:
- Estimate the **adjusted effect** of each variable while controlling for the others,
- Improve model accuracy and interpretability,
- Detect whether the apparent effect of one variable (e.g., treatment) persists after adjusting for others (e.g., age, comorbidities),
- Explore potential **interactions** or **effect modification**.

In this section, we'll first add a **continuous covariate** (`age`) and then a **binary variable** (`prior treatment`) to the model. We'll also assess whether the functional form of continuous variables is appropriate and ensure that model assumptions still hold.
:::

::: {.callout-tip title="Why not use logistic regression?" collapse="true"}
Logistic regression ignores time-to-event and censoring. When working with survival data, Cox models are preferred as they preserve the time structure and allow for censored cases.
:::

### Multivariable Cox Model with Age and Prior Treatment

In this section, we fit a multivariable Cox proportional hazards model that includes both age and prior treatment as covariates. This allows us to assess the independent effects of these variables on survival while controlling for treatment group.

```{r}
# Fit the multivariable Cox model including age and prior treatment
cox_multivariable <- coxph(Surv(time, status) ~ trt + age + prior, data = veteran)
summary(cox_multivariable)
```

::: {.callout-note title="Interpreting the Cox Model Output" collapse="true"}

Understanding the `coxph()` output is essential for interpreting survival models. Here's a breakdown of the most common components you'll encounter in both univariable and multivariable Cox regressions:

**🔹 _coef_**  
This is the **log hazard ratio** — the raw estimate from the model. It represents the change in the **log hazard** of the event (e.g., death) associated with a **one-unit increase** in the covariate (for continuous variables) or **relative to a reference group** (for categorical variables).

- Positive `coef` → increased hazard (i.e., higher risk)  
- Negative `coef` → decreased hazard (i.e., protective)  
- A `coef` close to 0 suggests little to no effect

🧠 _Note:_ Log hazards are difficult to interpret directly, so we usually exponentiate them.

**🔹 _exp(coef)_**  
This is the **hazard ratio (HR)** — the exponentiated `coef`, which expresses effect size more intuitively.

- `exp(coef) > 1` → Increased hazard  
- `exp(coef) < 1` → Decreased hazard  
- `exp(coef) = 1` → No effect

🔎 Example: `exp(coef) = 1.20` means a 20% higher hazard.  
`exp(coef) = 0.80` means a 20% lower hazard.

**🔹 _se(coef)_**  
The **standard error** of the coefficient — reflects how much uncertainty surrounds the estimate.

- Smaller values = more precise estimates  
- Larger values = more uncertainty (due to small samples or noisy data)

**🔹 _z_ and _Pr(>|z|)_**  
- `z` is the **Z-statistic**, calculated as `coef / se(coef)`  
- `Pr(>|z|)` is the **p-value** from the Wald test for each coefficient

This tests the **null hypothesis** that the covariate has **no effect** (`coef = 0`).

- `p < 0.05` → Statistically significant  
- `p > 0.05` → Not statistically significant

**📊 _Model-Level Statistics_**

**🔸 _Concordance_**  
- Measures the model's ability to correctly rank survival times (similar to AUC for classification)  
- Ranges from `0.5` (random) to `1.0` (perfect)  
- In clinical data, values of `0.6–0.75` are typical

**🔸 _Likelihood Ratio Test_**  
- Compares the full model to a null (intercept-only) model  
- Tests overall model fit  
- `p < 0.05` = your model significantly improves fit

**🔸 _Wald Test_**  
- Aggregates the individual Wald tests across all covariates  
- Tests whether any coefficients differ significantly from 0

**🔸 _Score (Logrank) Test_**  
- Rank-based test comparing observed vs expected survival  
- Can be slightly more sensitive in small samples  
- Often used as an overall model check

✅ _In practice:_ Use all three global tests and the concordance to evaluate how well the model explains variation in survival. If all are non-significant, the model may lack predictive utility.

:::

A multivariable Cox proportional hazards model was fitted to examine the association between survival and three covariates: treatment group, age, and prior therapy. After adjusting for all covariates:

- The hazard ratio for the intervention group relative to control was 1.00 (HR = 1.00; 95% CI: 0.70–1.44; p = 0.984), indicating no evidence of a treatment effect.

- Age had a hazard ratio of 1.01 per year (p = 0.462), suggesting no significant association with survival in this model.

- Prior therapy also showed no effect on hazard (HR = 0.99; p = 0.500).

Model discrimination was minimal (concordance = 0.507), and all global model tests were non-significant (p = 0.8), suggesting the model does not meaningfully predict survival in this dataset.

**Schoenfeld Residuals for Multivariable Model**

We now assess whether the **proportional hazards assumption** holds for each covariate in the multivariable Cox model. This step ensures that the effect of each variable — treatment group, age, and prior therapy — is consistent over time.

```{r}
# Test proportional hazards assumption for the multivariable model
cox_zph_multivariable <- cox.zph(cox_multivariable)
print(cox_zph_multivariable)
```

The proportional hazards assumption was assessed for the multivariable Cox model using Schoenfeld residuals. The global test was statistically significant (p = 0.032), suggesting that at least one covariate may have a time-varying effect on the hazard.

At the individual variable level:

- trt showed borderline evidence of non-proportionality (p = 0.060)

- prior also approached significance (p = 0.074)

- age did not show evidence of time-varying effects (p = 0.184)

While none of the individual tests crossed the 0.05 threshold, the cumulative result (GLOBAL p < 0.05) suggests some caution is warranted. Visual inspection of the Schoenfeld residual plots is recommended to evaluate whether these borderline findings correspond to meaningful deviations over time.

```{r}
# Visual check
plot(cox_zph_multivariable)
```

Visual inspection of the Schoenfeld residual plots for treatment group (trt), age, and prior treatment (prior) showed no substantial evidence of non-proportionality. The smoothed lines for each covariate remained relatively flat and were mostly contained within their respective 95% confidence bands across the follow-up period.

- For treatment group, the smoothed beta function showed some mild undulation, but it remained within the confidence bounds, indicating no major deviation from the proportional hazards assumption.

- For age, the trend line was nearly horizontal with minimal fluctuation, strongly supporting time-invariant effects.

- For prior treatment, the estimated beta also remained relatively stable over time, with the confidence interval consistently covering zero.

Together, these plots support the conclusion that the proportional hazards assumption is reasonably satisfied for each covariate in the multivariable model, aligning with the statistical test results where none of the individual covariate tests were significant at the 0.05 level.

**Martingale Residuals for Multivariable Model**

We now examine the Martingale residuals from the multivariable Cox model to evaluate the overall model fit and identify any patterns or potential outliers in the covariate effects.

```{r}
# Generate Martingale residuals
veteran$martingale_multivariable <- residuals(cox_multivariable, type = "martingale")
# Generate linear predictor
veteran$linear_pred_multivariable <- predict(cox_multivariable, type = "lp")
# Plot residuals against linear predictor
ggplot(veteran, aes(x = linear_pred_multivariable, y = martingale_multivariable, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Linear Predictor\n(Multivariable Model)",
    x = "Linear Predictor (Log Hazard)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```
The Martingale residuals plotted against the linear predictor for the multivariable model appear randomly scattered around the horizontal zero line, with no clear pattern, curvature, or systematic deviation. This suggests that the overall functional form of the covariates (treatment, age, and prior treatment) is reasonably specified. There is no strong evidence of model misspecification or non-linearity in the combined effects. However, the narrow spread of the linear predictor values (mostly between -0.2 and 0.2) reflects the modest effect sizes and limited variation in the log hazard estimates, which aligns with the model's lack of significant predictors.

::: {.callout-tip title="What This Plot Tells Us" collapse="true"}

Martingale residuals help assess how well the Cox model fits the data.

A good model will show residuals scattered symmetrically around zero with no systematic pattern.

Since we are plotting residuals against the linear predictor (the model's estimate of log hazard for each individual), this acts as a catch-all for checking overall model specification.

In this case:

- There's no strong non-linear trend.
- No excessive clustering or heteroskedasticity.
- Values are distributed fairly evenly above and below the zero line.

Conclusion: The model appears reasonably well-specified, and no major violations are detected.

:::

**Functional Form Check for Continuous Covariates**

We now recheck the functional form of the continuous variable age within the multivariable Cox model to ensure its effect on the log hazard remains appropriately modeled as linear.

```{r}
# Refit a univariable model with age (again, for functional form check)
cox_age_for_form_check <- coxph(Surv(time, status) ~ age, data = veteran)

# Plot functional form using ggcoxfunctional
ggcoxfunctional(
  fit = cox_age_for_form_check,
  data = veteran,
  point.col = "#E74C3C",
  ggtheme = theme_minimal(base_size = 13),
  title = "Functional Form Check: Age"
)
```

The Martingale residual plot for age reveals a clear U-shaped curve, suggesting that the assumption of linearity in the log hazard may not hold. This implies that age may have a non-linear effect on survival, and modeling it as a simple linear term may lead to misestimation of its effect. A transformation (e.g., quadratic or restricted cubic splines) may be warranted in further analysis.

::: {.callout-tip title="Why Check Functional Form Again in a Multivariable Model?" collapse="true"}

While functional form checks (e.g., using Martingale residuals) are **only necessary for continuous covariates**, it's important to **reassess them in the context of the full model** — even if they passed in a univariable setting.

- In a **univariable Cox model**, we check whether the continuous variable (e.g., age) has a linear relationship with the log hazard.
- However, once we add other covariates (like treatment or prior therapy), the relationship can **change due to confounding or interactions**.
- A variable might look linear on its own, but **show non-linearity when adjusted** for other predictors.

So we recheck **only the continuous variables** (e.g., age) in the multivariable model to ensure the assumption still holds.

:::

### Multivariable Cox Model with Spline-Transformed Age

We refit the Cox model using a restricted cubic spline to model the non-linear relationship between age and the log hazard. This allows for greater flexibility in capturing age-related risk while still accounting for treatment and prior therapy.

::: {.callout-note title="Understanding Spline and Non-Linear Transformations in Survival Models" collapse="true"}

**Why model transformations at all?**  
The Cox Proportional Hazards model assumes that each covariate has a **linear relationship** with the **log hazard**. For **continuous variables** like age, this means that each one-unit increase is expected to have the same multiplicative effect on the hazard — *regardless of whether someone is 40 or 80 years old*. This is often unrealistic.

To fix this, we apply **non-linear transformations** to more accurately capture the shape of the relationship between a variable (like age) and the outcome (e.g., death).

**What are spline transformations?**  
Splines are a **flexible way to model non-linear relationships** without explicitly specifying the form of the curve. They break a variable into *segments* using predefined cut points called **knots**, and fit separate curves (usually polynomials) between them, ensuring smoothness at the joins.

They allow the model to "bend" as needed — making them perfect for capturing complex patterns in the data.

**Common types of transformations:**

**• _Restricted Cubic Splines (Natural Splines)_:**  
This is the most common spline in Cox models. It creates piecewise cubic polynomials between knots, but constrains the ends to be linear — which prevents overfitting at the tails.

- ✅ Best for *flexible yet stable* curve fitting  
- 📌 Recommended when you suspect non-linearity but don't want to manually test transformations

**• _Quadratic or Polynomial Terms (e.g., age + I(age^2))_:**  
Adds a squared (or cubic) term to capture curvature.

- ✅ Simple to interpret
- ⚠️ Risky at the extremes — can overfit or behave erratically

**• _Log Transformations (e.g., log(age))_:**  
Used when the effect *decelerates* — i.e., rapid effect early on that flattens with increasing value.

- ✅ Great when data is skewed
- ⚠️ Doesn't work if the variable contains 0s or negatives

**• _Categorization (e.g., age groups)_:**  
Turns a continuous variable into bins (e.g., under 50, 50–65, over 65).

- ✅ Easy to explain
- ⚠️ Loses statistical power and may mislead if cutoffs are arbitrary

**When should you use each?**

| Transformation       | Use when...                                      |
|----------------------|--------------------------------------------------|
| **Splines (ns())**   | You want a smooth, data-driven relationship      |
| **Quadratic term**   | You suspect a simple "U-shaped" or curvilinear trend |
| **Log transform**    | The variable increases rapidly then levels off   |
| **Categories**       | Clinical guidelines or reporting require it      |

**How many knots do I use for splines?**  
A good rule of thumb (Harrell, 2015):

- 3 knots: small datasets (n < 100)
- 4–5 knots: medium datasets (n ~ 100–300)
- 5–7 knots: large datasets

Knots are often placed at **percentiles** of the variable — e.g., 5th, 35th, 65th, 95th percentile.

Final tip: Always re-check Schoenfeld residuals and Martingale residuals after applying any transformation.

:::

**Model Output**

```{r}
# Fit a Cox model with restricted cubic splines for age
cox_spline <- coxph(Surv(time, status) ~ ns(age, df = 3) + trt + prior, data = veteran)
summary(cox_spline)
```

A multivariable Cox proportional hazards model incorporating a restricted cubic spline transformation of age (df = 3) was fitted to explore potential non-linear effects of age on survival. None of the individual spline terms were statistically significant at the 0.05 level (p = 0.906, 0.143, and 0.157), suggesting that while the spline captures flexibility, the overall contribution of age remains statistically weak.

The treatment group (p = 0.626) and prior therapy (p = 0.640) also remained non-significant. Model-level tests showed that the likelihood ratio test (p = 0.1), Wald test (p = 0.1), and score test (p = 0.1) were all non-significant, indicating that the model, as specified, may not provide strong explanatory power.

However, the model's concordance improved slightly to 0.559 compared to the linear age model, suggesting a modest increase in predictive ability. While not conclusive, the use of splines helps satisfy the functional form assumption and could be valuable if future models identify stronger non-linear trends or interactions with age.

**Schoenfeld Residuals for Spline Model**

```{r}
# Test proportional hazards assumption for the spline model
cox.zph(cox_spline)
```

A Cox proportional hazards model incorporating a restricted cubic spline transformation of age (df = 3) demonstrated acceptable fit to the proportional hazards assumption across all covariates. The global Schoenfeld residuals test was non-significant (p = 0.116), suggesting that the model adequately satisfies the proportional hazards assumption. Individually, spline-transformed age (p = 0.211), treatment group (p = 0.114), and prior treatment (p = 0.079) all showed no significant time-dependent effects.

```{r}
# Visual check of Schoenfeld residuals for spline model
plot(cox.zph(cox_spline))
```

The Schoenfeld residuals were examined for the final multivariable Cox model, which included a spline-transformed age term. Across all covariates—age (using splines), treatment group (trt), and prior treatment (prior)—the smoothed residual lines remained relatively flat and within the 95% confidence intervals, suggesting no strong evidence of time-varying effects.

- Age (spline-transformed): Despite some visual variability, the three spline basis functions for age each produce smoothed lines that remain stable across time. Importantly, the confidence intervals broadly contain zero, indicating that the assumption of proportionality is not clearly violated despite the model’s non-linear specification.

- Treatment (trt): The smoothed residual curve shows a mild wave-like pattern, but this fluctuation stays well within the confidence bands and does not indicate a strong trend over time. While a borderline global p-value was observed in earlier models, the residuals here provide no compelling evidence against the proportional hazards assumption for treatment.

- Prior Treatment (prior): The smoothed residual line is relatively flat, and although the lower bound of the 95% confidence interval occasionally hovers just below zero, this does not strongly suggest non-proportionality. The overall trend appears stable, and the residuals provide visual reassurance that prior treatment's effect on hazard is consistent over time.

**Martingale Residuals for Spline Model**

```{r}
# Residuals vs linear predictor
veteran$martingale_spline <- residuals(cox_spline, type = "martingale")
veteran$linear_pred_spline <- predict(cox_spline, type = "lp")

ggplot(veteran, aes(x = linear_pred_spline, y = martingale_spline, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Linear Predictor\n(Spline Model)",
    x = "Linear Predictor (Log Hazard)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The plot displays Martingale residuals against the linear predictor (log hazard) for the spline-transformed Cox model. The residuals appear randomly scattered around the horizontal zero line, without a discernible systematic pattern or curvature. This suggests that the model has adequately captured the relationship between the covariates and the outcome.

The use of restricted cubic splines for age appears to have addressed any previously observed non-linearity, as there is no strong evidence of model misfit. The distribution of residuals across treatment groups also seems relatively balanced, with no obvious separation or heteroscedasticity.

The model’s functional form is appropriate, and the spline transformation has improved the model’s fit by flexibly accommodating the non-linear effect of age.

### Model Comparison: Linear vs Spline Age Term

We now compare the original multivariable Cox model with age modeled linearly to the updated model using a spline transformation for age. This allows us to formally test whether the added flexibility of the spline significantly improves model fit.

```{r}
# Compare linear age model vs spline-transformed age model
anova(cox_multivariable, cox_spline, test = "LRT")
```

A likelihood ratio test comparing a model with age modeled linearly to one using restricted cubic splines (df = 3) showed a statistically significant improvement in fit (χ² = 7.91, df = 2, p = 0.019). This suggests that the effect of age on survival is non-linear, and that modeling age with splines better captures its relationship with hazard.

**Visualizing the Effect of Spline-Transformed Age**

This section uses 'termplot()' to visualize how the hazard changes across age when modeled as a spline. This provides an intuitive look at the non-linear relationship.

::: {.callout-note title="Understanding ANOVA in Cox Regression" collapse="true"}

In the context of **Cox proportional hazards models**, ANOVA isn't used in the traditional sense (like comparing means). Instead, it's a way of **comparing nested models** using their **log-likelihoods**.

This is typically done using the **Likelihood Ratio Test (LRT)**, which answers the question:

> _"Does the more complex model fit significantly better than the simpler one?"_

**🔍 What the Output Shows:**

- **Model 1:** The **simpler model** (fewer terms or more constrained terms)
- **Model 2:** The **more complex model** (extra variables or flexible terms, e.g., splines)
- **Log-likelihoods:** How well each model fits the data (less negative = better)
- **Chisq:** Computed as `2 × (LL₂ − LL₁)`  
- **Degrees of freedom:** Difference in number of parameters
- **p-value:** Significance of improvement in fit

**✅ How to Interpret:**

- **p < 0.05** → The complex model fits significantly better → **keep it**
- **p ≥ 0.05** → No meaningful improvement → **stick with the simpler model**

This approach is especially useful for:
- Deciding whether to model continuous variables **non-linearly** (e.g., with splines)
- Evaluating whether new covariates **add predictive value**
- Performing **stepwise model refinement**

📦 **Takeaway:**  
Use ANOVA/LRT in Cox regression to guide decisions about **model complexity vs parsimony**, balancing better fit with interpretability.
:::


```{r}
termplot(cox_spline, se = TRUE, col.term = "blue")
```

The term plots from the spline-based Cox proportional hazards model illustrate the partial effect of each covariate on the log hazard:

- Age (modeled using natural splines with 3 degrees of freedom) exhibits a non-linear association with the log hazard. The curve displays a shallow U-shape, with the lowest hazard around age 55, and a gradual increase in hazard at both younger and older ages. While the trend is suggestive, the wide confidence intervals—particularly at the distribution tails—indicate uncertainty in the precise shape of this relationship.

- Treatment Group (trt) is represented as a binary variable with two flat lines, one for each group. The narrow confidence intervals overlap with zero, implying no statistically significant difference in log hazard between the intervention and control arms in this model.

- Prior Treatment (prior) also shows a flat trend across the range of prior treatments, with confidence bands consistently containing zero. This suggests that the number of prior therapies does not contribute meaningfully to hazard estimation when adjusting for age and treatment.

These visualizations support the statistical findings from the model summary. Age is the only covariate showing a potentially complex, non-linear effect, while treatment and prior exposure show no clear independent impact on survival in this cohort.

::: {.callout-note title="How to Interpret Term Plots in Cox Models" collapse="true"}

**Term plots** visualize the *partial effect* of each covariate in a fitted model while holding other variables constant. They are especially helpful for interpreting transformations like splines.

**🔹 What is a term plot?**  
A term plot shows how the **log hazard** (on the y-axis) changes with a covariate (on the x-axis), after accounting for all other variables in the model.

- The **blue line** is the estimated partial effect of that variable.
- The **dashed orange lines** show the 95% confidence interval.
- The interpretation depends on the type of variable:

**🔸 Continuous Variables (e.g., Age):**

- For **linear terms**, the blue line should be roughly straight.
- For **splines** or transformed terms:
  - A *curved* blue line indicates a non-linear relationship.
  - Areas where the confidence band **does not include 0** suggest a statistically significant effect on the log hazard.
  - Patterns like U-shapes, inflection points, or steep climbs reveal how risk evolves across the covariate range.

**🔸 Binary Variables (e.g., Treatment, Prior Therapy):**

- These are shown as **two horizontal segments**:
  - One for each group (e.g., Control = 0, Intervention = 1).
  - The **difference between the two** shows the log hazard shift due to that variable.
- If the confidence intervals **include zero**, there's no strong evidence that the variable affects the hazard.

**✅ Best Uses:**

- Confirming that spline-transformed terms behave as expected.
- Checking if a linear term should be transformed (e.g., age might be better modeled with splines).
- Double-checking whether binary variables have any meaningful effect.

🧠 _Tip:_ Term plots don't replace statistical tests, but they **visualize model estimates** clearly and intuitively.
:::

# Interaction or Additive Effects?

In Cox regression, interaction terms allow us to test whether the effect of one covariate on the hazard changes depending on the level of another covariate. If significant, this suggests a departure from simple additivity and may require stratified or more complex modelling. 

::: {.callout-note title="Understanding Interaction Effects in Cox Models" collapse=true}

In regression models—including **Cox proportional hazards models**—an *interaction effect* tests whether the relationship between a covariate and the outcome changes depending on the level of another covariate.

In practical terms, interaction terms allow us to ask:  
> *“Does the effect of Variable A on survival depend on Variable B?”*

*Why Test for Interaction?*

- To explore whether **treatment effects vary across subgroups** (e.g., age, gender, biomarker levels).
- To identify **non-additive relationships** that a main-effects-only model would miss.
- To **refine risk stratification** or support personalized approaches to care.

*What It Tells Us*

In the Cox model, including an interaction (e.g., `treatment * age`) allows the **hazard ratio for treatment** to vary at different values of age. If the interaction term is statistically significant, it suggests the **effect of treatment is modified by age**.

This can reveal:
- A treatment that works well in younger patients but poorly in older ones.
- A covariate that only has an impact in the presence of another factor.
- Complex survival dynamics that are **hidden in simpler additive models**.

*Interpreting Output*

- A significant interaction coefficient indicates a **departure from additivity**.
- It's essential to **cautiously interpret main effects** in the presence of interactions, as their meaning may be conditional.
- Visualization (e.g., stratified survival curves or effect plots) can aid interpretation.

*When to Use*

- After fitting and interpreting a main-effects model.
- When clinical theory or prior evidence suggests effect modification.
- When residual plots or diagnostics hint at **unexplained heterogeneity**.

:::

## Interaction of Treatment and Age

```{r}
# Fit a cox model with interaction between treatment and age 
cox_interaction <- coxph(Surv(time, status) ~ trt * ns(age, df = 3) + prior, data = veteran)
summary(cox_interaction)
```

We tested for interaction between treatment group and age (modeled using natural splines with 3 degrees of freedom) to explore whether the effect of treatment on survival varied across age groups. The interaction model revealed a statistically significant interaction for the second spline basis term (p = 0.037), suggesting a possible age-dependent treatment effect.

The overall Wald test for the model approached significance (p = 0.05), while the likelihood ratio test was marginal (p = 0.07), indicating that the model with interaction terms may provide a better fit than the additive model. These findings suggest that the benefit (or lack thereof) of treatment may differ depending on patient age, warranting further exploration or stratified modelling.

**Schoenfeld Residuals for Interaction Model**

```{r}
# Test proportional hazards assumption for the interaction model
cox.zph(cox_interaction)
```

The Schoenfeld residual test was used to assess whether the proportional hazards assumption holds for each covariate and interaction term in the model.

- trt: χ² = 1.39, p = 0.24 → No evidence of time-varying effects for treatment group.

- ns(age, df = 3): χ² = 3.90, p = 0.27 → The spline-transformed age term shows no strong time dependency.

- prior: χ² = 2.42, p = 0.12 → Also not statistically significant.

- trt × ns(age, df = 3): χ² = 5.37, p = 0.15 → The interaction between age and treatment shows no significant violation of proportionality either.

- GLOBAL test: χ² = 8.94 on 8 df, p = 0.35 → The model as a whole satisfies the PH assumption.

There is no strong evidence that the proportional hazards assumption is violated in this interaction model. The covariate effects, including the interaction, appear stable over time.

```{r}
# Visual check of Schoenfeld residuals for interaction model
plot(cox.zph(cox_interaction))
```

The global test using cox.zph() indicated no significant violation of the PH assumption (χ² = 8.94, df = 8, p = 0.35). Individual covariate tests also returned non-significant p-values, suggesting that the estimated hazard ratios remain approximately constant over time.

Visual inspection of the Schoenfeld residual plots supported this conclusion. The smoothed curves for treatment, age (spline terms), prior treatment, and their interaction remained relatively flat and within their 95% confidence bands. While a few outliers were present—particularly in the treatment and age components—these did not indicate strong evidence of non-proportionality.

Taken together, these results suggest that the proportional hazards assumption is reasonably met for all terms in the interaction model.

**Martingale Residuals for Interaction Model**

The Martingale residuals from the interaction model were plotted against the linear predictor (log hazard) to assess model fit and identify any potential patterns or outliers.

```{r}
# Residuals vs linear predictor
veteran$martingale_interaction <- residuals(cox_interaction, type = "martingale")
veteran$linear_pred_interaction <- predict(cox_interaction, type = "lp")
ggplot(veteran, aes(x = linear_pred_interaction, y = martingale_interaction, color = trt)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Martingale Residuals vs Linear Predictor\n(Interaction Model)",
    x = "Linear Predictor (Log Hazard)",
    y = "Martingale Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The Martingale residuals plot from the interaction model reveals a concentration of predicted log hazard values for the control group between 0.5 and 1.0, whereas predictions for the intervention group span a wider range. This may reflect the influence of the age-by-treatment interaction term, which adds complexity primarily for patients receiving the intervention. The residuals themselves are randomly scattered around zero, suggesting that the functional form of the covariates is appropriately specified and that no major nonlinearity remains unaccounted for.

**Deviance Residuals for Interaction Model**

Deviance residuals were generated to further assess the model fit and identify any potential outliers or influential observations.

::: {.callout-note title="Understanding Deviance Residuals in Cox Models" collapse=true}

**Deviance residuals** are a type of diagnostic measure used to assess how well a Cox proportional hazards model fits the observed survival data. Each deviance residual reflects the discrepancy between the observed outcome (event or censoring) and what the model predicts for an individual, after accounting for all covariates.

🔍 Why we use them:
- **Outlier detection**: Large residuals (positive or negative) may indicate subjects whose survival times were poorly predicted by the model.
- **Model fit check**: If residuals cluster systematically or display non-random patterns when plotted against the linear predictor, it may suggest issues such as:
  - Mis-specification of functional form (e.g., nonlinear effects not properly captured),
  - Omitted interactions or variables,
  - Violation of model assumptions.

📊 How to interpret:
- Deviance residuals are approximately normally distributed around 0.
- Values far from zero may indicate poor fit for specific observations.
- A balanced, symmetrical scatter around the horizontal axis suggests a well-fitting model.

🧠 When to use:
- After fitting a Cox model, especially when refining the covariate structure.
- In combination with **martingale** and **Schoenfeld residuals** to build a comprehensive diagnostic view.
- Before finalizing your model, particularly if you are considering transformations or interactions.

💡 Bonus tip:
Unlike martingale residuals, deviance residuals are **symmetrically distributed** and therefore better suited for identifying both under- and over-predictions, making them particularly useful for graphical diagnostics.

:::

```{r}
# Deviance residuals
veteran$deviance_interaction <- residuals(cox_interaction, type = "deviance")
# Plot deviance residuals
ggplot(veteran, aes(x = linear_pred_interaction, y = deviance_interaction, color = trt)) +
  geom_point(position = position_jitter(width = 0.1), alpha = 0.6) + # Adding jitter to avoid overlap
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey30") +
  theme_minimal(base_size = 13) +
  labs(
    title = "Deviance Residuals vs Linear Predictor\n(Interaction Model)",
    x = "Linear Predictor (Log Hazard)",
    y = "Deviance Residual",
    color = "Treatment Group"
  ) +
  scale_color_manual(values = c("Control" = "#A569BD", "Intervention" = "#45B39D")) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))
```

The deviance residual plot provides a diagnostic check on the adequacy of the Cox proportional hazards model that includes an interaction between treatment and age. In this plot, residuals are symmetrically distributed around zero with no discernible curvature or trend, suggesting that the model fits the data reasonably well.

Notably, the intervention group shows a broader range of linear predictor values (i.e., estimated log hazard), indicating greater variability in predicted risk compared to the control group, which clusters tightly between 0.5 and 1.0. Despite this asymmetry in predictor spread, the vertical dispersion of residuals remains balanced, and no strong outliers or patterns emerge. This supports the appropriateness of the interaction model and suggests no serious violations of model assumptions related to functional form or influential observations.

Overall, the deviance residuals confirm that the model adequately captures the observed survival data, with no clear evidence of misfit or heteroscedasticity.

## Interaction of Treatment and Prior Therapy

Prior treatments might alter a patient’s response to new therapies (e.g., due to resistance, immune suppression, or cumulative toxicity).

```{r}
# Fit a cox model with interaction between treatment and prior therapy
cox_interaction_prior <- coxph(Surv(time, status) ~ trt * prior + ns(age, df = 3), data = veteran)
summary(cox_interaction_prior)
```

A Cox proportional hazards model including an interaction between treatment and prior therapy was fit to assess potential effect modification. The interaction term (trt × prior) yielded a hazard ratio of 0.94 (95% CI: 0.87–1.02, p = 0.135), suggesting that the effect of treatment may be slightly attenuated in patients who had prior therapy, although this did not reach statistical significance. Model-level tests (LR, Wald, and Score) were borderline (p = 0.07–0.08), indicating weak evidence that the interaction model may offer improved fit. These findings suggest a potential interaction worth exploring in larger or more powered studies, but are not conclusive in this sample.

# Final Model Summary & Interpretation
# Conclusion
# Limitations